### 2012-04-12 Halos.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Sperical Overdensity Halo Finder}

Spherical overdensities, search algorithm, 
provided halo information, Comparison to PS and ST.

### 2012-04-12 Introduction.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Introduction}

Over the last few years, the science of cosmology has benefited from the increased computing power now available at many facilities,
as well as from high precision data from highly performing instruments like WMAP \cite{ref:WMAP5} and SDSS \cite{ref:SDSS} just to name  few.
After struggling with the lack of observational evidences, cosmology has now entered the realm of precision science, where the data are more and more abundant and, 
for the most part, its uncertainty can be brought down to the percent level. It is thus of the utmost importance to develop the theory to a high degree of precision
in order to constrain tightly the underlying free parameters of the theory.

At the moment, most observational data \cite{ref:WMAP5}\cite{ref:SN1A}\cite{ref:BAO} seem to agree on a concordance model, also called the {$\Lambda CDM$}, in which the energy budget of the universe today is dominated by a cosmological constant, negligible curvature, along with a significant amount of cold dark matter, while particles from the Standard Model only account for a few percent. In units of critical density, the above parameters are labeled $\Omega_{\Lambda}$, $\Omega_{k}$, $\Omega_{m}$ and $\Omega_{b}$.
In this framework, the universe has started from an early inflationary phase \cite{ref:inflation} which has ended up
creating the matter content from reheating \cite{ref:reheating} in a close to Gaussian distribution, in the sense that the two-point correlations, either in real or Fourrier space,  contain all possible statistical information. In this case, the initial power spectrum has an amplitude $\delta_{H}$ and a tilt $n_{s}$, which is consistent with being scale independent, i.e.$\alpha_{s} \equiv d n_{s}/d ln(k) =0$ \cite{ref:WMAP5}.
This radiation-matter plasma was originally extremely hot, but cooled as a result of adiabatic expansion, and eventually reached a temperature cold enough for electrons to recombine with protons, liberating a thermal radiation observed today as the cosmic microwave background (CMB). The universe was about 400 000 years old when that happened, and from then on, gravity was no longer suppressed by radiation pressure and started to grow structure out of the post-recombination inhomogeneities. In particular, the baryonic acoustic oscillation (BAO) \cite{ref:AO} present in the primordial plasma were frozen in the matter distribution and started to follow the Hubble flow. Much later, gravity grew regions over-dense enough to produce the first stars, supernovae, black holes and galaxies, which ignited the inter galactic medium and gradually re-ionized the gas. This process happened over an extended amount of time, probably a billion year \cite{ref:Ilian} and left very few pockets of neutral hydrogen.

One of the principal objectives of modern cosmology is thus to hunt down these parameters, in attempts to minimize the uncertainty around their values or break the degeneracy between constraints.
The CMB spectrum provides a lot of constrains on the theoretical parameters of the concordance model, but many of them are degenerate, and we must use other techniques in order to break the degeneracy. 
Weak lensing, baryonic acoustic oscillations (BAO), cluster formation and type 1A supernovae surveys
have been targeted as the most promising methods to constrain the dark energy equation of state \cite{ref:DETF}, but each contain
both theoretical and observational challenges.

To make a long story short, weak lensing is sensitive to XXX and allows XXX.

Since they can be used as standard rulers, the detection of BAO at various redshifts allows one to reconstruct the expansion history,
with some sensitivity to the evolution of the equation of state of dark energy. The observation requires a very large amount of galaxies whose
redshift are known to a high level of accuracy, the signal is partly erased by non-linear clustering and is furthermore sampled only through galaxies,
whose distribution could be biased with respect to the total mass. 
Structure formation is affected by the global expansion, which tends to slow down gravitational collapse, and it is possible to constrain
many parameters by measuring the growth factor of clusters at different redshifts \cite{ref:growth}. However, XXX
Finally, type 1A supernovae are though to be standard candles, whose redshift and luminosity detection also allow to measure the expansion history.
The assumption that they are actually standard candles as been questioned by many \cite{ref:SNdoubt}, since it is not obvious that these objects are redshift independent.

On the theoretical side, many of the methods above mentioned require a thorough understanding of the dynamics of structure formation, and
perturbation theory only works in the so-called linear regime. Smaller scales must be probed by solving Poisson's equation deep into the non-linear regime,
which requires the use of N-Body simulations. Many such codes have been written in the last few years, but not all of them are optimal for the same applications.
For example, XXXXX is a tree XXX, while XXX is a hybrid XXX. ATM, GOTPM, gadget2,...

PMFAST was developed in XXX \cite{ref:PMFAST} and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level. Also, the force was obtained by solving Poisson's equation for the the gravitational potential, which led to the force by finite difference. 
Finally, the simulation volume was decomposed into a set of slabs, which becomes problematic when we try to parallelize the calculation.
Each computing unit is assigned a sub-volume of the slab, which eventually reach the limit where one cannot divide the finest cell grid, in a non-adaptative mesh grid at least.

is not highly scalable to a large number of

Moreover, Poisson's equation was solved

### 2012-04-12 PID.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Other tools}

Particle ID, initial condition generator, projections, power spectrum code,  various expansions

### 2012-04-12 PoissonSolver.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Poisson Solver}
\label{sec:Poisson}

This section describes how Poisson's equation is solved
within {\small CUBEP3M}.

It is of course possible to shut down any of these components,

Double mesh levels, equation to match them, fudge factor, kernel, 
corrected kernel, pp and extended pp, fft to solve the force.

### 2012-04-12 ScalingPerformance.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Scaling Performances}

Real time versus size, bottle neck, extended pp, trials on scinet and ranger, 
p3dfft,

### 2012-04-12 Structure.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Presentation of the Code}
\label{sec:presentation}

An optimal N-body code is able to make efficient use of the available memory in order to maximize the size of the problem to solve,
and at the same time it must be fast for calculations to converge. In addition, communication time must be cut down to the strict minimum, 
and the load across all the available computing units must be as balanced as possible. 
In this section, we present the strategy adopted to address these challenges, and quickly walk through the code flow.
We refer the reader to future sections for detailed discussions on specific topics.
In its standard configuration, {\small CUBEP3M} is a {\small FORTRAN90} N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. The force of gravity from dark matter particles is decomposed into a sum of three distinct contribution,
which are constructed such as to optimize both speed and memory. The code has many extensions that departs from this basic scheme, and
we shall come back to these later.

\subsection{Memory foot-print and communication strategy}
\label{subsec:memory}
When designing a simulation geometry, one has to verify that the configuration fits in the system's memory.
The bulk of the memory foot-print is concentrated in a handful of large arrays that mostly concern the fine grid:
the coarse grid requires $4^3$ times less memory per node, hence has a minor contribution. 
In addition, many of these arrays are intermediate steps, hence it is possible to recycle the memory space required by the fine 
in the calculation of the coarse grid.  
Here are the largest  arrays used by the code:
\begin{enumerate}
\item{{\tt xv}, a phase-space array which stores the position and velocity of each particle} 
\item{{\tt ll}, a linked-list array that accelerate the access to particles in a local domain}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f}, a pair of three-dimensional array that stores 
the complete matter density for a given node in real and Fourier space respectively}
\item{{\tt force\_f}, an array that stores the force of gravity along the three Cartesian directions}
\item{{\tt kern\_f}, an array that stores the fine grid force kernel in the three directions}
\end{enumerate}

The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
Information is shared across multiple nodes with {\small MPI} communications. 
The short range force and the particle-particle interactions  are computed locally on a single processor,  
hence a node that has multiple cores can compute in parallel a series of such local contributions, or {\it tiles}, with the {\small OPENMP} technology.
The information exchange between cores that are part of the same motherboard is generally much faster than that across nodes,
hence it is generally a good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes.

\subsection{Code overview}
\label{subsec:overview}

The code flow is presented in Fig. \ref{fig:structure}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial condition generator,
and the force kernels on both grids are also constructed from the specific geometry of the simulation.
For clarity, these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceed to a determination of the redshift jump by comparing the step size constraints from each
force components and from the cosmic expansion.
The latter is found by Taylor expanding  Friedmann's equation up to the third order in the scale factor.
Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney}.
If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and write them to file.

\begin{figure}
\begin{verbatim}
program cubep3m
   call initialize
   do
       call timestep
       call particle_mesh
       if(checkpoint_step) then
          call checkpoint
       elseif(last_step)
          exit
       endif
   enddo
   call finalize
end program cubep3m
\end{verbatim}
\caption{Overall structure of the code.}
\label{fig:structure}
\end{figure}

\subsection{Two-level algorithm}
\label{subsec:particle_mesh}

The {\tt particle\_mesh} subroutine is conceptually identical to that of {\small PMFAST}, with the exception  that {\small CUBEP3M}
decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}, 
and we save the detailed description of the force calculation  in section \ref{sec:Poisson}.

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call particle_pass
   call link_list
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f      
      call link_list_pp
      call force_pp
      call update_velocity_pp
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm.}
\label{fig:particle_mesh}
\end{figure}

force calculation,

### 2012-04-12 Systematics.tex
Adding .tex, .bib, .bst, .cls and .sty files

\section{Systematics}
Random shift, resolution, initial redshift

### 2012-04-12 paper.tex
Adding .tex, .bib, .bst, .cls and .sty files

\documentclass[useAMS,usenatbib]{mn2e}

% The usenatbib command allows the use of Patrick Daly's natbib.sty for
% cross-referencing.
%
% If you wish to typeset the paper in Times font (if you do not have the
% PostScript Type 1 Computer Modern fonts you will need to do this to get
% smoother fonts in a PDF file) then uncomment the next line
% \usepackage{Times}

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
\input{abbrev}
\input{psfig.sty}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{txfonts}

\title[{\small} CUBEP3M: Closer to optimal N-body computations]{{\small} CUBEP3M: Closer to optimal N-body computations}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev, Hy Trac]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca} 
Hugh Merz$^{3}$ \thanks{E-mail: xxx},  Ilian T. Iliev$^{4}$ \thanks{E-mail: xxx} and
Hy Trac$^{5}$ \thanks{E-mail: xxx}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

\begin{document}

%\date{Accepted 1988 December 15. Received 1988 December 14; in original form 1988 October 11}
\date{\today}

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2011}

\maketitle

\label{firstpage}

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
a two-mesh gravity solver that was already among the fastest N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the subgrid level and across neighboring cells. The volume decomposition is now cubical,
and most of the calculations on the fine mesh are {\tt openmp} parallel.
Many utilities have been added and are briefly described, including a runtime halofinder
and particle identification tags. 
We discuss the structure of the code, its accuracy, and  its scaling performance.
\end{abstract}

\begin{keywords}
To specify...
\end{keywords}

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{Accuracy} %pairwise and density force, 2-points stats.
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
%\input{MHD}
\input{PID}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}

\bsp

\label{lastpage}

\end{document}

### 2012-04-13 Introduction.tex
Started to work on the Poissonsolver.tex and Structure.tex

@@
 
-PMFAST was developed in XXX \cite{ref:PMFAST} and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level.
+PMFAST was developed in XXX \cite{ref:PMFAST} (we refer to this paper as MPT hereafter) and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level.
 Also, the force was obtained by solving Poisson's equation for the the gravitational potential, which led to the force by finite difference.

### 2012-04-13 PoissonSolver.tex
Started to work on the Poissonsolver.tex and Structure.tex

@@
 
-This section describes how Poisson's equation is solved
-within {\small CUBEP3M}.
+This section describes how Poisson's equation is solved at each time step of the main loop. 
+Many parts of the algorithm are identical to {\small PMFAST}, hence we refer extensively 
+to section 2 of MPT.
 
-It is of course possible to shut down any of these components,
+The reason that motivates the double mesh configuration is to allow the force of gravity to be broken down into 
+a long and a short range components. This two-mesh decomposition was first seen in Hydra \citep{hydra}, 
+and is now used by many codes, including {\small PMFAST}, {\bf (TracPen2003)} and {\bf (Other codes?)}.
+The long range force requires knowledge of the complete simulated volume, 
+but can be computed on a coarser grid without degrading significantly the precision of the calculation.
+Using a grid 64 times coarser limits the global communication time, plus contribute only to a fraction of the total memory requirement.
+The short range force is computed on the finer mesh and is local in the sense that it only knows about the matter in the immediate surroundings.
+It can thus be computed simultaneously on many computing units with {\small OPENMP} parallel loops, each {\small CPU} solving for a different sub-volume,
+as long as the arrays are constructed such as to support parallel reading and writing\footnote{In practice, this is done by adding an additional dimension
+to the relevant arrays, such that each {\small CPU} accesses a unique memory location.}. 
+The particle-particle (pp) interactions are also local operations, and are computed inside the same {\small OPENMP} loops.
 
-Double mesh levels, equation to match them, fudge factor, kernel, 
-corrected kernel, pp and extended pp, fft to solve the force.
+The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
+  or a force  kernel $\omega_{F} ({\bf x})$.
+It is a curl-free field, which allows us to relate the potential $\phi({\bf x})$to the source term via Poisson's equation: 
+\begin{eqnarray}
+\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
+\label{eq:poisson}
+\end{eqnarray}
+$G$ being Newton's constant. We solve this equation in Fourier space, where we write
+\begin{eqnarray}
+ \tilde{\phi}({\bf k}) = \frac{4 \pi G \tilde{\rho}({\bf k})}{- k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
+\label{eq:poissonFourier}
+\end{eqnarray}
+The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
+Using the convolution theorem, we can write
+\begin{eqnarray}
+ \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}
+\label{eq:poisson_solution_pot}
+\end{eqnarray}
+Although this approach is fast, it involves a finite differentiation which enhances the numerical noise.
+We therefore opt for a force kernel, which is more accurate but has the inconvenient to require four extra Fourier transforms.
+In this case, we must solve the convolution in three dimensions:
+\begin{eqnarray}
+ F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) {\bf  \omega}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
+\label{eq:poisson_solution_force}
+\end{eqnarray}
+The differentiation does not affect the density since it only acts on un-prime variables,
+and the force kernel is given by 
+\begin{eqnarray}
+{\bf  \omega}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
+\end{eqnarray}
+
+Following MPT, we split  the force kernel into two components, for the short and long range respectively, and 
+match the overlapping regions with a polynomial. Namely, we have:
+\begin{eqnarray}
+{\bf \omega}_{s}(r) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
+0 & \mbox{otherwise} 
+\end{cases}
+\end{eqnarray}
+and
+\begin{eqnarray}
+{\bf \omega}_{l}(r) = \begin{cases} {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
+ {\bf\omega}_{F}(r)  &\mbox{otherwise} 
+\end{cases}
+\end{eqnarray}
+The vector $ {\bf \beta}(r)$ is related to the fourth order polynomial that is used in the potential case by
+ $ {\bf \beta} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
+ and we get
+  \begin{eqnarray}
+   {\bf \beta}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
+  \end{eqnarray}
+The long range force is subject to a correction that fixes {\bf (what exactly?)}.
+
+ equation to match them, fudge factor,
+corrected kernel, pp and extended pp.

### 2012-04-13 Structure.tex
Started to work on the Poissonsolver.tex and Structure.tex

@@
 and the force kernels on both grids are also constructed from the specific geometry of the simulation.
-For clarity, these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
+These kernels are slightly modified in the code configuration where the exact Newtonian force is extended outside a fine grid cell.
+For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
 although they are actually distinct calls in the code.
@@
 Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
-which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney}.
+which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney} {\bf (more on leapfrog?)}.
 If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
@@
 decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
-A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}, 
-and we save the detailed description of the force calculation  in section \ref{sec:Poisson}.
+A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}.
+We save the detailed description of the force calculation  in section \ref{sec:Poisson}.
 
@@
       call update_velocity_pp
+      if(extended_pp = .true.)
+         call link_list_pp_extended
+         call force_pp_extended
+         call update_velocity_pp_extended       
+      endif
    end do
@@
 \end{figure}
-
-force calculation,

### 2012-04-13 Introduction.tex
Worked on the intro

--- 2012-04-12 Introduction.tex
@@
 
-Over the last few years, the science of cosmology has benefited from the increased computing power now available at many facilities,
-as well as from high precision data from highly performing instruments like WMAP \cite{ref:WMAP5} and SDSS \cite{ref:SDSS} just to name  few.
-After struggling with the lack of observational evidences, cosmology has now entered the realm of precision science, where the data are more and more abundant and, 
-for the most part, its uncertainty can be brought down to the percent level. It is thus of the utmost importance to develop the theory to a high degree of precision
-in order to constrain tightly the underlying free parameters of the theory.
+Many physical and astrophysical systems are subject to non-linear dynamics
+and rely on N-body simulations to describe the evolution of bodies. 
+One of the main field of application is the modelling of large scale structures, 
+which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
+\citep{WMAP5-7} of galaxy clustering \citep{2df, SDSS, WiggleZ, BOSS} of weak gravitational lensing \citep{CFHTLS, SDSS}
+and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
+and collisionless dark matter occupy more than 95 per cent of the total energy density of the universe.
+In such a paradigm, pure N-body code are perfectly suited to describe the dynamics, as long as we
+understands where and how the baryonic fluid feeds back on the dark matter structure.
+The next generation of measurements aim at constraining the cosmological parameters at the per cent level, 
+and the theoretical understanding of the non-linear dynamics that govern structure formation heavily relies 
+on numerical simulations.
 
-At the moment, most observational data \cite{ref:WMAP5}\cite{ref:SN1A}\cite{ref:BAO} seem to agree on a concordance model, also called the {$\Lambda CDM$}, in which the energy budget of the universe today is dominated by a cosmological constant, negligible curvature, along with a significant amount of cold dark matter, while particles from the Standard Model only account for a few percent.
-In units of critical density, the above parameters are labeled $\Omega_{\Lambda}$, $\Omega_{k}$, $\Omega_{m}$ and $\Omega_{b}$.
-In this framework, the universe has started from an early inflationary phase \cite{ref:inflation} which has ended up
-creating the matter content from reheating \cite{ref:reheating} in a close to Gaussian distribution, in the sense that the two-point correlations, either in real or Fourrier space,  contain all possible statistical information.
-In this case, the initial power spectrum has an amplitude $\delta_{H}$ and a tilt $n_{s}$, which is consistent with being scale independent, i.e.$\alpha_{s} \equiv d n_{s}/d ln(k) =0$ \cite{ref:WMAP5}.
-This radiation-matter plasma was originally extremely hot, but cooled as a result of adiabatic expansion, and eventually reached a temperature cold enough for electrons to recombine with protons, liberating a thermal radiation observed today as the cosmic microwave background (CMB).
-The universe was about 400 000 years old when that happened, and from then on, gravity was no longer suppressed by radiation pressure and started to grow structure out of the post-recombination inhomogeneities.
-In particular, the baryonic acoustic oscillation (BAO) \cite{ref:AO} present in the primordial plasma were frozen in the matter distribution and started to follow the Hubble flow.
-Much later, gravity grew regions over-dense enough to produce the first stars, supernovae, black holes and galaxies, which ignited the inter galactic medium and gradually re-ionized the gas.
-This process happened over an extended amount of time, probably a billion year \cite{ref:Ilian} and left very few pockets of neutral hydrogen.
+For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation scale can provide
+tight constraints on the dark energy equation of scale \citep{Seo2003,2005}.
+The most optimal estimates of the uncertainty requires the knowledge of the 
+matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
+of N-body simulations \citep{RH2005, japan2,japan3, Ngan}.
+For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
+by propagating photons in past light cones that are extracted from simulated density fields \citep{HarnoisLudo, Takahashi }.
+{\bf (Ilian, could you say something about reionization here?)}
 
-One of the principal objectives of modern cosmology is thus to hunt down these parameters, in attempts to minimize the uncertainty around their values or break the degeneracy between constraints.
-The CMB spectrum provides a lot of constrains on the theoretical parameters of the concordance model, but many of them are degenerate, and we must use other techniques in order to break the degeneracy.
-Weak lensing, baryonic acoustic oscillations (BAO), cluster formation and type 1A supernovae surveys
-have been targeted as the most promising methods to constrain the dark energy equation of state \cite{ref:DETF}, but each contain
-both theoretical and observational challenges.
+The basic problem that is addressed with N-body codes is a time evolution of an ensemble of $N$ particles
+that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operation, a cost that 
+exceeds the memory and speed of current machines for large problems.
+Solving the problem  on a mesh \citep{Hockney} reduces to $O(N\mbox{log}N)$ the number of operations,
+as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques, 
+typically using high performance libraries such as {\small FFTW} \citep{FFTW}.
 
-To make a long story short, weak lensing is sensitive to XXX and allows XXX.
+With the advent of large computing facilities, parallel coding has now become common practice, 
+and N-body codes have evolved both in performance and complexity. 
+Many codes have opted for a tree algorithm \citep{Gadjet, Gadjet2, TPM, GOTPM}, in which 
+the local resolution increases with the density of the matter field. 
+These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
+The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
+Unlike Hydra \citep{couchman1991} and the PM code by \cite{FerrelBertschinger1995},
+ these are not designed to perform large scale PM calculations.
 
-Since they can be used as standard rulers, the detection of BAO at various redshifts allows one to reconstruct the expansion history,
-with some sensitivity to the evolution of the equation of state of dark energy. The observation requires a very large amount of galaxies whose
-redshift are known to a high level of accuracy, the signal is partly erased by non-linear clustering and is furthermore sampled only through galaxies,
-whose distribution could be biased with respect to the total mass. 
-Structure formation is affected by the global expansion, which tends to slow down gravitational collapse, and it is possible to constrain
-many parameters by measuring the growth factor of clusters at different redshifts \cite{ref:growth}. However, XXX
-Finally, type 1A supernovae are though to be standard candles, whose redshift and luminosity detection also allow to measure the expansion history.
-The assumption that they are actually standard candles as been questioned by many \cite{ref:SNdoubt}, since it is not obvious that these objects are redshift independent.
+{\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
+both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
+The long range gravitational force is computed on a  grid four times coarser, such as to minimize the communication time
+and to fit in system's memory. The short range is computed locally on a finer mesh, which can be as tight as the memory allows.
+This enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
 
-On the theoretical side, many of the methods above mentioned require a thorough understanding of the dynamics of structure formation, and
-perturbation theory only works in the so-called linear regime. Smaller scales must be probed by solving Poisson's equation deep into the non-linear regime,
-which requires the use of N-Body simulations. Many such codes have been written in the last few years, but not all of them are optimal for the same applications.
-For example, XXXXX is a tree XXX, while XXX is a hybrid XXX. ATM, GOTPM, gadget2,...
-
-PMFAST was developed in XXX \cite{ref:PMFAST} and was already highly efficient, although the resolution was limited by the absence of particle-particle interaction at sub-grid level.
-Also, the force was obtained by solving Poisson's equation for the the gravitational potential, which led to the force by finite difference.
-Finally, the simulation volume was decomposed into a set of slabs, which becomes problematic when we try to parallelize the calculation.
-Each computing unit is assigned a sub-volume of the slab, which eventually reach the limit where one cannot divide the finest cell grid, in a non-adaptative mesh grid at least.
-
-is not highly scalable to a large number of
-
-Moreover, Poisson's equation was solved
+{\bf (Describe briefly   Hydra, PM code by japanese, and others...)}

### 2012-04-16 Introduction.tex
Completed first draft of Introduction.tex

@@
 matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
-of N-body simulations \citep{RH2005, japan2,japan3, Ngan}.
+of N-body simulations \citep{RH2005, japan2,japan3}.
 For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
-by propagating photons in past light cones that are extracted from simulated density fields \citep{HarnoisLudo, Takahashi }.
-{\bf (Ilian, could you say something about reionization here?)}
+by propagating photons in past light cones that are extracted from simulated density fields \citep{White, Jain, Takahashi}.
+{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
+accurate predictions/forecasts, etc.)}
 
@@
 
+{\bf (Describe briefly   Hydra, PM code by japanese, and others...)}
+
 {\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
@@
 The long range gravitational force is computed on a  grid four times coarser, such as to minimize the communication time
-and to fit in system's memory. The short range is computed locally on a finer mesh, which can be as tight as the memory allows.
-This enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
+and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
+to be stored at a given time, allowing for parallel computation.
+This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
+One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
+and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
+and  for computing the long range FFTs.
+As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
+in buffer zones is done at the same time as particles are passed to adjacent nodes,
+and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
+designed specifically to preserve a high processor load.
 
-{\bf (Describe briefly   Hydra, PM code by japanese, and others...)}
+Since its first published version, {\small PMFAST} has evolved in many aspects. 
+The first major improvement was to transform the volume decomposition in multi-node configurations 
+from slabs to cubes. One of the problem with slabs is that they do not scale well to large runs: 
+as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
+until it reaches the hard limit of a single cell layer.  With this enhancement came a new name, {\small CUBEPM},
+which soon after incorporated particle-particle (pp) interactions at the sub-grid level, and the code was renamed
+{\small CUBEP3M}. It now includes a significant number of new features since MPT : the pp force
+can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
+a runtime halofinder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features
+{\bf (anything else?)}.
+{\small CUBEP3M} has already been involved in a number of scientific applications over the last few years,
+spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
+ \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)}
+
+This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
+It is structured as follow: XXXX

### 2012-04-16 paper.tex
Completed first draft of Introduction.tex

@@
 \title[{\small} CUBEP3M: Closer to optimal N-body computations]{{\small} CUBEP3M: Closer to optimal N-body computations}
-\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev, Hy Trac]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
+\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca} 
-Hugh Merz$^{3}$ \thanks{E-mail: xxx},  Ilian T. Iliev$^{4}$ \thanks{E-mail: xxx} and
-Hy Trac$^{5}$ \thanks{E-mail: xxx}\\
+Hugh Merz$^{3}$ \thanks{E-mail: xxx} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: xxx}\\
 %\footnotemark[1]\thanks{This file has been amended to

### 2012-04-17 Extendedpp.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

\subsection{Extended range of the pp force}
\label{subsec:extendedpp}

{\bf (JD, here will be one of your contribution!)}

### 2012-04-17 Initial.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

\subsection{Initial Conditions}
\label{subsec:init}

As mention in section \ref{sec:structure}, the code start off by reading a set of initial conditions.
These correspond to a  $6 \times N$ phase-space array, where $N$ is the number of particles in the
local node. Although most applications so far were based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial condition generator that we briefly describe in this section. 
{\bf Vincent, here you go!}

### 2012-04-17 Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

@@
 both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
-The long range gravitational force is computed on a  grid four times coarser, such as to minimize the communication time
+The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
 and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
-to be stored at a given time, allowing for parallel computation.
+to be stored at a given time, allowing for {\small OPENMP}parallel computation.
 This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
@@
 {\bf (anything else?)}.
-{\small CUBEP3M} has already been involved in a number of scientific applications over the last few years,
-spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
- \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)}
 
 This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
-It is structured as follow: XXXX
+It has already been involved in a number of scientific applications over the last few years,
+spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
+ \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
+and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
+The paper is structured as follow: XXXX

### 2012-04-17 PID.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

@@
-\section{Other tools}
+\subsection{Particle identification tags}
+\label{PID}
 
-Particle ID, initial condition generator, projections, power spectrum code,  various expansions
+A system of particle identification can be turned on, which basically allows to track each particle's trajectory
+between checkpoints. Such a tool is useful to a number of applications, from reconstruction of halo merging history to {\bf (what else?)}
+The particle tag system has been implemented as an array of double integers, {\tt PID}, 
+and assigns a unique integer to each particle during the initialization stage. The location of the tag on the {\tt PID} array 
+matches the location of the corresponding particle on the {\tt xv} array, hence it acts as if the latter array had an extra dimension.
+The location change only when particles exist the local volume, in which case the tag is sent along with the particle in the {\tt pass\_particle} subroutine.
+Deleted particles results in deleted flags, and the {\tt PID} array gets written to file at each particle checkpoint.
+{\bf (Anything else to say here?)}

### 2012-04-17 PoissonSolver.tex
Completed first draft of Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

@@
 
-This section describes how Poisson's equation is solved at each time step of the main loop. 
-Many parts of the algorithm are identical to {\small PMFAST}, hence we refer extensively 
-to section 2 of MPT.
-
-The reason that motivates the double mesh configuration is to allow the force of gravity to be broken down into 
-a long and a short range components. This two-mesh decomposition was first seen in Hydra \citep{hydra}, 
-and is now used by many codes, including {\small PMFAST}, {\bf (TracPen2003)} and {\bf (Other codes?)}.
-The long range force requires knowledge of the complete simulated volume, 
-but can be computed on a coarser grid without degrading significantly the precision of the calculation.
-Using a grid 64 times coarser limits the global communication time, plus contribute only to a fraction of the total memory requirement.
-The short range force is computed on the finer mesh and is local in the sense that it only knows about the matter in the immediate surroundings.
-It can thus be computed simultaneously on many computing units with {\small OPENMP} parallel loops, each {\small CPU} solving for a different sub-volume,
-as long as the arrays are constructed such as to support parallel reading and writing\footnote{In practice, this is done by adding an additional dimension
-to the relevant arrays, such that each {\small CPU} accesses a unique memory location.}. 
-The particle-particle (pp) interactions are also local operations, and are computed inside the same {\small OPENMP} loops.
+This section reviews how Poisson's equation is solved on a double-mesh configuration. 
+Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
+to section 2 of MPT for more details. In cubep3m, the mass default assignment scheme are
+a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
+for the fine grid \citep{Hockney}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
+for pairs closer to a tenth of a grid cell.
 
@@
   or a force  kernel $\omega_{F} ({\bf x})$.
-It is a curl-free field, which allows us to relate the potential $\phi({\bf x})$to the source term via Poisson's equation: 
+Gravity fields are curl-free, which allows us to relate the potential $\phi({\bf x})$ to the source term via Poisson's equation: 
 \begin{eqnarray}
@@
 
-Following MPT, we split  the force kernel into two components, for the short and long range respectively, and 
-match the overlapping regions with a polynomial. Namely, we have:
+Following the spherically symmetric matching technique of MPT (section 2.1), 
+we split  the force kernel into two components, for the short and long range respectively, and 
+match the overlapping region with a polynomial. Namely, we have:
 \begin{eqnarray}
-{\bf \omega}_{s}(r) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
+{\bf \omega}_{s}({\bf r}) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 0 & \mbox{otherwise} 
@@
   \end{eqnarray}
-The long range force is subject to a correction that fixes {\bf (what exactly?)}.
 
- equation to match them, fudge factor,
-corrected kernel, pp and extended pp.
+Because these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
+both with the density and the kernel (see [Eq. 7-8] of MPT).
+When matching the two  force kernels, it was realized that close to the cutoff region, the long range force is always on the low side, whereas 
+the short range force is scattered across the theoretical $1/r^2$ value. These behaviours are purely features of the CIC and NGP interpolation scheme 
+respectively. We identified a small range surrounding the cutoff length, in which we empirically adjust both kernels such as to improve 
+the match. Namely, for $14 \le r \le 16$, ${\bf \omega}_{s}({\bf r}) \rightarrow {\bf \omega}_{s}({\bf r})*0.985$,
+and for  $12 \le r \le 16$, ${\bf \omega}_{l}({\bf r}) \rightarrow {\bf \omega}_{l}({\bf r})*1.2$.
+The two fudge factors were found by performing force measurements on two particles randomly placed in the volume.
+
+In addition, the long range force is subject to a correction that fixes {\bf (what exactly fixes the LRCKCORR flag?)}.
+
+{\bf (Anything else here?)}

### 2012-04-17 Structure.tex
Completed first draft of Introduction.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

--- 2012-04-12 Structure.tex
@@
-\section{Presentation of the Code}
-\label{sec:presentation}
+\section{Review of the Code Structure}
+\label{sec:structure}
 
-An optimal N-body code is able to make efficient use of the available memory in order to maximize the size of the problem to solve,
-and at the same time it must be fast for calculations to converge. In addition, communication time must be cut down to the strict minimum, 
-and the load across all the available computing units must be as balanced as possible. 
-In this section, we present the strategy adopted to address these challenges, and quickly walk through the code flow.
-We refer the reader to future sections for detailed discussions on specific topics.
-In its standard configuration, {\small CUBEP3M} is a {\small FORTRAN90} N-body code that solves Poisson's equation on a two-level mesh, 
-with sub-cell accuracy thanks to particle-particle interactions. The force of gravity from dark matter particles is decomposed into a sum of three distinct contribution,
-which are constructed such as to optimize both speed and memory. The code has many extensions that departs from this basic scheme, and
-we shall come back to these later.
+An optimal large scale N-body code must address many challenges: minimize the memory footprint to allow larger dynamical range,
+minimize the passing of information across computing nodes, reduce and accelerate the memory accesses to the large scale arrays, 
+make efficient use of high performance libraries to speed up standard calculations like Fourier transforms, just to name a few.
+In the realm of parallel programming, high efficiency  can be assessed when a high load is balanced across all processors
+most of the time. In this section, we present the general strategies adopted to address these challenges\footnote{ 
+Many originate directly from MPT and were preserved in {\small CUBEP3M};
+those will be briefly mentioned, and we shall refer the reader to the original PMFAST paper for greater details.}.
+We start with a walkthrough the code flow, and briefly discuss some specific sections that depart from standard N-body codes,
+while referring the reader to future sections for detailed discussions on selected topics.
 
-\subsection{Memory foot-print and communication strategy}
-\label{subsec:memory}
-When designing a simulation geometry, one has to verify that the configuration fits in the system's memory.
-The bulk of the memory foot-print is concentrated in a handful of large arrays that mostly concern the fine grid:
-the coarse grid requires $4^3$ times less memory per node, hence has a minor contribution. 
-In addition, many of these arrays are intermediate steps, hence it is possible to recycle the memory space required by the fine 
-in the calculation of the coarse grid.  
-Here are the largest  arrays used by the code:
+As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
+N-body code that solves Poisson's equation on a two-level mesh, 
+with sub-cell accuracy thanks to particle-particle interactions. 
+The code has extensions that departs from this basic scheme, and
+we shall come back to these later, but for the moment, we adopt the 
+standard configuration. 
+The long range component of the gravity force is solved on the coarse grid, 
+and is global in the sense that the calculations require knowledge about the full simulated volume.
+The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing.
+In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles},
+and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.
+Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes, as the information exchange between cores that are part of the same motherboard is generally much faster.
+As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned.
+The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
+is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}
+
+%\subsection{Memory foot-print and communication strategy}
+%\label{subsec:memory}
+
+Because the coarse grid arrays require $4^3$ times less memory per node, 
+the bulk of the foot-print is concentrated in a handful of fine grid arrays.
+In addition, some of these are required for intermediate steps of the calculations only, 
+hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
+We present here the largest arrays used by the code:
 \begin{enumerate}
-\item{{\tt xv}, a phase-space array which stores the position and velocity of each particle} 
-\item{{\tt ll}, a linked-list array that accelerate the access to particles in a local domain}
-\item{{\tt rho\_f} and {\tt cmplx\_rho\_f}, a pair of three-dimensional array that stores 
-the complete matter density for a given node in real and Fourier space respectively}
-\item{{\tt force\_f}, an array that stores the force of gravity along the three Cartesian directions}
-\item{{\tt kern\_f}, an array that stores the fine grid force kernel in the three directions}
+\item{{\tt xv} stores the position and velocity of each particle} 
+\item{{\tt ll} stores the linked-list that accelerate the access to particles in a local domain}
+\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
+the local fine grid density  in real and Fourier space respectively}
+\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
+\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
 \end{enumerate}
 
-The long range component of the gravity force is solved on the coarse grid, 
-and is global in the sense that the calculations require knowledge about the full simulated volume.
-Information is shared across multiple nodes with {\small MPI} communications. 
-The short range force and the particle-particle interactions  are computed locally on a single processor,  
-hence a node that has multiple cores can compute in parallel a series of such local contributions, or {\it tiles}, with the {\small OPENMP} technology.
-The information exchange between cores that are part of the same motherboard is generally much faster than that across nodes,
-hence it is generally a good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes.
+%\subsection{Code overview}
+%\label{subsec:overview}
 
-\subsection{Code overview}
-\label{subsec:overview}
-
-The code flow is presented in Fig. \ref{fig:structure}.
+The code flow is presented in Fig. \ref{fig:structure} and \ref{fig:particle_mesh}.
 Before entering the main loop, the code starts with an initialization stage, 
@@
 the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
-The phase-space array  is obtained from the output of the initial condition generator,
-and the force kernels on both grids are also constructed from the specific geometry of the simulation.
-For clarity, these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
+The phase-space array  is obtained from the output of the initial conditions generator,
+and the force kernels on both grids are constructed from the specific geometry of the simulation.
+For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
 although they are actually distinct calls in the code.
@@
 Each iteration of the main loop starts with the {\tt timestep} subroutine, 
-which proceed to a determination of the redshift jump by comparing the step size constraints from each
-force components and from the cosmic expansion.
-The latter is found by Taylor expanding  Friedmann's equation up to the third order in the scale factor.
-Poisson's equation is then solved  in the {\tt particle\_mesh} subroutine,
-which updates the positions and velocities of the dark matter particles with a leapfrog scheme \citep{Hockney}.
-If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
-and write them to file.
+which proceeds to a determination of the redshift jump by comparing the step size constraints from each
+force components and from the scale factor.
+The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
+and can accommodate constant or running equation of state of dark energy.
+The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
+which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
+creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
+with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
+The loop over tiles and the particle exchange are thus performed in three dimensions.
+{\bf (Should I mention leapfrog here or anywhere else?)}
+When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
+and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
+do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
+only the head processor is active. Other libraries such as {\small P3DFFT} or  Intel {\small MKL-$\alpha$} currently permit this extra level of parallelization,
+and it is our plan to migrate to one of these in the near future.
 
@@
 
-\subsection{Two-level algorithm}
-\label{subsec:particle_mesh}
-
-The {\tt particle\_mesh} subroutine is conceptually identical to that of {\small PMFAST}, with the exception  that {\small CUBEP3M}
-decomposes the volume into cubes (as opposed to slabs). The loop over tiles is thus performed in three dimension.
-A schematic representation of the subroutine is shown in Fig. \ref{fig:particle_mesh}, 
-and we save the detailed description of the force calculation  in section \ref{sec:Poisson}.
-
 \begin{figure}
@@
    call update_position
+   call link_list
    call particle_pass
-   call link_list
    !$omp parallel do
@@
       call update_velocity_f
+      if(pp = .true.) then       
          call link_list_pp
@@
          call update_velocity_pp
+         if(extended_pp = .true.) then
+            call link_list_pp_extended
+            call force_pp_extended
+            call update_velocity_pp_extended       
+         endif
+      endif
    end do
@@
 \end{verbatim}
-\caption{Overall structure of the two-level mesh algorithm.}
+\caption{Overall structure of the two-level mesh algorithm.
+We have included the section that concerns the extended pp force calculation to show that it follows the same basic logic.
+We mention here that the three linked list arrays are distinct entities, and their structure differ slightly.
+}
 \label{fig:particle_mesh}
@@
 
-force calculation,
+If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
+and writes them to file.
+Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time {\bf (is this correct?)}.
+The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 
+and clears the {\small MPI} communicators. We have collected those operations under the subroutine {\tt finalize} for concision.
+
+{\bf (Anything else to say in this section?)}

### 2012-04-17 paper.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...

@@
 
-\title[{\small} CUBEP3M: Closer to optimal N-body computations]{{\small} CUBEP3M: Closer to optimal N-body computations}
+\title[{\small} CUBEP3M: High Performance P3M N-body code]{{\small} CUBEP3M: High Performance P3M N-body code}
 \author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
-\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca} 
-Hugh Merz$^{3}$ \thanks{E-mail: xxx} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: xxx}\\
+\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
+Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
 %\footnotemark[1]\thanks{This file has been amended to
@@
 \input{Halos}
-%\input{MHD}
-\input{PID}
+\input{Extensions}
 \input{Conclusion}

### 2012-04-17 Extendedpp.tex
More improvements

@@
 
-{\bf (JD, here will be one of your contribution!)}
+One of the main source of error in the calculation of the force occurs when on the smallest scales of the fine grid.
+The approximation by which particles in a neighbouring mesh grid can be placed at the centre of the cell
+is less accurate, which cause a maximal scatter around the exact $1/r^2$ law.
+A solution to minimize this error consists in extending the pp force calculation outside a single cell,
+which inevitably reintroduces a $N^2$ number of operations. Our goal is to add the flexibility to have a code
+that runs slower, but produces results with a higher precision.
+
+To allow this feature, we  have to choose how far outside a cell we want the exact pp force.  
+Since the force kernels on both meshes are organized in terms of grids, the simplest way to implement this 
+feature is to shut down the mesh kernels in a region of specified size, and allow the pp force to extend therein.
+Concretely, these regions are constructed as cubic layers of fine mesh grids around a central cell; 
+the freedom we have is to choose the number of such layers.
+
+ To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
+ to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell.
+We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
+ we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
+ To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
+ $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
+ \begin{itemize}
+ \item{$z_2 \ge z_1$ always}
+ \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
+ \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
+ \end{itemize}
+ The case where all three coordinates are equal is already calculated in the standard configuration of the code.
+
+ To quantify the accuracy improvement versus computing time requirements, we performed the following test.
+ We generate a set of initial conditions at a starting redshift of $z = XXX$, with a box size equal to $ XXX h^{-1}\mbox{Mpc}$,
+ and with $XXX^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
+ the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
+ such that the only difference between different runs is the range of the pp force.
+{\bf (JD, insert here your P(k) plot and a description of what you observe, in terms of accuracy improvement and real time cpu usage...)}

### 2012-04-17 Introduction.tex
More improvements

@@
 which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
-\citep{WMAP5-7} of galaxy clustering \citep{2df, SDSS, WiggleZ, BOSS} of weak gravitational lensing \citep{CFHTLS, SDSS}
+\citep{WMAP5-7} of galaxy clustering \citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
 and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
@@
 matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
-of N-body simulations \citep{RH2005, japan2,japan3}.
+of N-body simulations \citep{2005MNRAS.360L..82R, 2009ApJ...700..479T, 2011ApJ...726....7T}.
 For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
-by propagating photons in past light cones that are extracted from simulated density fields \citep{White, Jain, Takahashi}.
+by propagating photons in past light cones that are extracted from simulated density fields \citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
 {\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
@@
 exceeds the memory and speed of current machines for large problems.
-Solving the problem  on a mesh \citep{Hockney} reduces to $O(N\mbox{log}N)$ the number of operations,
+Solving the problem  on a mesh \citep{1981csup.book.....H} reduces to $O(N\mbox{log}N)$ the number of operations,
 as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques, 
-typically using high performance libraries such as {\small FFTW} \citep{FFTW}.
+typically using high performance libraries such as {\small FFTW} \citep{FFTW3}.
 
@@
 
-{\small PMFAST} (\cite{PMFAST}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
+{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
 both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
@@
 It has already been involved in a number of scientific applications over the last few years,
-spanning the field of weak lensing \citep{Sanaz_Lu, Lu_dore, Dore_Lu, Harnois_Vafaei, LuCMBLensing},  BAO
- \citep{HarnoisPen, Ngan,Zhang, Yu, HarnoisYu in prep.}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
+spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
+ \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
 and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 

### 2012-04-17 PoissonSolver.tex
More improvements

@@
 a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
-for the fine grid \citep{Hockney}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
+for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
 for pairs closer to a tenth of a grid cell.

### 2012-04-17 Systematics.tex
Significant changes to Poisson Solver, precised the skeleton, added a few discussion shere and there...
More improvements

@@
-\section{Systematics}
-Random shift, resolution, initial redshift
+\section{Systematics and Accuracy}
+\label{sec:systematics}
+
+To discuss :
+
+1) random shift to shuffle systematic clustering at the grid scale,
+
+2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
+{\bf JD, here you could discuss the ra\_max tempering...}
+
+3) limits of resolution caused by softening length,
+
+4) Poisson shot noise if too few particles or unrelaxed systems, 5) finite box size effects (with beat coupling?)
+
+Show plots of pairwise force (transverse and radial), density force, power spectrum, 
+mention the work on covariance matrix by Ngan, Harnoisetal.

### 2012-04-17 paper.tex
More improvements

@@
 Among the principal changes, the Poisson solver includes particle-particles interactions
-at the subgrid level and across neighboring cells. The volume decomposition is now cubical,
+at the sub-grid level and across neighbouring cells. The volume decomposition is now cubical,
 and most of the calculations on the fine mesh are {\tt openmp} parallel.
@@
 \input{PoissonSolver}
-\input{Accuracy} %pairwise and density force, 2-points stats.
 \input{ScalingPerformance}

### 2012-05-01 Conclusion.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

\section{Conclusion}

The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
compiling and running strategy is can be found at {\tt www.wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.

\section*{Acknowledgements}

### 2012-05-01 Extendedpp.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

@@
  To quantify the accuracy improvement versus computing time requirements, we performed the following test.
- We generate a set of initial conditions at a starting redshift of $z = XXX$, with a box size equal to $ XXX h^{-1}\mbox{Mpc}$,
- and with $XXX^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
+ We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
+ and with $128^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
  the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
  such that the only difference between different runs is the range of the pp force.
-{\bf (JD, insert here your P(k) plot and a description of what you observe, in terms of accuracy improvement and real time cpu usage...)}
+ Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
+ when extending  PM to P$^{3}$M first, and when adding successive layers of fine cell mesh where the pp force is extended.
+We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the fine grid calculations
+are more accurate as the distance increases. For this reason, it seems that a range of a few layers, between 3 and 6, suffices 
+to reduce most of the undesired NGP scatter.
+
+Extending the pp calculation comes at a price, since the number of operation scales as  $N^{2}$ in the sub-domain. 
+This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
+to evolved the particles to the final redshift. Table \ref{table:cpu_pp_ext} presents this usage.
+
+\begin{figure}
+  \begin{center}
+  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth,height=0.49\textwidth}
+  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
+    \label{fig:power}
+  \end{center}
+\end{figure}
+
+\begin{table}
+\begin{center}
+\caption{Scaling in {\small CPU} resources as a function of the range of the pp interaction.
+$N_{layers}$ refers to the number of fine mesh layers around a given cell, inside of which the force calculation
+is purely given by the pp contribution. }
+\begin{tabular}{|l|c|c|}
+\hline 
+             Type         & time (h)   \\
+                  \hline
+PM                         & 1.77 \\
+P$^{3}$M             & 2.09 \\
+\hline
+ $N_{layers}=1$ & 8.74 \\
+ $N_{layers}=2$ & 11.47\\
+ $N_{layers}=3$ & 14.30 \\
+ $N_{layers}=4$ & 18.87\\
+ $N_{layers}=5$ & 22.52\\
+ $N_{layers}=6$ & 29.62\\
+ $N_{layers}=7$ & 34.82\\
+ $N_{layers}=8$ & 47.00\\
+
+\hline \hline
+\end{tabular}
+\label{table:cpu_pp_ext}
+\end{center}
+\end{table}
+
+Another way to quantify the improvement of the calculation is to look at the halo mass function for these different runs.
+{\bf (To do...)}

### 2012-05-01 Introduction.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

@@
 and N-body codes have evolved both in performance and complexity. 
-Many codes have opted for a tree algorithm \citep{Gadjet, Gadjet2, TPM, GOTPM}, in which 
-the local resolution increases with the density of the matter field. 
+Many codes have opted for a tree algorithm, including Gadjet \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field.
 These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
 The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
-Unlike Hydra \citep{couchman1991} and the PM code by \cite{FerrelBertschinger1995},
+The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{FerrelBertschinger1995}.
  these are not designed to perform large scale PM calculations.
 
-{\bf (Describe briefly   Hydra, PM code by japanese, and others...)}
+{\bf ( PM code by japanese? Others codes I should mention? That of Hy Trac?)}
 
 {\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
-both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{TracPen2003},
+both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T},
 The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time

### 2012-05-01 ScalingPerformance.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

@@
 
-Real time versus size, bottle neck, extended pp, trials on scinet and ranger, 
-p3dfft,
+Real time versus size, bottle neck, description of the largest runs, etc.
+{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
+concerning the largest runs you did on Blue Gene? )}

### 2012-05-01 paper.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.

@@
 
-\title[{\small} CUBEP3M: High Performance P3M N-body code]{{\small} CUBEP3M: High Performance P3M N-body code}
+\title[{\small} CUBEP3M: High Performance P$^{3}$M N-body code]{{\small} CUBEP3M: High Performance P$^{3}$M N-body code}
 \author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
@@
 Among the principal changes, the Poisson solver includes particle-particles interactions
-at the sub-grid level and across neighbouring cells. The volume decomposition is now cubical,
-and most of the calculations on the fine mesh are {\tt openmp} parallel.
-Many utilities have been added and are briefly described, including a runtime halofinder
-and particle identification tags. 
+at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical.
+Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
 We discuss the structure of the code, its accuracy, and its scaling performance.
+In addition, many utilities have been added, including a runtime halofinder,
+particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
+and accuracy, when applicable .
+
 \end{abstract}
@@
 \begin{keywords}
-To specify...
+N-body simulations --- Large scale structure of Universe --- Dark matter
 \end{keywords}

### 2012-05-08 Halos_new.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).

\section{Runtime Halo Finder}
\label{sec:halo}

%Spherical overdensities, search algorithm, 
%provided halo information, halo bias, comparison to PS and ST.
%{\bf (Ilian, You can lead the way here...)}

We have implemented a halo finding procedure, which we have developed 
based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
In the interest of speed and efficiency the halo catalogues are constructed 
on-the-fly at a pre-determined list of redshifts. The halo finding is 
massively-parallel and threaded based on the CubeP$^3$M data structures 
discussed earlier. The code first builds the 
fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
proceeds to search for and record all local density maxima above certain
threshold (typically set to 100 above mean density) within the local 
sub-domain's physical volume (excluding the sub-domain buffer zones). It then 
uses parabolic interpolation on the density field to determine more precisely
the location of the maximum within the densest cell, and records the peak 
position and value. The halo center determined this way agrees closely with 
the center-of-mass of the halo particles discussed below.

Once the list of peak positions is generated, they are sorted from the highest 
to the lowest density value. Then each of the halo candidates is inspected 
independently, starting with the highest peak. The grid mass is accumulated 
in spherical shells of fine grid cells surrounding the maximum, until the 
mean density within the halo drops below a pre-defined overdensity cutoff 
(usually set to 178 in units of the mean, in accordance to the top-hat 
collapse model). As we accumulate mass we remove it from the mesh, so that no 
mass element is double-counted. This method is thus inappropriate for finding 
sub-halos as within this framework those are naturally incorporated in their 
host halos. Because the halos are found on a grid of finite-size cells, it is 
possible, especially for the low-mass halos, to overshoot the target overdensity.
When this occurs we use an analytical halo density profile to correct the 
halo mass and radius to the values corresponding to the target overdensity. 
This analytical density profile is given by Truncated Isothermal Sphere (TIS) 
profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
$\sim130$, and $1/r^2$ for lower overdensities. The TIS density profile has a
similar outer slope (the relevant one here) to the Navarro, Frenk and White 
\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
and matches well the virialization shock position given by the Bertschinger 
self-similar collapse solution \citep{1985ApJS...58...39B}.

Once the correct halo mass, radius and position are determined, we find all 
particles which are within the halo radius. Their positions and velocities are
used to calculate the halo center-of-mass, bulk velocity, internal velocity 
dispersion and the three angular momentum components, all of which are then 
included in the final halo catalogues. We also calculate the total mass of
all particles within the halo radius, also listed in the halo data. This mass
is very close, but  typically slightly lower, than the halo mass calculated 
based on the gridded density field. The particle center-of-mass corresponds 
very well to the halo centre found based on the gridded mass distribution.

{\bf II: do we need to say anything about having the option to use other
halo finders like AHF and FOF?}

{\bf Show some examples and comparisions to analytical fits and other halo 
finders.}

### 2012-05-08 PoissonSolver.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).

@@
 When matching the two  force kernels, it was realized that close to the cutoff region, the long range force is always on the low side, whereas 
-the short range force is scattered across the theoretical $1/r^2$ value. These behaviours are purely features of the CIC and NGP interpolation scheme 
+the short range force is scattered across the theoretical $1/r^2$ value. These behaviors are purely features of the CIC and NGP interpolation scheme 
 respectively. We identified a small range surrounding the cutoff length, in which we empirically adjust both kernels such as to improve 
@@
 
-In addition, the long range force is subject to a correction that fixes {\bf (what exactly fixes the LRCKCORR flag?)}.
+Finally, a choice must be done concerning the longest range of the force. Gravity can be either a) an accurate $1/r^2$ force, as far as the volume allows, 
+or b) modified to correctly match the periodicity of the boundary conditions. By default, the code is configured along to the second choice,
+which accurately models the growth of structure at very large scales. However, detailed studies of gravitational collapse would benefit 
+from the first settings.
 

### 2012-05-08 fnl.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).

The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
is extended to include non-Gaussian features of the ``local'' form,
$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
to the Bardeen potential $\Phi({\bf x})$ (see \cite{F_NL} for a review). 
We adopted the CMB convention,
in which $\Phi$ is calculated immediately after the matter-
radiation equality (and not at redshift $z=0$ as in the large scale
structure convention). For consistency, $\phi({\bf x})$ is normalized
to the amplitude of scalar perturbations inferred by CMB measurements
($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
before the inclusion of the matter transfer function, and the initial 
particle positions and velocities are finally computed from $\Phi({\bf x})$ 
according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.

### 2012-05-08 Halos.tex
More improvements
Updated references, Intro, scaling, halos with Ilian's input.

@@
-\section{Sperical Overdensity Halo Finder}
+\section{Runtime Halo Finder}
+\label{sec:halo}
 
-Spherical overdensities, search algorithm, 
-provided halo information, Comparison to PS and ST.
+%Spherical overdensities, search algorithm, 
+%provided halo information, halo bias, comparison to PS and ST.
+%{\bf (Ilian, You can lead the way here...)}
+
+We have implemented a halo finding procedure, which we have developed 
+based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
+In the interest of speed and efficiency the halo catalogues are constructed 
+on-the-fly at a pre-determined list of redshifts. The halo finding is 
+massively-parallel and threaded based on the CubeP$^3$M data structures 
+discussed earlier. The code first builds the 
+fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
+proceeds to search for and record all local density maxima above certain
+threshold (typically set to 100 above mean density) within the local 
+sub-domain's physical volume (excluding the sub-domain buffer zones). It then 
+uses parabolic interpolation on the density field to determine more precisely
+the location of the maximum within the densest cell, and records the peak 
+position and value. The halo center determined this way agrees closely with 
+the center-of-mass of the halo particles discussed below.
+
+Once the list of peak positions is generated, they are sorted from the highest 
+to the lowest density value. Then each of the halo candidates is inspected 
+independently, starting with the highest peak. The grid mass is accumulated 
+in spherical shells of fine grid cells surrounding the maximum, until the 
+mean density within the halo drops below a pre-defined overdensity cutoff 
+(usually set to 178 in units of the mean, in accordance to the top-hat 
+collapse model). As we accumulate mass we remove it from the mesh, so that no 
+mass element is double-counted. This method is thus inappropriate for finding 
+sub-halos as within this framework those are naturally incorporated in their 
+host halos. Because the halos are found on a grid of finite-size cells, it is 
+possible, especially for the low-mass halos, to overshoot the target overdensity.
+When this occurs we use an analytical halo density profile to correct the 
+halo mass and radius to the values corresponding to the target overdensity. 
+This analytical density profile is given by Truncated Isothermal Sphere (TIS) 
+profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
+$\sim130$, and $1/r^2$ for lower overdensities. The TIS density profile has a
+similar outer slope (the relevant one here) to the Navarro, Frenk and White 
+\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
+and matches well the virialization shock position given by the Bertschinger 
+self-similar collapse solution \citep{1985ApJS...58...39B}.
+
+Once the correct halo mass, radius and position are determined, we find all 
+particles which are within the halo radius. Their positions and velocities are
+used to calculate the halo center-of-mass, bulk velocity, internal velocity 
+dispersion and the three angular momentum components, all of which are then 
+included in the final halo catalogues. We also calculate the total mass of
+all particles within the halo radius, also listed in the halo data. This mass
+is very close, but  typically slightly lower, than the halo mass calculated 
+based on the gridded density field. The particle center-of-mass corresponds 
+very well to the halo centre found based on the gridded mass distribution.
+
+%Compared to Tinker
+%44% low for 50 particles
+%22% low for 400 particles
+%12% low for 1000 particles
+
+\begin{figure}%[ht]
+%  \vskip -0.5cm 
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
+  \end{center}
+  \caption{Simulated halo multiplicity function, 
+    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
+    CubeP$^3$M simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$ 
+    particles (red, solid). For reference we also show a widely-used 
+    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
+    \label{mf}}
+\end{figure}
+
+A sample halo mass function produced based on our inlined SO halo finder 
+at redshift $z=0$ from a simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$
+particles is shown in Figure~\ref{mf}. We compare our result to the 
+precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
+other widely-used fits like the one by Sheth and Tormen 
+\citep{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
+halo finders, the \citet{2008ApJ...688..709T} fit is based on a Spherical 
+Overdensity-determined halo masses, which are systematically different 
+from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
+making this fit a better base for comparison here. Results show excellent
+agreement, within $\sim10\%$ for all halos with masses corresponding to
+1000 particles or more. Lower-mass halos are somewhat under-counted compared
+to the \citet{2008ApJ...688..709T} fit, by $\sim20\%$ for 400 particles and 
+by $\sim40\%$ for 50 particles. This is due to the grid-based nature of our
+SO halo finder, which misses some of the low-mass halos. Using more sophisticated
+halo finders (available through post-processing due to their heavier memory
+footprint) recovers the expected mass function.
+
+{\bf II: do we need to say anything more about having the option to use other
+halo finders like AHF and FOF?}
+
+%{\bf Show some examples and comparisions to analytical fits and other halo finders.}

### 2012-05-08 Initial.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).
Updated references, Intro, scaling, halos with Ilian's input.

@@
 we have developed a non-Gaussian initial condition generator that we briefly describe in this section.
-{\bf Vincent, here you go!}
+
+The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
+is extended to include non-Gaussian features of the ``local'' form,
+$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
+\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
+to the Bardeen potential $\Phi({\bf x})$ (see \cite{F_NL} for a review). 
+We adopted the CMB convention,
+in which $\Phi$ is calculated immediately after the matter-
+radiation equality (and not at redshift $z=0$ as in the large scale
+structure convention). For consistency, $\phi({\bf x})$ is normalized
+to the amplitude of scalar perturbations inferred by CMB measurements
+($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
+before the inclusion of the matter transfer function, and the initial 
+particle positions and velocities are finally computed from $\Phi({\bf x})$ 
+according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.
+
+This code was tested by comparing simulations and theoretical predictions
+for the effect of local primordial non-Gaussianity on the halo mass 
+function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
+It has also been used to quantify the impact of local non-Gaussian initial
+conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
+2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
+ as well as the matter bispectrum \citep{2011arXiv1111.6966S}.

### 2012-05-08 Introduction.tex
Updated references, Intro, scaling, halos with Ilian's input.

@@
 One of the main field of application is the modelling of large scale structures, 
-which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
-\citep{WMAP5-7} of galaxy clustering \citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
-and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
-and collisionless dark matter occupy more than 95 per cent of the total energy density of the universe.
-In such a paradigm, pure N-body code are perfectly suited to describe the dynamics, as long as we
-understands where and how the baryonic fluid feeds back on the dark matter structure.
-The next generation of measurements aim at constraining the cosmological parameters at the per cent level, 
-and the theoretical understanding of the non-linear dynamics that govern structure formation heavily relies 
-on numerical simulations.
+which are driven by the sole force of gravity. Recent observations of the 
+cosmic microwave background \citep{WMAP5-7}, of galaxy clustering 
+\citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 
+2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
+and of supernovae redshift-distance relations all point towards a standard 
+model of cosmology, in which dark energy and collisionless dark matter occupy 
+more than 95 per cent of the total energy density of the universe. In such a 
+paradigm, pure N-body code are perfectly suited to describe the dynamics, as 
+long as baryonic physics is not very important, or at least we understand how 
+the baryonic fluid feeds back on the dark matter structure. The next generation 
+of measurements aim at constraining the cosmological parameters at the per cent 
+level, and the theoretical understanding of the non-linear dynamics that govern 
+structure formation heavily relies on numerical simulations.
 
-For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation scale can provide
-tight constraints on the dark energy equation of scale \citep{Seo2003,2005}.
-The most optimal estimates of the uncertainty requires the knowledge of the 
-matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
-of N-body simulations \citep{2005MNRAS.360L..82R, 2009ApJ...700..479T, 2011ApJ...726....7T}.
-For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
-by propagating photons in past light cones that are extracted from simulated density fields \citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
-{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
-accurate predictions/forecasts, etc.)}
+For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation 
+scale can provide tight constraints on the dark energy equation of scale 
+\citep{Seo2003,2005}. The most optimal estimates of the uncertainty requires the 
+knowledge of the matter power spectrum covariance matrix, which is only accurate 
+when measured from a large sample of N-body simulations \citep{2005MNRAS.360L..82R, 
+2009ApJ...700..479T, 2011ApJ...726....7T}. For the same reasons, the most accurate 
+estimates of weak gravitational lensing signal is obtained by propagating photons 
+in past light cones that are extracted from simulated density fields 
+\citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
+Another area where large-scale N-body simulations have in recent years been 
+instrumental are in simulations of early cosmic structures and reionization 
+\citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
+2011arXiv1107.4772I}. The reionization process is primarily driven by low-mass 
+galaxies, which for both observational and theoretical reasons, need to be resolved 
+in fairly large volumes, which demands simulations with a very large dynamic range.   
+%{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
+%accurate predictions/forecasts, etc.)}
 
@@
 
-With the advent of large computing facilities, parallel coding has now become common practice, 
-and N-body codes have evolved both in performance and complexity. 
-Many codes have opted for a tree algorithm, including Gadjet \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field.
+With the advent of large computing facilities, parallel coding has now become 
+common practice, and N-body codes have evolved both in performance and complexity. 
+Many codes have opted for a tree algorithm, including Gadget \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field.
 These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
@@
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
- \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
+ \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, 
+formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
+observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
+reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}, {\bf (anymore cubep3m papers to cite?)},  
 and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 

### 2012-05-08 ScalingPerformance.tex
Updated references, Intro, scaling, halos with Ilian's input.

@@
 Real time versus size, bottle neck, description of the largest runs, etc.
-{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
+{\bf (Hugh, do you have anything you would like to share 
 concerning the largest runs you did on Blue Gene? )}
+
+\begin{figure*}%[ht]
+%  \vskip -0.5cm 
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_curie.eps}
+    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_new.eps}
+%  \vskip -1.2cm 
+%  \vskip -0.5cm 
+  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
+    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
+    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
+    in each case. Dashed line indicates the ideal weak 
+    scaling. The data is listed in Table \ref{summary_scaling_table}.
+    \label{scaling}
+% \vskip -0.9cm 
+}
+\end{center}
+\end{figure*}
+
+\begin{table*}%[ht]
+  \vskip -0.5cm 
+  \begin{center}
+\caption{Scaling of {\small CUBEP$^3$M} on Curie. Speedup is 
+scaled to the smallest run.}
+\label{summary_scaling_table}
+\begin{tabular}{@{}|llllll|}
+\hline
+number of cores & speedup & ideal speedup & absolute timing (min) & 
+$N_{\rm particles}$& box size (Mpc$/h$)
+\\[2mm]\hline
+%\hline
+32  &  1.00 & - &3.2 & $256^3$ & 256\\
+256  & 7.21 & 8 &3.55 & $512^3$  & 512\\
+864  & 25.63 & 27 &4.8 & $864^3$  & 864\\
+2048  & 61.87 & 64 &26.48 & $2048^3$ & 2048 \\
+\hline
+\end{tabular}
+\caption{Scaling of the CubeP$^3$M code for large number of cores 
+(on Ranger). Speedup is scaled to the smallest run.}
+\label{summary_scaling_table2}
+\begin{tabular}{@{}|llllll|}
+\hline
+number of cores & speedup & ideal speedup & absolute timing (min) & 
+$N_{\rm particles}$& box size (Mpc$/h$)
+\\[2mm]\hline
+%\hline
+864    & 1.00  & -    &258   & $1728^3$  & 6.3\\
+4096   & 4.53  & 4.74 &320   & $3072^3$  & 11.4\\
+10976  & 9.78  & 12.7 &845   & $5488^3$  & 20\\
+21952  & 14.73 & 25.4 &561   & $5488^3$  & 20 \\
+\hline
+\end{tabular}
+\end{center}
+  \vskip -0.7cm 
+\end{table*}
+
+%the Ranger system at the Texas Supercomputing
+%Center (in top 20 in the world) which is a SunBlade x6420 
+%with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
+% ``Barcelona'' processors and Infiniband networking. It has 
+%a total of 62976 computing cores and 125952 GB of total 
+%memory. Its nodes consist of 4 Quad Core processors and 32 GB 
+%of shared RAM. For efficiency reasons (local memory access) we 
+%typically use smaller MPI 'nodes' consisting of one Quad Core 
+%processor and 8 GB of RAM.
+
+The parallel algorithm of {\small CUBEP$^3$M} is designed for ``weak'' 
+scaling, i.e. if the number of cores and the problem size 
+increase in proportion to each other, then for ideal scaling the 
+wall-clock time should remain the same, in contrast to ``strong'' 
+scaling, whereby the same problem solved on more cores should take 
+proportionately less wall-clock time. This weak scaling requirement 
+is dictated by the problems we are typically aiming towards (very 
+large and computationally-intensive) and our goals, which are to 
+address such large problems in the most efficient way, rather than 
+for the least wall-clock time. Furthermore, there is no explicit 
+load balancing, thus the code is most efficient if the sub-domains
+contain roughly equal number of particles, which is true for most
+cosmological-size volumes, but not for e.g. simulations of a single
+highly-resolved galaxy.
+
+%Since the current Curie is using Intel Nehalem architecture, while 
+%the thin Curie nodes will use the newer Westermere Intel architecture, 
+%we also show the scaling of our code on the Westermere-based Lonestar 
+%computer at the Texas Advanced Computing Centre
+
+The scaling was tested with a dedicated series 
+of simulations with increasing size and number of cores on the `fat'' 
+(i.e. large-memory) nodes of the supercomputers Curie at TGCC in France. 
+Our results are shown in Fig. \ref{scaling} and in 
+Table \ref{summary_scaling_table}. For appropriate direct comparison,
+all simulations were performed using the same particle mass 
+($M_{\rm particle}=1.07\times10^{11}M_\odot$) and force resolution 
+(softening length 50 $h^{-1}$kpc). The box sizes used range from 256 $h^{-1}$Mpc
+to 2048 $h^{-1}$Mpc, and the number of particles from $256^3$ to $2048^3$.
+Simulations were run on 32 up to 2048 computing cores starting from 
+redshift $z=100$, and eveolving until $z=0$. Our results show excellent scaling, within 
+$\sim3\%$ of the ideal one, for up to 2048 cores.
+
+We have also ran {\small CUBEP$^3$M} on a much larger number of cores, 
+from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
+particles on Ranger at the Texas Advanced Computing Centre (SunBlade 
+x6420 with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
+``Barcelona'' processors and Infiniband networking, a total of 62,976 
+computing cores and 125,952 GB of total memory. Its shared-memory nodes 
+consist of 4 Quad Core processors and 32 GB of RAM.) and JUROPA 
+at J\"ulich Supercomputing Centre in Germany (Intel Xeon X5570 
+(Nehalem-EP) quad-core, 2.93 GHz, Infiniband networking, a total of 17,664
+computing cores and 52 TB of total memory. Its shared-memory nodes 
+consist of 2 Quad Core processors and 24 GB of RAM). Since it is not 
+practical to perform dedicated scaling tests on such large number 
+computing cores, we instead use data extracted from production runs, 
+as listed in Table \ref{summary_scaling_table2}. We have found the 
+code to scale quite well, within 1.5\% of ideal up to 4,096 cores. 
+For very large number of cores (10,976) the scaling is slightly worse,
+but still within $\sim20\%$ of ideal, due to increased communication 
+costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
+and load balancing issues. These first three Ranger runs were performed 
+with 4 MPI processes and 4 threads per Ranger node ('4way') (using the 
+specially-provided {\it tacc\_affinity} NUMA script to bind the memory 
+usage to local core, thus ensuring memory affinity. This is important 
+because the 32 GB RAM/node on Ranger (and other computers with 
+multi-processor sockets) are actually not equal-access and the local 
+8 GB memory of each processor has much shorter access time).
+
+Furthermore, due to the very strong clustering of structures at those small 
+scales, some of the cuboid sub-domains came to contain well above the 
+average number of particles, thereby requiring more memory per MPI process
+in order to run. As a consequence, throughout most of the evolution the 
+largest two of these simulations were run with 4096 and 21,952 cores and 
+with only 2 MPI processes and 8 threads per Ranger node (2way), which on 
+Ranger allows using up to 16 GB of RAM per MPI process. In order to insure 
+local memory affinity a special NUMA control script (which we called {\it 
+tacc\_affinity\_2way}) 
+was developed for us by TACC staff and was used by us to run more efficiently 
+in this mode. However, this not-fully-local memory access does affect the 
+performance somewhat, as does the imperfect load balancing in such situations,
+as seen by the rightmost point on the scaling plot. The scaling in this case
+is 42\% below the ideal, although we note that we still get $\sim1.5$ speedup
+from doubling the core count, even given these issues. Overall the
+code scaling performance is thus quite satisfactory and it clearly 
+scales very well to extremely large number of cores and can therefore 
+do very large problems efficiently, utilizing the next generation 
+Petascale systems.
+
+Finally, we note that several special fixes had to be developed by TACC 
+and JUROPA staff in order for our largest runs work properly and to be 
+able to use the software libraries we required (MPICH and FFTW), as those 
+latter exhibited serious problems when applied to runs of such 
+unprecedented size.

### 2012-05-08 Systematics.tex
Updated references, Intro, scaling, halos with Ilian's input.

--- 2012-04-12 Systematics.tex
@@
+\section{Accuracy}
+\label{sec:accuracy}
+
+One of the most accurate way to assess the simulation's accuracy at evolving particles
+is to measure the density power spectrum at late time, and compare to non-linear prediction. 
+For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
+\begin{eqnarray}
+\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
+\label{eq:power}
+\end{eqnarray}
+where the angle bracket corresponds to an ensemble (or volume) average.
+We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
+which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
+We observe that the agreement with the non-linear prediction of \cite{Lewis} is at the per cent level over a large dynamical range.
+The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
+in the initial conditions.
+
+\begin{figure*}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/power_highres.eps}
+  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
+  from $4000^3$ particles. The simulation ran on 4000 cores on RANGER.
+    \label{fig:power_highres}}
+\end{center}
+\end{figure*}
+
+(Show plots of pairwise force (transverse and radial), density force, 
+mention the work on covariance matrix by Ngan, Harnoisetal.)
+
 \section{Systematics}
-Random shift, resolution, initial redshift
+\label{sec:systematics}
+
+To discuss :
+
+2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
+{\bf JD, here you could discuss the ra\_max tempering...}
+
+3) limits of resolution caused by softening length,
+
+4) Poisson shot noise if too few particles or unrelaxed systems,
+
+5) finite box size effects (with beat coupling?)
+
+The biggest problem with the pp force calculation is that it 
+is anisotropic and depends on the location of the fine mesh with respect 
+to the particles. An example are two particles on either side of a grid 
+cell boundary. These particles will experience the 1-grid cell separation
+NGP mesh force, but if the mesh were shifted such that they were
+within the same cell, they would experience the (larger) pp force instead. 
+This effect is especially pronounced at the early stages of the simulation where
+the density is more homogeneous, and leads to mesh artefacts appearing
+in the density field. In order to minimize this systematic effect, 
+we randomly shift the particle distribution relative to the mesh by a small
+amount -- up to 2 fine grid cells in magnitude -- in each
+dimension and at each time step.  This adds negligible computational
+overhead as it can be applied during the particle position update
+and suppresses the mesh behavior over multiple time steps.
+It is possible to shift back the particles at the end of each time steps,
+which prevents a random drift of the whole population.
+This is convenient if one needs to correlate the initial and final positions of the particles.
+
+We ensure that, on average, this solution balances out the mesh feature,
+by tuning the NGP force kernel such as to provide  force as unbalanced as possible at grid cell distances.
+This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.

### 2012-05-11 Conclusion.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 \section{Conclusion}
+
+This paper describes {\small CUBEP3M}, a P$^3$M N-body public code that is massively parallel, memory local, 
+and that scales well to 20,000 cores, thus raising the limits of the cosmological problem size one can handle.
+We summarize the code structure, the two-mesh Poisson solver algorithm, we present scaling and systematic tests
+that have been performed. We also describe a few utilities and extensions that come with the public release, 
+including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
 
@@
 \section*{Acknowledgements}
+
+TI was supported by The Southeast Physics
+Network (SEPNet) and the Science and Technology Facilities Council
+grants ST/F002858/1 and ST/I000976/1. The authors acknowledge the
+TeraGrid and the Texas Advanced Computing Center (TACC) at The
+University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
+providing HPC and visualization resources that have contributed to the
+research results reported within this paper and Partnership for Advanced
+Computing in Europe (PRACE) grant 2010PA0442 to ITI used for the code
+scaling studies.

### 2012-05-11 Halos.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 on-the-fly at a pre-determined list of redshifts. The halo finding is 
-massively-parallel and threaded based on the CubeP$^3$M data structures 
-discussed earlier. The code first builds the 
+massively-parallel and threaded based on the main {\small CUBEP3M} data structures 
+discussed in section \ref{sec:structure}. The code first builds the 
 fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
@@
 the location of the maximum within the densest cell, and records the peak 
-position and value. The halo center determined this way agrees closely with 
-the center-of-mass of the halo particles discussed below.
+position and value. The halo centre determined this way agrees closely with 
+the centre-of-mass of the halo particles discussed below.
 
@@
 particles which are within the halo radius. Their positions and velocities are
-used to calculate the halo center-of-mass, bulk velocity, internal velocity 
+used to calculate the halo centre-of-mass, bulk velocity, internal velocity 
 dispersion and the three angular momentum components, all of which are then 
@@
 is very close, but  typically slightly lower, than the halo mass calculated 
-based on the gridded density field. The particle center-of-mass corresponds 
+based on the gridded density field. The particle centre-of-mass corresponds 
 very well to the halo centre found based on the gridded mass distribution.
@@
     $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
-    CubeP$^3$M simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$ 
+    simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
     particles (red, solid). For reference we also show a widely-used 
@@
 at redshift $z=0$ from a simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$
-particles is shown in Figure~\ref{mf}. We compare our result to the 
+particles is shown in Fig. \ref{mf}. We compare our result to the 
 precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
-other widely-used fits like the one by Sheth and Tormen 
-\citep{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
-halo finders, the \citet{2008ApJ...688..709T} fit is based on a Spherical 
-Overdensity-determined halo masses, which are systematically different 
+other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
+halo finders, the \citet{2008ApJ...688..709T} fit is based on the
+SO search algorithm, whose masses are systematically different 
 from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
 making this fit a better base for comparison here. Results show excellent
-agreement, within $\sim10\%$ for all halos with masses corresponding to
+agreement, within $\sim10$ per cent for all halos with masses corresponding to
 1000 particles or more. Lower-mass halos are somewhat under-counted compared
-to the \citet{2008ApJ...688..709T} fit, by $\sim20\%$ for 400 particles and 
-by $\sim40\%$ for 50 particles. This is due to the grid-based nature of our
+to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
+by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
 SO halo finder, which misses some of the low-mass halos. Using more sophisticated
@@
 
+A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
+with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
+
+\begin{figure}%[ht]
+%  \vskip -0.5cm 
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/bias.eps}
+  \end{center}
+  \caption{Halo bias, measured at $z=0$ from a $4000^3$ particles simulations with a side length of $3.2 h^{-1}$Gpc.
+  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
+  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2,000 and so on.
+    \label{fig:halo}}
+\end{figure}
+
 %{\bf Show some examples and comparisions to analytical fits and other halo finders.}

### 2012-05-11 Introduction.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
-to be stored at a given time, allowing for {\small OPENMP}parallel computation.
-This this setup enable the code to evolve very large cosmological systems both rapidly and accurately, on relatively modest clusters.
+to be stored at a given time, allowing for {\small OPENMP} parallel computation.
+This this setup enable the code to evolve large cosmological systems both rapidly and accurately, on relatively modest clusters.
 One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
@@
 as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
-until it reaches the hard limit of a single cell layer.  With this enhancement came a new name, {\small CUBEPM},
-which soon after incorporated particle-particle (pp) interactions at the sub-grid level, and the code was renamed
-{\small CUBEP3M}. It now includes a significant number of new features since MPT : the pp force
+until it reaches the hard limit of a single cell layer.  With this enhancement, the code name was changed to {\small CUBEPM},
+which soon after incorporated particle-particle (pp) interactions at the sub-grid level. 
+The code was finally renamed {\small CUBEP3M}, and now includes a significant number of new features: the pp force
 can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
-a runtime halofinder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features
-{\bf (anything else?)}.
+a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features {\bf (anything else since PMFAST?)}.
 
@@
 observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
-reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}, {\bf (anymore cubep3m papers to cite?)},  
-and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
+reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F},  
+and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
+
+Since {\small CUBEP3M} is not a new code, the accuracy and systematic tests were performed by different groups, on different machines,
+and with different configurations. It is not an ideal situation in which to quantify the performance, and each test must be viewed as a
+separate measurement that quantifies a specific aspect of the code. 
+We have tried to keep to a minimum the number of such different runs, and although the detailed numbers might vary across different runs, 
+the general trends are common to all of them.
+
 The paper is structured as follow: XXXX

### 2012-05-11 PoissonSolver.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
-to section 2 of MPT for more details. In cubep3m, the mass default assignment scheme are
+to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
 a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
@@
 for pairs closer to a tenth of a grid cell.
+
+The code units inherit from \citep{2004NewA....9..443T}, which we summarized here for completeness.
+In simulation units, the mean comoving mass density is set to unity, the length of a fine grid cell is set to one,
+and the Newton gravitational  constant $G$ is set to $1/(6 \pi a)$, where $a$ is the scale factor.
+With these three choices, one can transform any physical units into simulation units and vice versa.
 
@@
 \begin{eqnarray}
- F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) {\bf  \omega}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
+ F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
 \label{eq:poisson_solution_force}
@@
 \begin{eqnarray}
-{\bf  \omega}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
+\mbox{\boldmath $\omega$}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
 \end{eqnarray}
@@
 \begin{eqnarray}
-{\bf \omega}_{s}({\bf r}) = \begin{cases} {\bf \omega}_{F}(r) -  {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
+\mbox{\boldmath $\omega$}_{s}(r) = \begin{cases} \mbox{\boldmath $\omega$}_{F}(r) -  \mbox{\boldmath $\beta$}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 0 & \mbox{otherwise} 
@@
 \begin{eqnarray}
-{\bf \omega}_{l}(r) = \begin{cases} {\bf \beta}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
- {\bf\omega}_{F}(r)  &\mbox{otherwise} 
+\mbox{\boldmath $\omega$}_{l}(r) = \begin{cases} \mbox{\boldmath $\beta$}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
+ \mbox{\boldmath $\omega$}_{F}(r)  &\mbox{otherwise} 
 \end{cases}
 \end{eqnarray}
-The vector $ {\bf \beta}(r)$ is related to the fourth order polynomial that is used in the potential case by
- $ {\bf \beta} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
+The vector $\mbox{\boldmath $\beta$}(r)$ is related to the fourth order polynomial that is used in the potential case by
+ $ \mbox{\boldmath $\beta$} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
  and we get
   \begin{eqnarray}
-   {\bf \beta}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
+   \mbox{\boldmath $\beta$}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
   \end{eqnarray}
@@
 respectively. We identified a small range surrounding the cutoff length, in which we empirically adjust both kernels such as to improve 
-the match. Namely, for $14 \le r \le 16$, ${\bf \omega}_{s}({\bf r}) \rightarrow {\bf \omega}_{s}({\bf r})*0.985$,
-and for  $12 \le r \le 16$, ${\bf \omega}_{l}({\bf r}) \rightarrow {\bf \omega}_{l}({\bf r})*1.2$.
+the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow \mbox{\boldmath $\omega$}_{s}({ r})*0.985$,
+and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow \mbox{\boldmath $\omega$}_{l}({ r})*1.2$.
 The two fudge factors were found by performing force measurements on two particles randomly placed in the volume.
@@
 which accurately models the growth of structure at very large scales. However, detailed studies of gravitational collapse would benefit 
-from the first settings.
+from the first settings. {\bf (Hugh, are the above two paragraphs correct?)}
 
-{\bf (Anything else here?)}
+{\bf Least square kernel?}

### 2012-05-11 ScalingPerformance.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 \section{Scaling Performances}
-
-Real time versus size, bottle neck, description of the largest runs, etc.
-{\bf (Hugh, do you have anything you would like to share 
-concerning the largest runs you did on Blue Gene? )}
 
@@
 %  \vskip -0.5cm 
-  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
+  \caption{Scaling of {\small CUBEP3M} on Curie fat nodes (left) and 
     on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
@@
   \begin{center}
-\caption{Scaling of {\small CUBEP$^3$M} on Curie. Speedup is 
+\caption{Scaling of {\small CUBEP3M} on Curie. Speedup is 
 scaled to the smallest run.}
@@
 number of cores & speedup & ideal speedup & absolute timing (min) & 
-$N_{\rm particles}$& box size (Mpc$/h$)
+$N_{\rm particles}$& box size ($h^{-1}$Mpc)
 \\[2mm]\hline
@@
 \end{tabular}
-\caption{Scaling of the CubeP$^3$M code for large number of cores 
+\caption{Scaling of  {\small CUBEP3M} for large number of cores 
 (on Ranger). Speedup is scaled to the smallest run.}
@@
 Simulations were run on 32 up to 2048 computing cores starting from 
-redshift $z=100$, and eveolving until $z=0$. Our results show excellent scaling, within 
+redshift $z=100$, and evolving until $z=0$. Our results show excellent scaling, within 
 $\sim3\%$ of the ideal one, for up to 2048 cores.
@@
 Finally, we note that several special fixes had to be developed by TACC 
-and JUROPA staff in order for our largest runs work properly and to be 
+and JUROPA staff in order for our largest runs to work properly and to be 
 able to use the software libraries we required (MPICH and FFTW), as those 
@@
 unprecedented size.
+
+Real time versus size, bottle neck, description of the largest runs, etc.
+{\bf (Hugh, do you have anything you would like to share 
+concerning the largest runs you did on Blue Gene? )}

### 2012-05-11 Structure.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
-only the head processor is active. Other libraries such as {\small P3DFFT} or  Intel {\small MKL-$\alpha$} currently permit this extra level of parallelization,
+only the head processor is active. Other libraries such as {\small P3DFFT} \citep{p3dfft} or  Intel {\small MKL-$\alpha$} \citep{fft_mkl} currently permit this extra level of parallelization,
 and it is our plan to migrate to one of these in the near future.
@@
 and writes them to file.
-Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time {\bf (is this correct?)}.
+Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time.
 The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 

### 2012-05-11 Systematics.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 
-One of the most accurate way to assess the simulation's accuracy at evolving particles
+One of the most reliable way to assess the simulation's accuracy at evolving particles
 is to measure the density power spectrum at late time, and compare to non-linear prediction. 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/power_highres.eps}
+    \includegraphics[width=5.2in]{graphs/power_highres.eps}
   \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
-  from $4000^3$ particles. The simulation ran on 4000 cores on RANGER.
+  from $4000^3$ particles. The simulation ran on 4000 cores on Ranger.
     \label{fig:power_highres}}
@@
 \label{sec:systematics}
-
-To discuss :
-
-2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
-{\bf JD, here you could discuss the ra\_max tempering...}
-
-3) limits of resolution caused by softening length,
-
-4) Poisson shot noise if too few particles or unrelaxed systems,
-
-5) finite box size effects (with beat coupling?)
 
@@
 overhead as it can be applied during the particle position update
-and suppresses the mesh behavior over multiple time steps.
+and suppresses the mesh behaviour over multiple time steps.
 It is possible to shift back the particles at the end of each time steps,
@@
 This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.
+We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
+since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
+As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
+a choice that  must be carefully balanced.
+
+At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
+rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
+If the expansion jump is too large, the size of the residual errors can become significant, and one can observe, for instance,
+a growth of structure that does not match the predictions of  linear theory even at the largest scales.
+One therefore needs to choose a maximum step size. This is controlled by $r_{max}$, which is the fractional step size,
+$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  It is possible to reduce this number and improve significantly 
+the accuracy of the code, at the cost of increasing the total number of time steps. Fig. \ref{fig:ra_max}
+shows how $r_{max}$ affects the code. It is a comparison of late time power spectra of density fields that originate from the same initial conditions, 
+and used the same random seeds to control the fine mesh shift (mentioned above). We observe that XXXXX.
+The {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.
+
+a good compromise seems to be XXXXX.
+
+\begin{figure}%[ht]
+  \begin{center}
+    %\includegraphics[width=3.2in]{graphs/power_highres.eps}
+  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $XXX h^{-1}\mbox{Mpc}$ per side,
+  from $256^3$ particles and a starting redshift of $z=200$. The different curves show different values of $r_{max}$. 
+  The resources required to run these simulations increase rapidly as $r_{max}$ decreases.    \label{fig:ra_max}}
+\end{center}
+\end{figure}
+
+\begin{table}
+\begin{center}
+\caption{Scaling in {\small CPU} resources as a function of the value of $r_{max}$. The tests were performed 
+on the CITA Sunnyvale cluster, and general trends could vary slightly on other machines.}
+\begin{tabular}{|l|c|c|}
+\hline 
+$r_{max}$         & time (h)   \\                 
+\hline
+ $0.1$ & 8.74 \\
+ $0.05$ & 11.47\\
+ $0.01$ & 14.30 \\
+ $0.005$ & 18.87\\
+ $0.002$ & 22.52\\
+ $0.001$ & 29.62\\
+\hline
+\end{tabular}
+\label{table:ra_max}
+\end{center}
+\end{table}
+
+To discuss :
+
+2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
+{\bf JD, here you could discuss the ra\_max tempering...}
+
+3) limits of resolution caused by softening length,
+
+4) Poisson shot noise if too few particles or unrelaxed systems,
+
+5) finite box size effects (with beat coupling?)

### 2012-05-11 paper.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).
Updated references, Intro, scaling, halos with Ilian's input.
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

@@
 
-\title[{\small} CUBEP3M: High Performance P$^{3}$M N-body code]{{\small} CUBEP3M: High Performance P$^{3}$M N-body code}
-\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
+\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
+\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
-Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
+Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk} and Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}\\
 %\footnotemark[1]\thanks{This file has been amended to
@@
 $^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
-$^{3}$Department of XXX\\
-$^{4}$Department of XXX\\
-$^{5}$Department of XXX}
+$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
+$^{4}$Department of XXX}
 
@@
 \bibliographystyle{mn2e}
-\bibliography{mybib3}{}
+\bibliography{mybib3_new}{}
 %\bibliographystyle{amsplain}

### 2012-05-15 Extendedpp.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
  The case where all three coordinates are equal is already calculated in the standard configuration of the code.
+ To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
+ plot, analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
+ We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
+ The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
+ which are now at least 5 time smaller than in the default p$^{3}$m algorithm.
+
+ \begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
+  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
+  with the pp force extended to six fine mesh layers.
+  Particles in the same cell follow the exact curve, then we observe a large scatter at 
+  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
+    \label{fig:den_force_ppext6}}
+\end{center}
+\end{figure}
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
+  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions,
+  with the pp force extended to six fine mesh layers.
+    \label{fig:den_force_fracErr_ppext6}}
+\end{center}
+\end{figure}
 
@@
   \begin{center}
-  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth,height=0.49\textwidth}
+  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
   \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
@@
 \hline
- $N_{layers}=1$ & 8.74 \\
- $N_{layers}=2$ & 11.47\\
- $N_{layers}=3$ & 14.30 \\
- $N_{layers}=4$ & 18.87\\
- $N_{layers}=5$ & 22.52\\
- $N_{layers}=6$ & 29.62\\
- $N_{layers}=7$ & 34.82\\
- $N_{layers}=8$ & 47.00\\
+$N_{layers}$       &          \\
+\hline
+ $1$ & 8.74 \\
+ $2$ & 11.47\\
+ $3$ & 14.30 \\
+ $4$ & 18.87\\
+ $5$ & 22.52\\
+ $6$ & 29.62\\
+ $7$ & 34.82\\
+ $8$ & 47.00\\
 

### 2012-05-15 Initial.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
 \phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
-to the Bardeen potential $\Phi({\bf x})$ (see \cite{F_NL} for a review). 
+to the Bardeen potential $\Phi({\bf x})$ (see \cite{2004PhR...402..103B} for a review). 
 We adopted the CMB convention,
@@
  as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
+Fig. \ref{fig:init} shows the power spectrum of two realizations with the same initial power spectrum, 
+one of which was started of with non-Gaussian initial conditions.
+We see that the power spectra agree at the sub-per cent level, as expected {\bf (?????)}
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50.eps}
+  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
+  from $4000^3$ particles.
+The two curves represent two realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator.
+The to curves differ at the sub-per cent level, as seen in the top panel.
+    \label{fig:init}}
+\end{center}
+\end{figure}

### 2012-05-15 Introduction.tex
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
 which are driven by the sole force of gravity. Recent observations of the 
-cosmic microwave background \citep{WMAP5-7}, of galaxy clustering 
+cosmic microwave background \citep{2009ApJS..180..330K,2011ApJS..192...18K}, of galaxy clustering 
 \citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 
-2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
+2010MNRAS.401.1429D} of weak gravitational lensing \citep{2012AAS...21913001H, 2009ApJ...703.2232S}
 and of supernovae redshift-distance relations all point towards a standard 
-model of cosmology, in which dark energy and collisionless dark matter occupy 
+model of cosmology, in which dark energy and collision-less dark matter occupy 
 more than 95 per cent of the total energy density of the universe. In such a 
@@
 scale can provide tight constraints on the dark energy equation of scale 
-\citep{Seo2003,2005}. The most optimal estimates of the uncertainty requires the 
+\citep{Eisenstein:2005su,2006PhRvD..74l3507T, 2007MNRAS.381.1053P,2009arXiv0902.4680S}. 
+The most optimal estimates of the uncertainty requires the 
 knowledge of the matter power spectrum covariance matrix, which is only accurate 
@@
 common practice, and N-body codes have evolved both in performance and complexity. 
-Many codes have opted for a tree algorithm, including Gadget \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field.
+Many codes have opted for a tree algorithm, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field.
 These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
 The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
-The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{FerrelBertschinger1995}.
+The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{1995astro.ph..3042F}.
  these are not designed to perform large scale PM calculations.
 
-{\bf ( PM code by japanese? Others codes I should mention? That of Hy Trac?)}
+{\bf ( PM code by japanese? Others codes I should mention?)}
 
@@
 It has already been involved in a number of scientific applications over the last few years,
-spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
- \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, 
+spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
+ (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 (in preparation)), 
 formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},

### 2012-05-15 PID.tex
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

--- 2012-04-12 PID.tex
@@
-\section{Other tools}
+\subsection{Particle identification tags}
+\label{PID}
 
-Particle ID, initial condition generator, projections, power spectrum code,  various expansions
+A system of particle identification can be turned on, which basically allows to track each particle's trajectory
+between checkpoints. Such a tool is useful for a number of applications, from reconstruction of halo merging history to tracking individual particle
+trajectories.
+The particle tag system has been implemented as an array of double integers, {\tt PID}, 
+and assigns a unique integer to each particle during the initialization stage. The location of the tag on the {\tt PID} array 
+matches the location of the corresponding particle on the {\tt xv} array, hence it acts as if the latter array had an extra dimension.
+The location change only when particles exist the local volume, in which case the tag is sent along adjacent nodes 
+with the particle in the {\tt pass\_particle} subroutine, and deleted particles result in deleted flags.
+Similarly to the phase space array, the {\tt PID} array gets written to file at each particle checkpoint.

### 2012-05-15 Structure.tex
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
 do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
-only the head processor is active. Other libraries such as {\small P3DFFT} \citep{p3dfft} or  Intel {\small MKL-$\alpha$} \citep{fft_mkl} currently permit this extra level of parallelization,
-and it is our plan to migrate to one of these in the near future.
+only the head processor is active\footnote{Other libraries such as {\small P3DFFT}({\tt http://code.google.com/p/p3dfft/}) currently permit
+ an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.
 

### 2012-05-15 Systematics.tex
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
 which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
-We observe that the agreement with the non-linear prediction of \cite{Lewis} is at the per cent level over a large dynamical range.
+We observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
 The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
@@
 
-(Show plots of pairwise force (transverse and radial), density force, 
-mention the work on covariance matrix by Ngan, Harnoisetal.)
+The actual force of gravity on the p$^3$m algorithm is presented in Fig. \ref{fig:den_force_ppext0}.
+This is a force versus distance plot, where the force is calculated in two steps. 
+We first compute the force on each particles of a given distribution as done at every time step.
+We then remove one particle, compute the force again on all particles, and record on a file the 
+force difference between the two calculations and the distance to the `hole'.
+
+Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
+It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value.
+{\bf (I would like to mention why we opted for NGP.
+Can't we do p3m with CIC on the fine grid?
+)}
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
+  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law.
+  Particles in the same cell follow the exact curve, then we observe a large scatter at 
+  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
+    \label{fig:den_force_ppext0}}
+\end{center}
+\end{figure}
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
+  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions.
+    \label{fig:den_force_fracErr}}
+\end{center}
+\end{figure}
 
@@
 \label{sec:systematics}
+
+\subsection{Mesh force at grid distances}
 
@@
 
+\subsection{Constraining redshift jumps}
+
 At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
@@
 a growth of structure that does not match the predictions of  linear theory even at the largest scales.
-One therefore needs to choose a maximum step size. This is controlled by $r_{max}$, which is the fractional step size,
-$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  It is possible to reduce this number and improve significantly 
-the accuracy of the code, at the cost of increasing the total number of time steps. Fig. \ref{fig:ra_max}
-shows how $r_{max}$ affects the code. It is a comparison of late time power spectra of density fields that originate from the same initial conditions, 
-and used the same random seeds to control the fine mesh shift (mentioned above). We observe that XXXXX.
-The {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.
+One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
+$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  Generally, a simulation should start at a redshift high enough that
+the initial dimensionless power spectrum is well under unity at all scales. This ensures that the Zel'dovich approximation
+ holds at the pre cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
+ truncation error will be significant at the early time steps.
 
-a good compromise seems to be XXXXX.
+It is possible to reduce this effect, and thereby improve significantly 
+the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
+Fig.
+\ref{fig:ra_max} shows a comparison of late time power spectra of density fields that originate from the same initial conditions, and used the same random seeds to control the fine mesh shifts (mentioned above), but the value of $r_{max}$ was modified at each run.
+We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
+allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
+This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
+test runs in order to optimize a given configuration.  
+As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
+In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the gain in time is more than a factor of four.
 
@@
   \begin{center}
-    %\includegraphics[width=3.2in]{graphs/power_highres.eps}
-  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $XXX h^{-1}\mbox{Mpc}$ per side,
+    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
+  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $500 h^{-1}\mbox{Mpc}$ per side,
   from $256^3$ particles and a starting redshift of $z=200$. The different curves show different values of $r_{max}$. 
@@
 \hline
- $0.1$ & 8.74 \\
- $0.05$ & 11.47\\
- $0.01$ & 14.30 \\
- $0.005$ & 18.87\\
- $0.002$ & 22.52\\
- $0.001$ & 29.62\\
+ $0.1$ & 1.46 \\
+ $0.06$ & 1.48\\
+ $0.01$ & 1.67 \\
+ $0.006$ & 1.91\\
+ $0.003$ & 2.83 \\
+ $0.002$ & 4.13\\
+ $0.001$ & 8.15\\
 \hline
@@
 
-To discuss :
-
-2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
-{\bf JD, here you could discuss the ra\_max tempering...}
-
-3) limits of resolution caused by softening length,
-
-4) Poisson shot noise if too few particles or unrelaxed systems,
-
-5) finite box size effects (with beat coupling?)
+{\bf (Any other systematics effects we want to discuss here?)}

### 2012-05-15 paper.tex
Added figures, and many changes to the main body of the text. Updated and completed the reference list. Included force tests.

@@
 \title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
-\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
+\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
-Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk} and Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}\\
+Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, 
+JD Emberson$^{1,5}$ \thanks{E-mail: emberson@astro.utoronto.ca} and Vincent Desjacques$^{6,7}$ \thanks{E-mail:vincent.desjacques@unige.ch}\\
 %\footnotemark[1]\thanks{This file has been amended to
@@
 $^{1}$Canadian Institute for Theoretical Astrophysics, University of
-Toronto, M5S 3H8, Canada\\
+Toronto, M5S 3H8, Ontario, Canada\\
 $^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
 $^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
-$^{4}$Department of XXX}
+$^{4}$Department of Mathematics and Computer Science, Laurentian University, P3E 2C6, Ontario, Canada\\
+$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
+$^{6}$Institute for Theoretical Physics, University of Z\"{u}rich, Z\"{u}rich, CH 8057, Switzerland\\
+$^{7}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}
 

### 2012-05-23 Extendedpp.tex
Incorporated Ilian annotations on the pdf

@@
   \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
-  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
+  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
+  These realizations are evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
+ and with $128^{3}$ particles, until $z=0$. The only difference is the ranges of the pp calculation.}
     \label{fig:power}

### 2012-05-23 Initial.tex
Incorporated Ilian annotations on the pdf

@@
 
-As mention in section \ref{sec:structure}, the code start off by reading a set of initial conditions.
-These correspond to a  $6 \times N$ phase-space array, where $N$ is the number of particles in the
-local node. Although most applications so far were based on initial conditions that follow Gaussian statistics,
-we have developed a non-Gaussian initial condition generator that we briefly describe in this section.
+As mention in section \ref{sec:structure}, the code starts off by reading a set of initial conditions.
+These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
+local volume. Although many applications are typically based on initial conditions that follow Gaussian statistics,
+we have developed a non-Gaussian initial conditions generator that we briefly describe in this section.
 
 The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
-is extended to include non-Gaussian features of the ``local'' form,
+is extended to include non-Gaussian features of the `local' form,
 $\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
@@
  as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
-Fig. \ref{fig:init} shows the power spectrum of two realizations with the same initial power spectrum, 
-one of which was started of with non-Gaussian initial conditions.
-We see that the power spectra agree at the sub-per cent level, as expected {\bf (?????)}
+Fig. \ref{fig:init} shows the late time power spectrum of two realizations that started off the same initial power spectrum, 
+but one of which had non-Gaussian initial conditions.
+We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
+agrees much better with one loop perturbation theory \citep{XXX}.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50.eps}
+    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
   \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,

### 2012-05-23 Introduction.tex
Incorporated Ilian annotations on the pdf

@@
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
- (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 (in preparation)), 
+ (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 ({\bf update with last paper!})), 
 formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
@@
 reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F},  
-and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
+and it is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
+
+%\subsection{Simulation suites}
+%\label{subsec:suites}
 
@@
 
-The paper is structured as follow: XXXX
+Tests on constraints of the redshift step size were performed 
+on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
+Each node contains 2 quad core Intel(R) Xeon(R) E5310 @ 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
+We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
+with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
+$128^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
+Tests on improvements of the force calculations were performed on the same machine, but with 8 times more particles and node, 
+and with a box size of $500 h^{-1}\mbox{Mpc}$.
+Hereafter we refer to these simulation modes as the CITA128 and CITA256 configurations respectively.
+
+We then test the dynamical range of the simulation, the non-Gaussian initial conditions generator and the run time halo finder algorithm 
+with a third simulation configuration, which we denote the  RANGER4000, that also evolves $4000^{3}$ particles 
+from $z=100$ to $z=0$ but with a box side of $3.2 h^{-1}\mbox{Gpc}$.
+ Much larger runs were produced at various high performance computing centers
+ around the world; their description and a comparison of their performance are also described in this paper.
+
+The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
+section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
+we then present in section \ref{sec:scaling} the scaling of the code to very large problems;
+sections \ref{sec:accuracy} and \ref{sec:systematics} discuss the accuracy and systematic effects of the code respectively.
+ We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions to 
+ to the default configuration, and conclude afterwards.

### 2012-05-23 PoissonSolver.tex
Incorporated Ilian annotations on the pdf

@@
 a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
-for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, but a sharp cutoff kills the force
-for pairs closer to a tenth of a grid cell.
+for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, 
+but we instead impose a sharp cutoff on the force
+for pairs that air closer than a tenth of a grid cell.
 
-The code units inherit from \citep{2004NewA....9..443T}, which we summarized here for completeness.
-In simulation units, the mean comoving mass density is set to unity, the length of a fine grid cell is set to one,
-and the Newton gravitational  constant $G$ is set to $1/(6 \pi a)$, where $a$ is the scale factor.
-With these three choices, one can transform any physical units into simulation units and vice versa.
+The code units inherit from \citep{2004NewA....9..443T}, which we summarize here for completeness.
+In simulation units,  the comoving length of a fine grid cell is set to one,
+such that the unit length in simulation unit is 
+\begin{eqnarray}
+1\mathcal{L} = a \frac{L}{N} 
+\end{eqnarray}
+where $a$ is the scale factor, $N$ is the total number of cells along one dimension,
+and $L$ is the comoving volume in $h^{-1}\mbox{Mpc}$.
+The mean comoving mass density is also set to unity in simulation units, 
+which, in physical units, corresponds to 
+\begin{eqnarray}
+1\mathcal{D} = \rho_{m}(0) a^{-3} = \Omega_{m} \rho_{c} a^{-3} = \frac{3 \Omega_{m} H_{o}^{2}}{8 \pi G a^{3} }
+\end{eqnarray}
+$\Omega_{m}$ is the matter density, $H_{o}$ is the Hubble's constant, $\rho_{c}$ is the critical density today,
+and $G$ is Newton's constant. The mass unit is found with $\mathcal{M} = \mathcal{DL}^{3}$.
+Specifying the value of $G$ on the grid fixes the time unit, and with the choice $G_{grid}$ = 1/$(6 \pi a)$,
+we get:
+\begin{eqnarray}
+1 \mathcal{T} = \frac{2a^{2}}{3}\frac{1}{\sqrt{\Omega_{m}H_{o}^{2}}}
+\end{eqnarray}
+These choices completely determine the convertion between physical units into simulation units and vice versa.
+For instance, the velocity units are given by $1\mathcal{V} = \mathcal{L}$/$\mathcal{T}$.
 

### 2012-05-23 ScalingPerformance.tex
Incorporated Ilian annotations on the pdf

--- 2012-05-01 ScalingPerformance.tex
@@
 \section{Scaling Performances}
+\label{sec:scaling}
 
-Real time versus size, bottle neck, description of the largest runs, etc.
-{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
+\begin{figure*}%[ht]
+%  \vskip -0.5cm 
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_curie.eps}
+    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_new.eps}
+%  \vskip -1.2cm 
+%  \vskip -0.5cm 
+  \caption{Scaling of {\small CUBEP3M} on Curie fat nodes (left) and 
+    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
+    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
+    in each case. Dashed line indicates the ideal weak 
+    scaling. The data is listed in Table \ref{summary_scaling_table}.
+    \label{scaling}
+% \vskip -0.9cm 
+}
+\end{center}
+\end{figure*}
+
+\begin{table*}%[ht]
+  \vskip -0.5cm 
+  \begin{center}
+\caption{Scaling of {\small CUBEP3M} on Curie. Speedup is 
+scaled to the smallest run.}
+\label{summary_scaling_table}
+\begin{tabular}{@{}|llllll|}
+\hline
+number of cores & speedup & ideal speedup & absolute timing (min) & 
+$N_{\rm particles}$& box size ($h^{-1}$Mpc)
+\\[2mm]\hline
+%\hline
+32  &  1.00 & - &3.2 & $256^3$ & 256\\
+256  & 7.21 & 8 &3.55 & $512^3$  & 512\\
+864  & 25.63 & 27 &4.8 & $864^3$  & 864\\
+2048  & 61.87 & 64 &26.48 & $2048^3$ & 2048 \\
+\hline
+\end{tabular}
+\caption{Scaling of  {\small CUBEP3M} for large number of cores 
+(on Ranger). Speedup is scaled to the smallest run.}
+\label{summary_scaling_table2}
+\begin{tabular}{@{}|llllll|}
+\hline
+number of cores & speedup & ideal speedup & absolute timing (min) & 
+$N_{\rm particles}$& box size ($h^{-1}$Mpc)
+\\[2mm]\hline
+%\hline
+864    & 1.00  & -    &258   & $1728^3$  & 6.3\\
+4096   & 4.53  & 4.74 &320   & $3072^3$  & 11.4\\
+10976  & 9.78  & 12.7 &845   & $5488^3$  & 20\\
+21952  & 14.73 & 25.4 &561   & $5488^3$  & 20 \\
+\hline
+\end{tabular}
+\end{center}
+  \vskip -0.7cm 
+\end{table*}
+
+%the Ranger system at the Texas Supercomputing
+%Center (in top 20 in the world) which is a SunBlade x6420 
+%with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
+% ``Barcelona'' processors and Infiniband networking. It has 
+%a total of 62976 computing cores and 125952 GB of total 
+%memory. Its nodes consist of 4 Quad Core processors and 32 GB 
+%of shared RAM. For efficiency reasons (local memory access) we 
+%typically use smaller MPI 'nodes' consisting of one Quad Core 
+%processor and 8 GB of RAM.
+
+The parallel algorithm of {\small CUBEP3M} is designed for `weak' 
+scaling, i.e. if the number of cores and the problem size 
+increase in proportion to each other, then for ideal scaling the 
+wall-clock time should remain the same, in contrast to `strong' 
+scaling, whereby the same problem solved on more cores should take 
+proportionately less wall-clock time. This weak scaling requirement 
+is dictated by the problems we are typically investigating (very 
+large and computationally-intensive) and our goals, which are to 
+address such large problems in the most efficient way, rather than 
+for the least wall-clock time. Furthermore, there is no explicit 
+load balancing, thus the code is most efficient if the sub-domains
+contain roughly equal number of particles, which is true for most
+cosmological-size volumes, but not for e.g. simulations of a single
+highly-resolved galaxy.
+
+%Since the current Curie is using Intel Nehalem architecture, while 
+%the thin Curie nodes will use the newer Westermere Intel architecture, 
+%we also show the scaling of our code on the Westermere-based Lonestar 
+%computer at the Texas Advanced Computing Centre
+
+The scaling was tested with a dedicated series 
+of simulations -- the CURIE simulation suite-- with increasing size and number of cores on the `fat' 
+(i.e. large-memory) nodes of the supercomputers Curie at the Tr\`{e}s Grand Centre de Calcul (TGCC) in France. 
+Our results are shown in Fig. \ref{scaling} and in 
+Table \ref{summary_scaling_table}. For appropriate direct comparison,
+all these simulations were performed using the same particle mass 
+($M_{\rm particle}=1.07\times10^{11}M_\odot$) and force resolution 
+(softening length 50 $h^{-1}$kpc). The box sizes used range from 256 $h^{-1}$Mpc
+to 2048 $h^{-1}$Mpc, and the number of particles from $256^3$ to $2048^3$.
+Simulations were run on 32 up to 2048 computing cores, also starting from 
+redshift $z=100$, and evolving until $z=0$. Our results show excellent scaling, within 
+$\sim3$ per cent of the ideal one, for up to 2048 cores.
+
+We have also ran {\small CUBEP3M} on a much larger number of cores, 
+from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
+particles on Ranger at the Texas Advanced Computing Centre (TACC), a SunBlade 
+x6420 system with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
+`Barcelona' processors and Infiniband networking,%, a total of 62,976 
+%computing cores and 125,952 GB of total memory. 
+%Its shared-memory nodes consist of 4 Quad Core processors and 32 GB of RAM.) 
+and on JUROPA at the J\"ulich Supercomputing Centre in Germany, an Intel Xeon X5570 
+(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.%, a total of 17,664
+%computing cores and 52 TB of total memory. Its shared-memory nodes 
+%consist of 2 Quad Core processors and 24 GB of RAM). 
+Since it is not practical to perform dedicated scaling tests on such large number 
+computing cores, we instead use the data extracted from production runs, 
+as listed in Table \ref{summary_scaling_table2}. We have found the 
+code to scale quite well, within 1.5 per cent of ideal up to 4,096 cores. 
+For very large number of cores (10,976) the scaling is slightly worse,
+due to increased communication 
+costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
+and load balancing issues, but still within $\sim20$ per cent of ideal. 
+These first three Ranger runs were performed 
+with 4 {\small MPI} processes and 4 threads per Ranger node ('4way') (using the 
+specially-provided {\it tacc\_affinity} NUMA script to bind the memory 
+usage to local sockets, thus ensuring memory affinity. This is important 
+because the 32 GB RAM/node on Ranger (and other computers with 
+multi-processor nodes) are actually not equal-access and the local 
+8 GB memory of each processor has much shorter access time).
+
+Furthermore, due to the very strong clustering of structures at those small 
+scales, some of the cuboid sub-domains came to contain well above the 
+average number of particles, thereby requiring more memory per {\small MPI} process
+in order to run. As a consequence, throughout most of the evolution the 
+largest two of these simulations were run with 4096 and 21,952 cores and 
+with only 2 {\small MPI} processes and 8 threads per Ranger node (2way), which on 
+Ranger allows using up to 16 GB of RAM per {\small MPI }process. In order to insure 
+local memory affinity a special NUMA control script (which we called {\it 
+tacc\_affinity\_2way}) 
+was developed for us by TACC staff and was used by us to run more efficiently 
+in this mode. However, this not-fully-local memory access does affect the 
+performance somewhat, as does the imperfect load balancing in such situations,
+as seen by the rightmost point on the scaling plot. The scaling in this case
+is 42 per cent below the ideal, although we note that we still get $\sim1.5$ speedup
+from doubling the core count, even given these issues. Overall the
+code scaling performance is thus quite satisfactory and it clearly 
+scales very well to extremely large number of cores and can therefore 
+do very large problems efficiently, utilizing the next generation 
+Petascale systems.
+
+Finally, we note that several special fixes had to be developed by TACC 
+and JUROPA staff in order for our largest runs to work properly and to be 
+able to use the software libraries required ({\small MPICH} and {\small FFTW}), as those 
+exhibited serious problems when applied to calculations of such 
+unprecedented size.
+
+{\bf Bottle neck in the largest runs?}
+
+{\bf (Hugh, do you have anything you would like to share 
 concerning the largest runs you did on Blue Gene? )}

### 2012-05-23 Structure.tex
Incorporated Ilian annotations on the pdf

@@
 and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.
-Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes, as the information exchange between cores that are part of the same motherboard is generally much faster.
+
+Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes in relatively small runs, as the information exchange between cores that are part of the same motherboard is generally much faster.
+As one scales to larger problems, however, the formation of dense objects can cause memory problems in such configurations, and increasing 
+the number of {\small MPI} processes helps to ensure memory locality.
+
 As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned.
 The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
-is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}
+is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}.
+Only when it comes to finding haloes at run time can this buffer create a problem, because a halo located close to the boundary
+can have a radius larger than the buffer zone.
 
@@
 \item{{\tt xv} stores the position and velocity of each particle} 
-\item{{\tt ll} stores the linked-list that accelerate the access to particles in a local domain}
+\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
 \item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 

### 2012-05-23 Systematics.tex
Incorporated Ilian annotations on the pdf

@@
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
-  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law.
-  Particles in the same cell follow the exact curve, then we observe a large scatter at 
+  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+  Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
   distance of the order of the fine grid, caused by the NGP interpolation scheme. 

### 2012-05-23 paper.tex
Incorporated Ilian annotations on the pdf

@@
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
-Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, 
+Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, \newauthor
 JD Emberson$^{1,5}$ \thanks{E-mail: emberson@astro.utoronto.ca} and Vincent Desjacques$^{6,7}$ \thanks{E-mail:vincent.desjacques@unige.ch}\\
@@
 \begin{abstract}
-This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST}, 
-a two-mesh gravity solver that was already among the fastest N-body codes. 
+This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
+solves gravity on a two-level mesh and that is among the fastest publicly available N-body codes. 
 Among the principal changes, the Poisson solver includes particle-particles interactions
-at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical.
+at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical,
+which allows for massive parallelization of the code.
 Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.

### 2012-05-23 Extendedpp.tex
Introduced a naming of the simulation runs for easier reference

--- 2012-05-15 Extendedpp.tex
@@
  To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
- plot, analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
+ plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
  We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
@@
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
-  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
-  with the pp force extended to six fine mesh layers.
-  Particles in the same cell follow the exact curve, then we observe a large scatter at 
-  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
+  \caption{Gravity force in the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
+  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
+  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
+  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_ppext0}. 
+  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances. 
     \label{fig:den_force_ppext6}}
@@
     \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
-  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions,
-  with the pp force extended to six fine mesh layers.
+  \caption{Fractional error on the force in the p$^3$m algorithm, in the radial (top) and tangential (bottom) directions,
+  in a CITA256 simulation in which the exact pp force extends to six fine mesh layers.
     \label{fig:den_force_fracErr_ppext6}}
@@
  We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
- and with $128^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
+ and with $128^{3}$ particles. We evolve these CITA128 realizations with different ranges for the pp calculation, and compare 
  the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
@@
   \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
-  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
+  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
+  These CITA128 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$ 
+  until $z=0$. The only difference between the runs is the ranges of the pp calculation.}
     \label{fig:power}

### 2012-05-23 Halos.tex
Introduced a naming of the simulation runs for easier reference

@@
     $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
-    simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
+    RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
     particles (red, solid). For reference we also show a widely-used 
@@
 A sample halo mass function produced based on our inlined SO halo finder 
-at redshift $z=0$ from a simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$
-particles is shown in Fig. \ref{mf}. We compare our result to the 
+at redshift $z=0$ from a RANGER4000 simulation is shown in Fig. \ref{mf}. We compare our result to the 
 precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most

### 2012-05-23 Initial.tex
Introduced a naming of the simulation runs for easier reference

@@
  as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
-Fig. \ref{fig:init} shows the late time power spectrum of two realizations that started off the same initial power spectrum, 
+Fig. \ref{fig:init} shows the late time power spectrum of two RANGER4000 realizations that started off the same initial power spectrum, 
 but one of which had non-Gaussian initial conditions.
@@
   from $4000^3$ particles.
-The two curves represent two realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator.
-The to curves differ at the sub-per cent level, as seen in the top panel.
+The two curves represent two RANGER4000 realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator.
+The two curves differ at the sub-per cent level, as seen in the top panel,
+  and the non-Gaussian realization is in much better agreement with the one-loop perturbation theory calculations.   
     \label{fig:init}}

### 2012-05-23 Introduction.tex
Introduced a naming of the simulation runs for easier reference

@@
 on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
-Each node contains 2 quad core Intel(R) Xeon(R) E5310 @ 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
+Each node contains 2 Quad Core Intel(R) Xeon(R) E5310 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
 We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
@@
 
-We then test the dynamical range of the simulation, the non-Gaussian initial conditions generator and the run time halo finder algorithm 
-with a third simulation configuration, which we denote the  RANGER4000, that also evolves $4000^{3}$ particles 
-from $z=100$ to $z=0$ but with a box side of $3.2 h^{-1}\mbox{Gpc}$.
+For tests of the accuracy over a large dynamical range of the simulation, of the non-Gaussian initial conditions generator and of the run time halo finder algorithm, 
+we used a third simulation configuration that were run at the Texas Advanced Computing Centre (TACC) on Ranger, a SunBlade 
+x6420 system with AMD x86\_64 Opteron Quad Core 2.3 GHz `Barcelona' processors and Infiniband networking.
+These RANGER4000 simulations evolved $4000^{3}$ particles 
+from $z=100$ to $z=0$ with a box side of $3.2 h^{-1}\mbox{Gpc}$
+on 4000 cores.
+
  Much larger runs were produced at various high performance computing centers
- around the world; their description and a comparison of their performance are also described in this paper.
+ around the world; their description and a comparison of their performance are also described in a later section of this paper.
 

### 2012-05-23 Systematics.tex
Introduced a naming of the simulation runs for easier reference

@@
 where the angle bracket corresponds to an ensemble (or volume) average.
-We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
-which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
-We observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
-The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
+We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
+observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
+The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed
 in the initial conditions.
@@
     \includegraphics[width=5.2in]{graphs/power_highres.eps}
-  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
-  from $4000^3$ particles. The simulation ran on 4000 cores on Ranger.
+  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation.
     \label{fig:power_highres}}
@@
 
-The actual force of gravity on the p$^3$m algorithm is presented in Fig. \ref{fig:den_force_ppext0}.
-This is a force versus distance plot, where the force is calculated in two steps. 
-We first compute the force on each particles of a given distribution as done at every time step.
-We then remove one particle, compute the force again on all particles, and record on a file the 
-force difference between the two calculations and the distance to the `hole'.
+The actual force of gravity in the p$^3$m algorithm,
+as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
+This shows the force versus distance -- in fine cell units -- a calculation that was performed in a CITA256 realization in two steps: 
+1- we compute the force on each particle in a given distribution, as done at every time step.
+2- we remove a selected particle, compute the force again on all particles, and record on a file the 
+force difference (before  and after the removal) as a function of the distance to the `hole'.
 
@@
   Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
-  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
+  distance of the order of the fine grid, caused by the NGP interpolation scheme
+  This calculation was obtained in a CITA256 realization. 
     \label{fig:den_force_ppext0}}
@@
 One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
-$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  Generally, a simulation should start at a redshift high enough that
+$\mbox{d}a/(a + \mbox{d}a)$ and is set to $0.05$ by default.  Generally, a simulation should start at a redshift high enough that
 the initial dimensionless power spectrum is well under unity at all scales. This ensures that the Zel'dovich approximation
- holds at the pre cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
+ holds at the per cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
  truncation error will be significant at the early time steps.
@@
 the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
-Fig.
-\ref{fig:ra_max} shows a comparison of late time power spectra of density fields that originate from the same initial conditions, and used the same random seeds to control the fine mesh shifts (mentioned above), but the value of $r_{max}$ was modified at each run.
+Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of a series of CITA256 realizations that originate from the same initial conditions, 
+and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
 We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
@@
     \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
-  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $500 h^{-1}\mbox{Mpc}$ per side,
-  from $256^3$ particles and a starting redshift of $z=200$. The different curves show different values of $r_{max}$. 
-  The resources required to run these simulations increase rapidly as $r_{max}$ decreases.    \label{fig:ra_max}}
+  \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
+ The starting redshift ofwas raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
+  The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
 \end{center}

### 2012-05-31 Halos.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

@@
 profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
-$\sim130$, and $1/r^2$ for lower overdensities. The TIS density profile has a
+$\sim130$, and $1/r^2$ for lower overdensities. The density profile has a
 similar outer slope (the relevant one here) to the Navarro, Frenk and White 
@@
 by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
-SO halo finder, which misses some of the low-mass halos. Using more sophisticated
+SO halo finder, which misses some of the low-mass halos. It was shown, however, that using more sophisticated
 halo finders (available through post-processing due to their heavier memory
-footprint) recovers the expected mass function.
-
-{\bf II: do we need to say anything more about having the option to use other
-halo finders like AHF and FOF?}
+footprint) allows to recover the expected mass function.
 
@@
 with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
-
+{\bf Expand here, describe, match with theory, etc...}
 \begin{figure}%[ht]

### 2012-05-31 Introduction.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

@@
 
-The basic problem that is addressed with N-body codes is a time evolution of an ensemble of $N$ particles
-that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operation, a cost that 
+The basic problem that is addressed with N-body codes is the time evolution of an ensemble of $N$ particles
+that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operations, a cost that 
 exceeds the memory and speed of current machines for large problems.
 Solving the problem  on a mesh \citep{1981csup.book.....H} reduces to $O(N\mbox{log}N)$ the number of operations,
-as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques, 
-typically using high performance libraries such as {\small FFTW} \citep{FFTW3}.
+as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques with high performance libraries such as {\small FFTW} \citep{FFTW3}.
 
-With the advent of large computing facilities, parallel coding has now become 
+With the advent of large computing facilities, parallel computations have now become 
 common practice, and N-body codes have evolved both in performance and complexity. 
-Many codes have opted for a tree algorithm, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field.
-These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
+Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field.
+These often have the advantage to balance the work load across the computing units, which enable fast calculations even in high density regions. 
 The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
-The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{1995astro.ph..3042F}.
- these are not designed to perform large scale PM calculations.
+The same reasoning applies to mesh-refined codes  \citep{1991ApJ...368L..23C}, 
+which in the end are not designed to perform fast PM calculations on large scales.
 
-{\bf ( PM code by japanese? Others codes I should mention?)}
+%{\bf ( PM code by japanese? Others codes I should mention?)}
 
-{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code that is designed such as to optimize the PM algorithm,
+{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code designed specifically to optimize the PM algorithm,
 both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T},
@@
 can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
-a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features {\bf (anything else since PMFAST?)}.
+a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features.
+ {\small PMFAST} was also equipped with a multi time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available. 
+ It also contains support for gas cosmological evolution through a TVD MHD module,
+ and a coupling interface with the radiative transfer code C2-Ray \citep{C2-Ray...}.
 
@@
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
- (\cite{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H}; Harnois-D\'{eraps} et al 2012 ({\bf update with last paper!})), 
+ \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, 2012arXiv1205.4989H}, 
 formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},

### 2012-05-31 PoissonSolver.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

@@
 to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
-a `cloud-in-cell' (cic) interpolation for the coarse grid,  and a `nearest-grid-point' interpolation 
-for the fine grid \citep{1981csup.book.....H}. Particle-particle interactions are not interpolated, 
-but we instead impose a sharp cutoff on the force
-for pairs that air closer than a tenth of a grid cell.
+a `cloud-in-cell' (CIC) interpolation for the coarse grid,  and a `nearest-grid-point' (NGP) interpolation 
+for the fine grid \citep{1981csup.book.....H}. This choice is motivated by the fact that the most straightforward 
+way to implement a p$^3$m algorithm on a mesh is to have exactly zero mesh force inside a grid, 
+which is only true for the NGP interpolation. Although CIC generally has a smoother and more accurate force,
+the pp implementation enhances the code resolution almost by an order of magnitude. 
+We impose a softening length of one tenth of a grid cell to prevent
+large scattering; particles seperated by less than this distance have their 
+particle-particle force set to zero.
 
@@
 
+In addition, {\small PMFAST} could run with a different kernel, described in MPT as the `least square kernel', 
+which basically adjust the kernel on cell by cell basis based on minimization of the deviation with respect
+to the Newtonian predictions. This was originally computed such as to optimize the force calculation for
+the case where both grids are computed from CIC interpolation. Moving to a mix CIC/NGP scheme
+requires solving the system of equations with the new configuration, a straightforward operation. 
+{\bf (results are presented somewhere?)}
+
 Finally, a choice must be done concerning the longest range of the force. Gravity can be either a) an accurate $1/r^2$ force, as far as the volume allows, 
@@
 which accurately models the growth of structure at very large scales. However, detailed studies of gravitational collapse would benefit 
-from the first settings. {\bf (Hugh, are the above two paragraphs correct?)}
+from the first settings.
 

### 2012-05-31 ScalingPerformance.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

@@
     in each case. Dashed line indicates the ideal weak 
-    scaling. The data is listed in Table \ref{summary_scaling_table}.
+    scaling. The data are listed in Table \ref{summary_scaling_table}.
     \label{scaling}
@@
 \end{tabular}
-\caption{Scaling of  {\small CUBEP3M} for large number of cores 
-(on Ranger). Speedup is scaled to the smallest run.}
+\caption{Scaling of  {\small CUBEP3M} on Ranger. Speedup is scaled to the smallest run.}
 \label{summary_scaling_table2}
@@
 %processor and 8 GB of RAM. 
+{\bf (This whole section needs to be edited for readability... Plus some things are already in the introduction.)}
 
@@
 highly-resolved galaxy.
+
+Obviously, for maximal efficiency, the number of tiles per node should be a multiple of the number of available processors per {\small MPI} process.
+Given the available freedom in the parallel configuration, as long as the load is balanced, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes.
+The information exchange between cores that are part of the same motherboard is generally much faster,
+so we minimize the overhead; in addition, having fewer {\small MPI} processes reduces the total amount of buffer zones, a memory that can be used to increase the physical resolution. 
+As one scales probes deeper into the non-linear regime however, 
+the formation of dense objects can cause memory problems in such configurations, and increasing 
+the number of {\small MPI} processes helps to ensure memory locality,
+especially in non-uniform memory access (numa) environments.
 
@@
 $\sim3$ per cent of the ideal one, for up to 2048 cores.
+
+The intermediary version between {\small PMFAST} and  the current p$^3$m code -- {\small CUBEPM} --
+was also ported to the IBM Blue Gene/L platform, and achieved 
+weak-scaling up to 4096 processes, with {\small CUBEPM} only incurring a 10 per cent overhead 
+at runtime (compared to 8 processes) for a balanced workload.  In order to 
+accomdate the limited amount of memory available per processing core on the 
+Blue Gene/L platform machines, it was necessary to perform the (long range) MPI FFT
+with a volumetric decomposition \citep{3DFFT}.
+Slab decomposition would have required a volume too large to fit in system 
+memory given the constraints in the simulation geometry.
 
@@
 unprecedented size.
-
-{\bf Bottle neck in the largest runs?}
-
-{\bf (Hugh, do you have anything you would like to share 
-concerning the largest runs you did on Blue Gene? )}

### 2012-05-31 Structure.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

--- 2012-05-15 Structure.tex
@@
 Many originate directly from MPT and were preserved in {\small CUBEP3M};
-those will be briefly mentioned, and we shall refer the reader to the original PMFAST paper for greater details.}.
+those will be briefly mentioned, and we shall refer the reader to the original {\small PMFAST} paper for greater details.}.
 We start with a walkthrough the code flow, and briefly discuss some specific sections that depart from standard N-body codes,
@@
 The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing.
-In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles},
-and for maximal efficiency, the number of tiles per node should be at least equal to the number of available processors.
-Given the available freedom in the parallel configuration, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes, as the information exchange between cores that are part of the same motherboard is generally much faster.
-As mentioned in MPT, the computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned.
-The thickness of this surface must be larger than the cutoff length, and we find that a 24 cells deep buffer
-is a good compromise between memory usage and accuracy {\bf (Hugh, is that correct?)}
+In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles}.
+The force matching of the two meshes is performed by introducing a cutoff in the two force kernel.
+The value of $r_{c}=16$ fine cells was chosen such as to balance the communication
+overhead between processes and the accuracy of the match between the two meshes.
+
+The computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned.
+The thickness of this surface must be larger than the cutoff length (of 16 cells), and we find that a 24 cells deep buffer
+is a good compromise between memory usage and accuracy.
+When it comes to finding haloes at run time can this buffer create a problem, because a large halo located close to the boundary
+can have a radius larger than the buffer zone, in which case it would be truncated, thus falling in the wrong mass bin, producing an incorrect center of mass, etc. 
+It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume.
 
@@
 \item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
+\item{{\tt PID} stores the unique particle identification tags.}
 \end{enumerate}
-
+The particle ID is a feature that can be switched off by removing a compilation flag, 
+and allows to optimize the code for higher resolution configurations.
 %\subsection{Code overview}
@@
 do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
-only the head processor is active\footnote{Other libraries such as {\small P3DFFT}({\tt http://code.google.com/p/p3dfft/}) currently permit
+only the single core {\small MPI} process is active\footnote{Other libraries such as {\small P3DFFT} ({\tt http://code.google.com/p/p3dfft/}) currently permit
  an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.
@@
 \caption{Overall structure of the two-level mesh algorithm.
-We have included the section that concerns the extended pp force calculation to show that it follows the same basic logic.
+We have included the section that concerns the extended pp force calculation to illustrate that it follows the same basic logic.
 We mention here that the three linked list arrays are distinct entities, and their structure differ slightly.
@@
 
-{\bf (Anything else to say in this section?)}
+Other considerations that need to be considered is that any decomposition geometry somehow 
+limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
+--  such as {\small CUBEP3M} -- requires the number of MPI processes to be a perfect cube.
+Also,  since the decomposition is volumetric, as opposed to density dependent -- 
+it suffers from load imbalance when probing deep into the non-linear regime.

### 2012-05-31 Systematics.tex
Incorporated Hughs comments, plus a catastrophic plot showing what happens if we turn off the random offset in the current code

@@
 
+Particles in the same fine cell follow the exact $1/r^{2}$ curve. The scatter at 
+  distances of the order of the fine grid is caused by the NGP interpolation scheme:
+  Particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
+  but feel the mesh force at grid cell distances,
+  creating up to an order of magnitude loss in the force.
+As the separation approaches a tenth of the box size or so, the force on the coarse mesh deviates
+from Newton's law in order to preserve periodic boundary conditions. 
+This feature, however, can be turned off at compilation time.
+
 Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
 It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value.
-{\bf (I would like to mention why we opted for NGP.
-Can't we do p3m with CIC on the fine grid?
-)}
+As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
+such that this undesired scatter is average out over a few time steps.
 
@@
   \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
-  Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
-  distance of the order of the fine grid, caused by the NGP interpolation scheme
-  This calculation was obtained in a CITA256 realization. 
+    This particular calculation was obtained in a CITA256 realization. 
     \label{fig:den_force_ppext0}}
@@
 
-The biggest problem with the pp force calculation is that it 
-is anisotropic and depends on the location of the fine mesh with respect 
-to the particles. An example are two particles on either side of a grid 
-cell boundary. These particles will experience the 1-grid cell separation
-NGP mesh force, but if the mesh were shifted such that they were
-within the same cell, they would experience the (larger) pp force instead. 
+The biggest problem with the straightforward pp force calculation is that the results 
+are anisotropic and depend on the location of the fine mesh with respect 
+to the particles. As an example, consider two particles on either side of a grid 
+cell boundary, experiencing their mutual gravity attraction via the (NGP) mesh force with a 1-grid cell separation.
+ If, instead, the mesh was shifted such that they were
+within the same cell, they would experience the (larger) pp force. 
 This effect is especially pronounced at the early stages of the simulation where
@@
 dimension and at each time step.  This adds negligible computational
-overhead as it can be applied during the particle position update
-and suppresses the mesh behaviour over multiple time steps.
+overhead as it is applied during the particle position update,
+and suppresses the mesh behaviour that otherwise grows over multiple time steps.
 It is possible to shift back the particles at the end of each time steps,
-which prevents a random drift of the whole population.
-This is convenient if one needs to correlate the initial and final positions of the particles.
+which prevents a random drift of the whole population, a necessary step 
+if one needs to correlate the initial and final positions of the particles for instance,
+or for hybrid dark matter -- MHD simulations.
 
 We ensure that, on average, this solution balances out the mesh feature,
-by tuning the NGP force kernel such as to provide  force as unbalanced as possible at grid cell distances.
-This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.
+by tuning the force kernels such as to provide  force as evenly balanced as possible at grid cell distances
+and at the cutoff length ($r_{c}=16$ fine cells).
+These adjustments are performed from the pairwise force test described in section \ref{sec:accuracy}.
 We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
@@
 a choice that  must be carefully balanced.
+
+We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
+The power spectrum is completely wrong, caused by the large scatter in the force on the fine mesh.
+In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
+did not have this problem since it used CIC interpolation on both meshes.  
+This test was performed with CITA128 simulations of very large box size,
+which should in principle agree with linear predictions up to the resolution.
+The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
+about 60 timesteps.
+We see that deviations are enormous when the random offset is turned off.
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
+  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA128 realizations that are 1000 Mpc/$h$ per side.
+The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard p$^3$m configuration.
+  The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
+\end{center}
+\end{figure}
 
@@
 As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
-In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the gain in time is more than a factor of four.
+In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
 

### 2012-06-01 Introduction.tex
Minor things, C2Ray reference, etc.

@@
  It also contains support for gas cosmological evolution through a TVD MHD module,
- and a coupling interface with the radiative transfer code C2-Ray \citep{C2-Ray...}.
+ and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
 

### 2012-06-01 paper.tex
Minor things, C2Ray reference, etc.

@@
 
-\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
+\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M}
 \author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
-\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
-Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}, Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}, \newauthor
-JD Emberson$^{1,5}$ \thanks{E-mail: emberson@astro.utoronto.ca} and Vincent Desjacques$^{6,7}$ \thanks{E-mail:vincent.desjacques@unige.ch}\\
+\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
+Ilian T. Iliev$^{3}$, Hugh Merz$^{4}$, \newauthor
+JD Emberson$^{1,5}$ and Vincent Desjacques$^{6,7}$\\
 %\footnotemark[1]\thanks{This file has been amended to
@@
 $^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
-$^{4}$Department of Mathematics and Computer Science, Laurentian University, P3E 2C6, Ontario, Canada\\
+$^{4}$SHARCNET, Laurentian University, P3E 2C6, Ontario, Canada\\
 $^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\

### 2012-06-06 Conclusion.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

--- 2012-05-01 Conclusion.tex
@@
 \section{Conclusion}
+
+This paper describes {\small CUBEP3M}, a public P$^3$M N-body code that is massively parallel, memory local, 
+and that scales well to 20,000 cores, thus raising the limits of the cosmological problem size one can handle.
+We summarize the code structure, the two-mesh Poisson solver algorithm, we present scaling and systematic tests
+that have been performed. We also describe a few utilities and extensions that come with the public release, 
+including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
 
@@
 \section*{Acknowledgements}
+
+The CITA256 simulations were run on the Sunnyvale cluster at CITA.
+ITI was supported by The Southeast Physics
+Network (SEPNet) and the Science and Technology Facilities Council
+grants ST/F002858/1 and ST/I000976/1. The authors acknowledge the
+TeraGrid and the Texas Advanced Computing Center (TACC) at The
+University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
+providing HPC and visualization resources that have contributed to the
+research results reported within this paper. ITI also acknowledges the Partnership for Advanced
+Computing in Europe (PRACE) grant 2010PA0442 which supported the code
+scaling studies.

### 2012-06-06 Extendedpp.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
  The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
- which are now at least 5 time smaller than in the default p$^{3}$m algorithm.
+ which are now at least 5 time smaller than in the default P$^{3}$M algorithm.
 
@@
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
-  \caption{Gravity force in the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
+  \caption{Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
   in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
@@
   distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_ppext0}. 
-  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances. 
+  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.. 
     \label{fig:den_force_ppext6}}
@@
     \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
-  \caption{Fractional error on the force in the p$^3$m algorithm, in the radial (top) and tangential (bottom) directions,
+  \caption{Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
   in a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
+  This was obtained over a single time step.
     \label{fig:den_force_fracErr_ppext6}}
@@
  We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
- and with $128^{3}$ particles. We evolve these CITA128 realizations with different ranges for the pp calculation, and compare 
+ and with $256^{3}$ particles. We evolve these CITA256 realizations with different ranges for the pp calculation, and compare 
  the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
@@
   \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
-  These CITA128 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$ 
+  These CITA256 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$ 
   until $z=0$. The only difference between the runs is the ranges of the pp calculation.}
@@
 
-Another way to quantify the improvement of the calculation is to look at the halo mass function for these different runs.
-{\bf (To do...)}
+Perhaps the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different CITA256 runs.
+Fig.
+\ref{fig:MassFunction_extpp} presents this comparison at redshift zero.
+About 11,000 haloes were found in the end, and we observe a good match with the Sheth-Tormen predictions.
+
+\begin{figure}
+  \begin{center}
+  \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
+  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. {\bf (y-axis?)}}
+    \label{fig:MassFunction_extpp}
+  \end{center}
+\end{figure}

### 2012-06-06 Halos.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
     RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
-    particles (red, solid). For reference we also show a widely-used 
+    particles (solid, red in the online version). For reference we also show a widely-used 
     precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 

### 2012-06-06 Introduction.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
 The first major improvement was to transform the volume decomposition in multi-node configurations 
-from slabs to cubes. One of the problem with slabs is that they do not scale well to large runs: 
+from slabs to cubes. The problem with slabs is that they do not scale well to large runs: 
 as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
-until it reaches the hard limit of a single cell layer.  With this enhancement, the code name was changed to {\small CUBEPM},
-which soon after incorporated particle-particle (pp) interactions at the sub-grid level. 
-The code was finally renamed {\small CUBEP3M}, and now includes a significant number of new features: the pp force
-can be extended to an arbitrary range, the size of the redshift jump can be constrained for improved accuracy during the first time steps,
-a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification which can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features.
- {\small PMFAST} was also equipped with a multi time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available. 
+until it reaches the hard limit of a single cell layer.
+With this enhancement, the code name was changed to {\small CUBEPM}.
+Soon after, it incorporated particle-particle (pp) interactions at the sub-grid level,
+and was finally renamed {\small CUBEP3M}. The public package now includes a significant number of new features: the pp force
+can be extended to an arbitrary range, the size of the redshift jumps can be constrained for improved accuracy during the first time steps,
+a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification that can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features.
+ {\small PMFAST} was equipped with a multi-time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available. 
  It also contains support for gas cosmological evolution through a TVD MHD module,
@@
 
-This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
-It has already been involved in a number of scientific applications over the last few years,
+{\small CUBEP3M} is one of the most competitive public N-body code and
+has been involved in a number of scientific applications over the last few years,
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
@@
 observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
-reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F},  
-and it is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
+reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}. 
+It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
 
@@
 
-Since {\small CUBEP3M} is not a new code, the accuracy and systematic tests were performed by different groups, on different machines,
-and with different configurations. It is not an ideal situation in which to quantify the performance, and each test must be viewed as a
+Since {\small CUBEP3M} is not new, the existing  accuracy and systematic tests were performed by different groups, on different machines,
+and with different geometric and parallelization configurations. It is not an ideal situation in which to quantify the performance, and each test must be viewed as a
 separate measurement that quantifies a specific aspect of the code. 
-We have tried to keep to a minimum the number of such different runs, and although the detailed numbers might vary across different runs, 
-the general trends are common to all of them.
+We have tried to keep to a minimum the number of such different runs, and although the detailed numbers vary 
+with the problem size and the machines, the general trends are are rather universal.
 
@@
 with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
-$128^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
-Tests on improvements of the force calculations were performed on the same machine, but with 8 times more particles and node, 
-and with a box size of $500 h^{-1}\mbox{Mpc}$.
-Hereafter we refer to these simulation modes as the CITA128 and CITA256 configurations respectively.
+$256^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
+Tests on improvements of the force calculations were performed on the same machine, but with  a box size of $500 h^{-1}\mbox{Mpc}$.
+Hereafter we refer to these simulation sets as the CITA256 configurations, and specify which volume was used in each context.
 
-For tests of the accuracy over a large dynamical range of the simulation, of the non-Gaussian initial conditions generator and of the run time halo finder algorithm, 
-we used a third simulation configuration that were run at the Texas Advanced Computing Centre (TACC) on Ranger, a SunBlade 
+For tests of the code accuracy, of the non-Gaussian initial conditions generator and of the run time halo finder algorithm, 
+we used a third simulation configuration series that was run at the Texas Advanced Computing Centre (TACC) on Ranger, a SunBlade 
 x6420 system with AMD x86\_64 Opteron Quad Core 2.3 GHz `Barcelona' processors and Infiniband networking.
@@
 
- Much larger runs were produced at various high performance computing centers
- around the world; their description and a comparison of their performance are also described in a later section of this paper.
-
 The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
 section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
-we then present in section \ref{sec:scaling} the scaling of the code to very large problems;
+we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
+including much larger runs that were produced at various high performance computing centers;
 sections \ref{sec:accuracy} and \ref{sec:systematics} discuss the accuracy and systematic effects of the code respectively.
- We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions to 
+ We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
  to the default configuration, and conclude afterwards.

### 2012-06-06 PoissonSolver.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
 for the fine grid \citep{1981csup.book.....H}. This choice is motivated by the fact that the most straightforward 
-way to implement a p$^3$m algorithm on a mesh is to have exactly zero mesh force inside a grid, 
+way to implement a P$^3$M algorithm on a mesh is to have exactly zero mesh force inside a grid, 
 which is only true for the NGP interpolation. Although CIC generally has a smoother and more accurate force,
-the pp implementation enhances the code resolution almost by an order of magnitude. 
+the pp implementation enhances the code resolution by almost an order of magnitude. 
 We impose a softening length of one tenth of a grid cell to prevent
@@
 
-The code units inherit from \citep{2004NewA....9..443T}, which we summarize here for completeness.
-In simulation units,  the comoving length of a fine grid cell is set to one,
+The code units inherit from \citep{2004NewA....9..443T} and are summarized here for completeness.
+The comoving length of a fine grid cell is set to one,
 such that the unit length in simulation unit is 
@@
 and $G$ is Newton's constant. The mass unit is found with $\mathcal{M} = \mathcal{DL}^{3}$.
-Specifying the value of $G$ on the grid fixes the time unit, and with the choice $G_{grid}$ = 1/$(6 \pi a)$,
+Specifying the value of $G$ on the grid fixes the time unit, and with $G_{grid}$ = 1/$(6 \pi a)$,
 we get:
@@
 \end{eqnarray}
-These choices completely determine the convertion between physical units into simulation units and vice versa.
+These choices completely determine the convertion between physical and simulation units.
 For instance, the velocity units are given by $1\mathcal{V} = \mathcal{L}$/$\mathcal{T}$.
@@
 \end{eqnarray}
-$G$ being Newton's constant. We solve this equation in Fourier space, where we write
+We solve this equation in Fourier space, where we write:
 \begin{eqnarray}
- \tilde{\phi}({\bf k}) = \frac{4 \pi G \tilde{\rho}({\bf k})}{- k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
+ \tilde{\phi}({\bf k}) = \frac{-4 \pi G \tilde{\rho}({\bf k})}{k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
 \label{eq:poissonFourier}
@@
 \end{eqnarray}
-Although this approach is fast, it involves a finite differentiation which enhances the numerical noise.
-We therefore opt for a force kernel, which is more accurate but has the inconvenient to require four extra Fourier transforms.
-In this case, we must solve the convolution in three dimensions:
+and
 \begin{eqnarray}
- F({\bf x}) = - m {\bf \nabla} \phi({\bf x})   = \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
+{\bf F}({\bf x}) = - m {\bf \nabla} \phi({\bf x}) 
+\label{eq:Force_sol}
+\end{eqnarray}
+Although this approach is fast, it involves a finite differentiation at the final step, which enhances the numerical noise.
+We therefore opt for a force kernel, which is more accurate, even though it requires four extra Fourier transforms.
+In this case, we must solve the convolution in three dimensions and define the force kernel {\boldmath $\omega$}$_{F}$ such as:
+\begin{eqnarray}
+ {\bf F}({\bf x}) =  \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
 \label{eq:poisson_solution_force}
 \end{eqnarray}
-The differentiation does not affect the density since it only acts on un-prime variables,
-and the force kernel is given by 
+Because the gradient acting on \ref{eq:poisson_solution_pot} affects only un-prime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
 \begin{eqnarray}
@@
 \end{eqnarray}
-The vector $\mbox{\boldmath $\beta$}(r)$ is related to the fourth order polynomial that is used in the potential case by
+The vector $\mbox{\boldmath $\beta$}(r)$ is related to the fourth order polynomial that is used in the potential case described in MPT by
  $ \mbox{\boldmath $\beta$} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
@@
 both with the density and the kernel (see [Eq. 7-8] of MPT).
-When matching the two  force kernels, it was realized that close to the cutoff region, the long range force is always on the low side, whereas 
-the short range force is scattered across the theoretical $1/r^2$ value. These behaviors are purely features of the CIC and NGP interpolation scheme 
-respectively. We identified a small range surrounding the cutoff length, in which we empirically adjust both kernels such as to improve 
-the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow \mbox{\boldmath $\omega$}_{s}({ r})*0.985$,
-and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow \mbox{\boldmath $\omega$}_{l}({ r})*1.2$.
-The two fudge factors were found by performing force measurements on two particles randomly placed in the volume.
+When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.
+By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match.
+Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
+and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
 
-In addition, {\small PMFAST} could run with a different kernel, described in MPT as the `least square kernel', 
-which basically adjust the kernel on cell by cell basis based on minimization of the deviation with respect
+As mentioned in section \ref{sec:structure} an summarized in FIg. \ref{fig:particle_mesh},
+the force kernels are first  in the code initialization stage.
+Eq. \ref{eq:poisson_solution_force} is then solved with fast Fourier transform along each directions, 
+ and is applied onto particles in the {\tt update\_velocity} subroutine.
+
+As an option, {\small PMFAST} could run with a different set of kernels, described in MPT as the `least square matching' , 
+which basically adjust the kernels on cell-by-cell basis based on minimization of the deviation with respect
 to the Newtonian predictions. This was originally computed such as to optimize the force calculation for
-the case where both grids are computed from CIC interpolation. Moving to a mix CIC/NGP scheme
+the case where both grids are obtained from CIC interpolation. Moving to a mix CIC/NGP scheme
 requires solving the system of equations with the new configuration, a straightforward operation. 
@@
 
-Finally, a choice must be done concerning the longest range of the force. Gravity can be either a) an accurate $1/r^2$ force, as far as the volume allows, 
+Finally, a choice must be done concerning the longest range of the coarse mesh force. Gravity can be either a) an accurate $1/r^2$ force, as far as the volume allows, 
 or b) modified to correctly match the periodicity of the boundary conditions. By default, the code is configured along to the second choice,
-which accurately models the growth of structure at very large scales. However, detailed studies of gravitational collapse would benefit 
-from the first settings.
-
-{\bf Least square kernel?}
+which accurately models the growth of structures at very large scales. However, detailed studies of gravitational collapse of a single large object would benefit 
+from the first settings, even though the code is not optimal for such settings.

### 2012-06-06 ScalingPerformance.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

--- 2012-05-23 ScalingPerformance.tex
@@
 increase in proportion to each other, then for ideal scaling the 
-wall-clock time should remain the same, in contrast to `strong' 
-scaling, whereby the same problem solved on more cores should take 
+wall-clock time should remain the same. This is to be in contrasted with `strong' 
+scaling codes, whereby the same problem solved on more cores should take 
 proportionately less wall-clock time. This weak scaling requirement 
@@
 address such large problems in the most efficient way, rather than 
-for the least wall-clock time. Furthermore, there is no explicit 
-load balancing, thus the code is most efficient if the sub-domains
-contain roughly equal number of particles, which is true for most
-cosmological-size volumes, but not for e.g. simulations of a single
-highly-resolved galaxy.
+for the least wall-clock time. Furthermore, we recall that there is no explicit 
+load balancing feature, thus the code is maximally efficient when the sub-domains
+contain roughly an  number of particles. This is true for most
+cosmological-size volumes that do not resolve too deep in the non-linear regime, 
+but not for e.g. simulations of a single highly-resolved galaxy.
+
+Because of the volumetric decomposition, the total number of {\small MPI} processes needs
+to be a perfect cube. Also, for maximal resource usage, the number of tiles per node 
+should be a multiple of the number of available {\small CPU}s per {\small MPI} process,
+such that no core sits idle in the threaded block.
+Given the available freedom in the parallel configuration, as long as the load is balanced, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes:
+the information exchange between cores that are part of the same motherboard is generally much faster.
+In addition, having fewer {\small MPI} processes reduces the total amount of buffer zones, 
+freeing memory that can be used to increase the mesh resolution. 
+As one probes deeper into the non-linear regime however, 
+the formation of dense objects can cause memory problems in such configurations, and increasing 
+the number of {\small MPI} processes helps to ensure memory locality,
+especially in non-uniform memory access (NUMA) environments.
 
@@
 
-The scaling was tested with a dedicated series 
-of simulations -- the CURIE simulation suite-- with increasing size and number of cores on the `fat' 
-(i.e. large-memory) nodes of the supercomputers Curie at the Tr\`{e}s Grand Centre de Calcul (TGCC) in France. 
-Our results are shown in Fig. \ref{scaling} and in 
-Table \ref{summary_scaling_table}. For appropriate direct comparison,
+ The intermediary version of the code -- {\small CUBEPM} --
+was first  ported to the IBM Blue Gene/L platform, and achieved 
+weak-scaling up to 4096 processes (over a billion particles), with the N-body calculation only incurring a 10 per cent overhead 
+at runtime (compared to 8 processes) for a balanced workload {\bf (How do we cite Hugh's work on blue gene?)}.  In order to 
+accommodate the limited amount of memory available per processing core on the 
+Blue Gene/L platform machines, it was necessary to perform the long range {\small MPI FFT}
+with a volumetric decomposition \citep{3DFFT}.
+Slab decomposition would have required a volume too large to fit in system 
+memory given the constraints in the simulation geometry.
+
+The scaling of {\small CUBEP3M}  was first  tested with a dedicated series 
+of simulations -- the CURIE simulation suite-- by increasing the size and number of cores on the `fat' 
+(i.e. large-memory) nodes of the  Curie supercomputer at the Tr\`{e}s Grand Centre de Calcul (TGCC) in France. 
+ For appropriate direct comparison,
 all these simulations were performed using the same particle mass 
@@
 Simulations were run on 32 up to 2048 computing cores, also starting from 
-redshift $z=100$, and evolving until $z=0$. Our results show excellent scaling, within 
-$\sim3$ per cent of the ideal one, for up to 2048 cores.
+redshift $z=100$, and evolving until $z=0$. Our results are shown in Fig. \ref{scaling} and in Table \ref{summary_scaling_table}, and present excellent scaling, within 
+$\sim3$ per cent of ideal, at least for up to 2048 cores.
 
@@
 from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
-particles on Ranger at the Texas Advanced Computing Centre (TACC), a SunBlade 
-x6420 system with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
-`Barcelona' processors and Infiniband networking,%, a total of 62,976 
-%computing cores and 125,952 GB of total memory. 
-%Its shared-memory nodes consist of 4 Quad Core processors and 32 GB of RAM.) 
-and on JUROPA at the J\"ulich Supercomputing Centre in Germany, an Intel Xeon X5570 
-(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.%, a total of 17,664
-%computing cores and 52 TB of total memory. Its shared-memory nodes 
-%consist of 2 Quad Core processors and 24 GB of RAM). 
-Since it is not practical to perform dedicated scaling tests on such large number 
-computing cores, we instead use the data extracted from production runs, 
-as listed in Table \ref{summary_scaling_table2}. We have found the 
-code to scale quite well, within 1.5 per cent of ideal up to 4,096 cores. 
-For very large number of cores (10,976) the scaling is slightly worse,
+particles on Ranger and on JUROPA at the J\"ulich Supercomputing Centre in Germany, 
+which is an Intel Xeon X5570 
+(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.
+Since it is not practical to perform dedicated scaling tests on such a large number of
+computing cores, we instead list in Table \ref{summary_scaling_table2} 
+the data directly extracted from production runs. We have found the 
+code to scale within 1.5 per cent of ideal up to 4096 cores. 
+For larger sizes ($\ge$10,976 cores), the scaling is less ideal,
 due to increased communication 
@@
 These first three Ranger runs were performed 
-with 4 {\small MPI} processes and 4 threads per Ranger node ('4way') (using the 
-specially-provided {\it tacc\_affinity} NUMA script to bind the memory 
-usage to local sockets, thus ensuring memory affinity. This is important 
-because the 32 GB RAM/node on Ranger (and other computers with 
-multi-processor nodes) are actually not equal-access and the local 
-8 GB memory of each processor has much shorter access time).
+with 4 {\small MPI} processes and 4 threads per Ranger node (`4way')\footnote{For these very large runs, 
+we used a NUMA script {\it tacc\_affinity}, specially-provided by the technical staff, 
+that bind the memory usage to local sockets, thus ensuring memory affinity. 
+This becomes important because the memory sockets per node 
+(32 GB RAM/node on Ranger) are actually not equal-access. Generally, the local 
+memory of each processor has much shorter access time.}.
 
-Furthermore, due to the very strong clustering of structures at those small 
-scales, some of the cuboid sub-domains came to contain well above the 
-average number of particles, thereby requiring more memory per {\small MPI} process
-in order to run. As a consequence, throughout most of the evolution the 
+Furthermore, due to the increasing clustering of structures at those small 
+scales, some of the cuboid sub-domains came to contain a number of particles
+well above the 
+average, thereby requiring more memory per {\small MPI} process
+in order to run until the end. 
+As a consequence, {\it throughout most of the evolution}{\bf (Ilian, did you actually restarted simulations and switched 4ways to 2ways?)} the 
 largest two of these simulations were run with 4096 and 21,952 cores and 
-with only 2 {\small MPI} processes and 8 threads per Ranger node (2way), which on 
-Ranger allows using up to 16 GB of RAM per {\small MPI }process. In order to insure 
-local memory affinity a special NUMA control script (which we called {\it 
-tacc\_affinity\_2way}) 
-was developed for us by TACC staff and was used by us to run more efficiently 
-in this mode. However, this not-fully-local memory access does affect the 
-performance somewhat, as does the imperfect load balancing in such situations,
-as seen by the rightmost point on the scaling plot. The scaling in this case
-is 42 per cent below the ideal, although we note that we still get $\sim1.5$ speedup
+with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
+Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
+local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}) 
+was developed for us by the TACC technical staff and allowed to run more efficiently 
+in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
+performance somewhat, as does the imperfect load balancing that arises in such situations.
+This can be seen in the rightmost point of FIg. \ref{scaling} (right panel), where the scaling 42 per cent below the ideal.
+We note that we still get $\sim1.5$ speedup
 from doubling the core count, even given these issues. Overall the
-code scaling performance is thus quite satisfactory and it clearly 
-scales very well to extremely large number of cores and can therefore 
-do very large problems efficiently, utilizing the next generation 
-Petascale systems.
+code scaling performance is thus satisfactory even at extremely large number of cores.
+ We expect the code to handle even larger problems efficiently, and is thus well suited to run on
+  next generation Petascale systems.
 
-Finally, we note that several special fixes had to be developed by TACC 
-and JUROPA staff in order for our largest runs to work properly and to be 
-able to use the software libraries required ({\small MPICH} and {\small FFTW}), as those 
-exhibited serious problems when applied to calculations of such 
+Finally, we note that several special fixes had to be developed by  the TACC 
+and JUROPA technical staff in order for our largest runs to work properly.
+In particular, we encountered unexpected problems from software libraries such as 
+{\small MPICH} and {\small FFTW} when applied to calculations of such 
 unprecedented size.
 
-{\bf Bottle neck in the largest runs?}
-
-{\bf (Hugh, do you have anything you would like to share 
-concerning the largest runs you did on Blue Gene? )}
+{\bf (We should also compare the memory usage to other codes like Gadget. Ilian mentioned Gadget uses 90 Bytes per particle. Is cubep3m really using 130 Bytes per particle? Check with mem\_usage.)}

### 2012-06-06 Structure.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
 and is global in the sense that the calculations require knowledge about the full simulated volume.
-The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing.
-In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles}.
-The force matching of the two meshes is performed by introducing a cutoff in the two force kernel.
-The value of $r_{c}=16$ fine cells was chosen such as to balance the communication
+The short range force and the particle-particle interactions are computed in parallel on
+a second level of cubical decomposition of the local volumes, the {\it tiles}.
+To make this possible, the fine grid arrays are constructed such as to support parallel memory access.
+In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.
+The force matching between the two meshes is performed by introducing a cutoff length, $r_{c}$, in the definition of the two force kernels.
+The value of $r_{c}=16$ fine cells was found to balance the communication
 overhead between processes and the accuracy of the match between the two meshes.
 
-The computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned.
-The thickness of this surface must be larger than the cutoff length (of 16 cells), and we find that a 24 cells deep buffer
+The computation of the short range force requires each tile to store the fine grid density of a region that includes a buffer surface around the physical volume it is assigned.
+The thickness of this surface must be larger than $r_{c}$, and we find that a 24 cells deep buffer
 is a good compromise between memory usage and accuracy.
-When it comes to finding haloes at run time can this buffer create a problem, because a large halo located close to the boundary
-can have a radius larger than the buffer zone, in which case it would be truncated, thus falling in the wrong mass bin, producing an incorrect center of mass, etc. 
-It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume.
+This is basically to fully compensate for the coarse mesh calculations, whose CIC interpolation scheme 
+reaches two coarse cells deep beyond the cutoff.
+
+When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, center of mass, etc.
+It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.
 
@@
 Because the coarse grid arrays require $4^3$ times less memory per node, 
-the bulk of the foot-print is concentrated in a handful of fine grid arrays.
-In addition, some of these are required for intermediate steps of the calculations only, 
-hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
+it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
+concentrated in a handful of fine grid arrays.
+Some of these are only required for intermediate steps of the calculations, 
+hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
 We present here the largest arrays used by the code:
@@
 The loop over tiles and the particle exchange are thus performed in three dimensions.
-{\bf (Should I mention leapfrog here or anywhere else?)}
 When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
-and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
+and proceeds to the long range. This section of the code is also parallelized on many occasions, but, unfortunately, the current {\small MPI-FFTW}
 do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
@@
  an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.
+ As in {\small PMFAST}, the particle position and velocity updates are performed in a leap frog scheme \citep{1981csup.book.....H}.
 
@@
 \end{verbatim}
-\caption{Overall structure of the code.}
+\caption{Overall structure of the code simplified for readability.}
 \label{fig:structure}
@@
 subroutine particle_mesh
-   call update_position
+   call update_position + apply random offset
    call link_list
@@
 \caption{Overall structure of the two-level mesh algorithm.
-We have included the section that concerns the extended pp force calculation to illustrate that it follows the same basic logic.
-We mention here that the three linked list arrays are distinct entities, and their structure differ slightly.
+We have included the section that concerns the standard pp and the extended pp force calculation, to illustrate that they follows similar linked-list logic.
 }
@@
 limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
---  such as {\small CUBEP3M} -- requires the number of MPI processes to be a perfect cube.
+--  such as {\small CUBEP3M} -- requires the number of {\small MPI} processes to be a perfect cube.
 Also,  since the decomposition is volumetric, as opposed to density dependent -- 

### 2012-06-06 Systematics.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
 observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
-The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed
-in the initial conditions.
+The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed in the initial conditions.
+{\b (I should add a vertical line at the k of the coarse mesh for visualization purposes.)}
 
@@
     \includegraphics[width=5.2in]{graphs/power_highres.eps}
-  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation.
+  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
+  compared to the non-linear predictions of {\small CAMB}.
     \label{fig:power_highres}}
@@
 
-The actual force of gravity in the p$^3$m algorithm,
+The actual force of gravity in the P$^3$M algorithm,
 as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
-This shows the force versus distance -- in fine cell units -- a calculation that was performed in a CITA256 realization in two steps: 
-1- we compute the force on each particle in a given distribution, as done at every time step.
-2- we remove a selected particle, compute the force again on all particles, and record on a file the 
+This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
+1- we compute the force on each particle in a given time step.
+2- we remove a selected particle, compute the force again on all particles, and record on file the 
 force difference (before  and after the removal) as a function of the distance to the `hole'.
 
-Particles in the same fine cell follow the exact $1/r^{2}$ curve. The scatter at 
+Particles in the same fine cell as the hole follow the exact $1/r^{2}$ curve. The scatter at 
   distances of the order of the fine grid is caused by the NGP interpolation scheme:
-  Particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
-  but feel the mesh force at grid cell distances,
-  creating up to an order of magnitude loss in the force.
-As the separation approaches a tenth of the box size or so, the force on the coarse mesh deviates
+  particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
+  but still feel the mesh force at grid cell distances,
+  creating up to an order of magnitude loss.
+As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
 from Newton's law in order to preserve periodic boundary conditions. 
@@
 As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
-such that this undesired scatter is average out over a few time steps.
+such that this undesired scatter is averaged out over a few time steps.
 
@@
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
-  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
-    This particular calculation was obtained in a CITA256 realization. 
+  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
+    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
     \label{fig:den_force_ppext0}}
@@
     \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
-  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions.
+  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
     \label{fig:den_force_fracErr}}
@@
 to the particles. As an example, consider two particles on either side of a grid 
-cell boundary, experiencing their mutual gravity attraction via the (NGP) mesh force with a 1-grid cell separation.
+cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized 1-grid cell separation.
  If, instead, the mesh was shifted such that they were
-within the same cell, they would experience the (larger) pp force. 
+within the same cell, they would experience the much larger pp force. 
 This effect is especially pronounced at the early stages of the simulation where
@@
 We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
-The power spectrum is completely wrong, caused by the large scatter in the force on the fine mesh.
+The power spectrum is completely wrong, due to the large scatter in the force from the fine mesh.
 In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
 did not have this problem since it used CIC interpolation on both meshes.  
-This test was performed with CITA128 simulations of very large box size,
+This test was performed with CITA256 simulations of very large box size,
 which should in principle agree with linear predictions up to the resolution.
 The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
-about 60 timesteps.
+about 60 time steps.
 We see that deviations are enormous when the random offset is turned off.
@@
     \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
-  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA128 realizations that are 1000 Mpc/$h$ per side.
-The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard p$^3$m configuration.
+  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA256 realizations that are 1000 Mpc/$h$ per side.
+The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration.
   The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
@@
 \end{figure}
+
+{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}
 
@@
 In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
+We should mention that with a proper use of the non-Gaussian initial conditions, it is possible to start the simulations at much later redshifts, where the time step sizes are dominated by a stronger mesh force.
 

### 2012-06-06 paper.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

@@
 This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
-solves gravity on a two-level mesh and that is among the fastest publicly available N-body codes. 
-Among the principal changes, the Poisson solver includes particle-particles interactions
-at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical,
-which allows for massive parallelization of the code.
-Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
-We discuss the structure of the code, its accuracy, and its scaling performance.
-In addition, many utilities have been added, including a runtime halofinder,
-particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
-and accuracy, when applicable .
-
+is now one of the fastest highest performance public N-body code. 
+Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
+interactions inside and across neighboring cells.
+Among important changes since {\small PMFAST},  {\small CUBEP3M} has a volumetric decomposition,
+allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
+even at this scale.
+In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
+In parallel, many utilities and extensions have been added to the standard code package, 
+including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
+We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
+of the code and its utilities, when applicable.
 \end{abstract}

### 2012-06-08 Initial.tex
Included force tests over 10 steps

@@
 We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
-agrees much better with one loop perturbation theory \citep{XXX}.
+agrees much better with one loop perturbation theory \citep{2004PhRvD..69j3513S,2008PhRvD..78l3534T}.
 

### 2012-06-08 Introduction.tex
Included force tests over 10 steps

@@
 including much larger runs that were produced at various high performance computing centers;
-sections \ref{sec:accuracy} and \ref{sec:systematics} discuss the accuracy and systematic effects of the code respectively.
+section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
  We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions

### 2012-06-08 Systematics.tex
Included force tests over 10 steps

@@
-\section{Accuracy}
+\section{Accuracy and Systematics}
 \label{sec:accuracy}
+
+ This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
+ We start with a demontration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
+ The halo mass function also assess the code capacity to model gravitational collapse, 
+ but depends on the particular halo finder used, which we discuss in section \ref{sec:halo}.
+ For the moment, we focus on the two-point function, and further quantify the accuracy of the force calculation
+ with comparisons to Newton's law of gravitation.
+
+ \subsection{Power Spectrum}
+ \label{subsec:powerspectrum}
 
@@
 
+\subsection{Mesh force at grid distances}
+\label{subsec:force}
+
 The actual force of gravity in the P$^3$M algorithm,
-as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
-This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
+as felt by a single particle, is presented in the left panel of Fig. \ref{fig:den_force_fracErr}.
+The left panel shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
 1- we compute the force on each particle in a given time step.
@@
 
-Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
+The right panel of Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions, calculated from a single time step.
 It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value.
-As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
-such that this undesired scatter is averaged out over a few time steps.
+This change of interpolation is essential for the pp part.
 
-\begin{figure}%[ht]
+\begin{figure*}%[ht]
   \begin{center}
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
-  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+  %\caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    %This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
+    %in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
+ %   \label{fig:den_force_ppext0}}
+%\end{center}
+%\end{figure}
+%
+%\begin{figure}%[ht]
+ % \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
+  \caption{({\it left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
     This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
-    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
-    \label{fig:den_force_ppext0}}
-\end{center}
-\end{figure}
-
-\begin{figure}%[ht]
-  \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
-  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
-  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
+    in a single time step. 
+    ({\it right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+  In a full simulation run, the scatter averages over many time steps, 
+    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
     \label{fig:den_force_fracErr}}
 \end{center}
-\end{figure}
-
-\section{Systematics}
-\label{sec:systematics}
-
-\subsection{Mesh force at grid distances}
+\end{figure*}
 
@@
 and at the cutoff length ($r_{c}=16$ fine cells).
-These adjustments are performed from the pairwise force test described in section \ref{sec:accuracy}.
+These adjustments are performed from a pairwise force test described in MPT.
 We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
@@
 
-{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}
+To illustrate how well the random offset performs, we present in Fig. \ref{fig:den_force_N10}
+the impact of averaging over 10 time steps. We recast the same force test in which we dig a hole, 
+with the addition that we apply a random offset the whole particle population at the end of the calculation.
+We repeat this operation ten times, always extracting the same particle, and average the mean force and the deviation.
+When compared with a single time step, we see that the scatter has reduced by almost a factor of three in the radial direction, 
+while the transverse deviation shows a milder improvement.
+
+\begin{figure*}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
+  \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
+    ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+    The left panels are calculated from a single time step, while the right panels show the average over time time steps.
+    \label{fig:den_force_N10}}
+\end{center}
+\end{figure*}
 

### 2012-06-08 paper.tex
Included force tests over 10 steps

@@
 interactions inside and across neighboring cells.
-Among important changes since {\small PMFAST},  {\small CUBEP3M} has a volumetric decomposition,
+Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
 allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling

### 2012-06-11 Halos.tex
Included force tests over 10 steps
included predictions for bias

@@
 with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
-{\bf Expand here, describe, match with theory, etc...}
+The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
+contains 200-2000 particles and so on. Comparisons with the linear predictions of \cite{Sheth_et_al} shows
+results consistent findings of \cite{Thinker}, namely that low-mass bias is over-predicted and high-mass bias is under-predicted.
+
 \begin{figure}%[ht]
@@
   \end{center}
-  \caption{Halo bias, measured at $z=0$ from a $4000^3$ particles simulations with a side length of $3.2 h^{-1}$Gpc.
+  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
   Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
   of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2,000 and so on.
+  The dashed lines are linear predictions from \protect \cite{Sheth_et_al}.
     \label{fig:halo}}

### 2012-06-12 Extendedpp.tex
Included force tests over 10 steps
added a rebined force plot to show how the mean behaves

@@
  The case where all three coordinates are equal is already calculated in the standard configuration of the code.
- To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
- plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
+ To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
+ plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to six layers of fine cells. 
  We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
- The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
+ The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
  which are now at least 5 time smaller than in the default P$^{3}$M algorithm.
 
- \begin{figure}%[ht]
+ \begin{figure*}%[ht]
   \begin{center}
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
-  \caption{Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
+%    \label{fig:den_force_ppext6}}
+%\end{center}
+%\end{figure}
+%\begin{figure}%[ht]
+%  \begin{center}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
+    \caption{({\it left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
   in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
   Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
-  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_ppext0}. 
-  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.. 
-    \label{fig:den_force_ppext6}}
-\end{center}
-\end{figure}
-
-\begin{figure}%[ht]
-  \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
-  \caption{Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
-  in a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
-  This was obtained over a single time step.
+  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
+  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
+  %\caption{
+  ({\it right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
+  calculated from a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
+  This was also obtained over a single time step.
     \label{fig:den_force_fracErr_ppext6}}
 \end{center}
-\end{figure}
+\end{figure*}
 
@@
   \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
-  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. {\bf (y-axis?)}}
+  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. 
+  The smallest haloes have a mass equivalent to 20 particles.{\bf (y-axis?)}}
     \label{fig:MassFunction_extpp}

### 2012-06-12 Halos.tex
added a rebined force plot to show how the mean behaves

@@
 The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
-contains 200-2000 particles and so on. Comparisons with the linear predictions of \cite{Sheth_et_al} shows
-results consistent findings of \cite{Thinker}, namely that low-mass bias is over-predicted and high-mass bias is under-predicted.
+contains 200-2000 particles and so on. Comparisons with the linear predictions of \citet{2001MNRAS.323....1S} shows
+results consistent the known flaws of the model, namely that low-mass bias is over-predicted and high-mass bias is under-predicted \citep{2010ApJ...724..878T}.
 
@@
   Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
-  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2,000 and so on.
-  The dashed lines are linear predictions from \protect \cite{Sheth_et_al}.
+  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2000 and so on.
+  The dashed lines are linear predictions from \protect \citet{2001MNRAS.323....1S}.
     \label{fig:halo}}

### 2012-06-12 Systematics.tex
added a rebined force plot to show how the mean behaves

@@
 observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
-The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed in the initial conditions.
-{\b (I should add a vertical line at the k of the coarse mesh for visualization purposes.)}
+The drop of power at high-$k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
+The fluctuations at low-$k$ come from the white noise imposed in the initial conditions. The vertical line corresponds to the scale of the coarse mesh.
 
@@
   compared to the non-linear predictions of {\small CAMB}. 
+  The vertical line corresponds to the scale of the coarse mesh.
     \label{fig:power_highres}}
@@
 When compared with a single time step, we see that the scatter has reduced by almost a factor of three in the radial direction, 
-while the transverse deviation shows a milder improvement.
+while the transverse deviation shows a milder improvement. When rebining these results, we observe that the mean mesh force
+calculation agrees well down to the grid cell, as seen in Fig. \ref{fig:F_rebin}. For this calculation, the hole was not part of a very dense halo,
+hence most of the force it experiences comes from the mesh.
 
@@
   \begin{center}
+     \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
     \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
   \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
@@
 \end{figure*}
+
+\begin{figure}%[ht]
+  \begin{center}
+     \includegraphics[width=3.2in]{graphs/densityForce_ppext=10_rebin.eps}
+  \caption{ Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$, averaged over 10 time steps.
+    Results are rebined to show how well the mean value follows the thoretical curve down to the grid size.
+    \label{fig:F_rebin}}
+\end{center}
+\end{figure}
 

### 2012-06-21 Extendedpp.tex
Updated force plots with rebins and N=10 averages

@@
  To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
- plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to six layers of fine cells. 
+ plot in a CITA128 realization, analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to  six layers of fine cells
+ in the force test part of the code {\bf (do pp\_ext2 as well?)}. 
  We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
  The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
- which are now at least 5 time smaller than in the default P$^{3}$M algorithm.
+ which are now at least 5 time smaller than in the default P$^{3}$M algorithm {\bf (Check these numbers again)}.
+ When averaging over 10 time steps, the improvement is only mild, as the calculation is already very accurate.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
-%    \label{fig:den_force_ppext6}}
-%\end{center}
-%\end{figure}
-%\begin{figure}%[ht]
-%  \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
-    \caption{({\it left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
-  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin.eps}
+     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin_N10.eps}
+    \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
+  in the same CITA128 realization as that shown in Fig. \ref{fig:den_force_fracErr}, 
+  except that the exact pp force has been extended to six fine mesh layers around each particle
+  in the force test code.
   Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
@@
   %\caption{
-  ({\it right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions,
-  calculated from a CITA256 simulation in which the exact pp force extends to six fine mesh layers. 
+  ({\it top right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
   This was also obtained over a single time step.
+  ({\it bottom row}:) Same as top row, but averaged over ten time steps.
+{ \bf (I should add pp\_ext = 2. ) }
     \label{fig:den_force_fracErr_ppext6}}

### 2012-06-21 Halos.tex
Updated force plots with rebins and N=10 averages

@@
 fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
-proceeds to search for and record all local density maxima above certain
-threshold (typically set to 100 above mean density) within the local 
-sub-domain's physical volume (excluding the sub-domain buffer zones). It then 
-uses parabolic interpolation on the density field to determine more precisely
+proceeds to search for all local density maxima above a certain
+threshold (typically set to a factor of 100 above mean density) within the local tile. 
+It then uses parabolic interpolation on the density field to determine more precisely
 the location of the maximum within the densest cell, and records the peak 
-position and value. The halo centre determined this way agrees closely with 
-the centre-of-mass of the halo particles discussed below.
+position and value. %The halo centre determined this way agrees closely with 
+%the centre-of-mass of the halo particles discussed below.
 
@@
 mean density within the halo drops below a pre-defined overdensity cutoff 
-(usually set to 178 in units of the mean, in accordance to the top-hat 
-collapse model). As we accumulate mass we remove it from the mesh, so that no 
+(usually set to 178 in units of the mean, in accordance with the top-hat 
+collapse model). As we accumulate the mass, we remove it from the mesh, so that no 
 mass element is double-counted. This method is thus inappropriate for finding 
-sub-halos as within this framework those are naturally incorporated in their 
-host halos. Because the halos are found on a grid of finite-size cells, it is 
-possible, especially for the low-mass halos, to overshoot the target overdensity.
-When this occurs we use an analytical halo density profile to correct the 
-halo mass and radius to the values corresponding to the target overdensity. 
-This analytical density profile is given by Truncated Isothermal Sphere (TIS) 
+sub-haloes since within this framework, those are naturally incorporated in their 
+host haloes. Because the haloes are found on a discrete grid, it is 
+possible, especially for those with lower mass, to overshoot the target overdensity.
+To minimize this efect, we correct the halo mass and radius with an analytical density profile. 
+We use the Truncated Isothermal Sphere (TIS) 
 profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
-$\sim130$, and $1/r^2$ for lower overdensities. The density profile has a
-similar outer slope (the relevant one here) to the Navarro, Frenk and White 
+$\sim130$, and a simple $1/r^2$ law for lower overdensities {\bf (lower or higher?)}. 
+These yield a similar outer slope to the Navarro, Frenk and White 
 \citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
@@
 Once the correct halo mass, radius and position are determined, we find all 
-particles which are within the halo radius. Their positions and velocities are
+particles that are within the halo radius. Their positions and velocities are
 used to calculate the halo centre-of-mass, bulk velocity, internal velocity 
@@
 is very close, but  typically slightly lower, than the halo mass calculated 
-based on the gridded density field. The particle centre-of-mass corresponds 
-very well to the halo centre found based on the gridded mass distribution.
+based on the gridded density field. The  centre-of-mass found this way closely follows 
+that found from the peak location, which is based on the gridded mass distribution.
 
@@
   \caption{Simulated halo multiplicity function, 
-    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
+    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$, based on a
     RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
@@
 
-A sample halo mass function produced based on our inlined SO halo finder 
-at redshift $z=0$ from a RANGER4000 simulation is shown in Fig. \ref{mf}. We compare our result to the 
+A sample halo mass function produced  from a RANGER4000 simulation at $z=0$ is shown in Fig. \ref{mf}. We compare our result to the 
 precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
 other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
-halo finders, the \citet{2008ApJ...688..709T} fit is based on the
+halo finders, this one is based on the
 SO search algorithm, whose masses are systematically different 
@@
 making this fit a better base for comparison here. Results show excellent
-agreement, within $\sim10$ per cent for all halos with masses corresponding to
-1000 particles or more. Lower-mass halos are somewhat under-counted compared
+agreement, within $\sim10$ per cent for all haloes with masses corresponding to
+1000 particles or more. Lower-mass haloes are somewhat under-counted compared
 to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
 by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
-SO halo finder, which misses some of the low-mass halos. It was shown, however, that using more sophisticated
-halo finders (available through post-processing due to their heavier memory
-footprint) allows to recover the expected mass function.
+SO halo finder, which misses some of the low-mass haloes. It was shown, however, that using more sophisticated
+halo finders (available only through post-processing calculations due to their heavier memory
+footprint) it is possible to recover the expected mass function.
 

### 2012-06-21 ScalingPerformance.tex
Updated force plots with rebins and N=10 averages

@@
 in order to run until the end. 
-As a consequence, {\it throughout most of the evolution}{\bf (Ilian, did you actually restarted simulations and switched 4ways to 2ways?)} the 
+As a consequence,  throughout most of their late evolution, the 
 largest two of these simulations were run with 4096 and 21,952 cores and 
@@
 Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
-local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}) 
+local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
 was developed for us by the TACC technical staff and allowed to run more efficiently 

### 2012-06-21 Systematics.tex
Updated force plots with rebins and N=10 averages

--- 2012-06-06 Systematics.tex
@@
-\section{Accuracy}
+\section{Accuracy and Systematics}
 \label{sec:accuracy}
+
+ This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
+ We start with a demonstration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
+ The halo mass function also assess the code capacity to model gravitational collapse, 
+ but depends on the particular halo finder used.
+ We thus post-pone the discussion on this aspect until in section \ref{sec:halo},
+and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
+ with comparisons to Newton's law of gravitation.
+
+ \subsection{Power Spectrum}
+ \label{subsec:powerspectrum}
 
@@
   compared to the non-linear predictions of {\small CAMB}. 
-    \label{fig:power_highres}}
+  The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
+    \label{fig:power_highres} {\bf Should I show another redshift instead? Also, the legend box cuts the text: need to fix this. Finally, I could also show linear theory.}}
 \end{center}
@@
 
+\subsection{Mesh force at grid distances}
+\label{subsec:force}
+
+The force test presented in MPT was carried by placing two particles at random locations on the grid, calculate the force between them, then iterate
+over different locations. This pairwise force test is useful to quantify the accuracy on a cell-by-cell basis, but lacks the statistics that occur in an actual time step calculation.  
 The actual force of gravity in the P$^3$M algorithm,
-as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
-This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
+as felt by a single particle during a single step, is presented in the top left panel of Fig. \ref{fig:den_force_fracErr}.
+This force versus distance plot was obtained from a CITA128 realization, and the calculation proceeds in two steps: 
 1- we compute the force on each particle in a given time step.
-2- we remove a selected particle, compute the force again on all particles, and record on file the 
-force difference (before  and after the removal) as a function of the distance to the `hole'.
+2- we remove a selected particle, thus creating a `hole', compute the force again on all particles, and record on file the 
+force difference (before  and after the removal) as a function of the distance to the hole.
 
@@
   particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
-  but still feel the mesh force at grid cell distances,
-  creating up to an order of magnitude loss.
+  but still feel the same mesh force at grid cell distances.
+This  creates a discrepancy up to an order of magnitude, in loss or gain, depending on the location of the pair with respect to the centre of the cell.
 As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
 from Newton's law in order to preserve periodic boundary conditions.
-This feature, however, can be turned off at compilation time.
 
-Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
-It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value.
-As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
-such that this undesired scatter is averaged out over a few time steps.
+It is not clear from this scatter plot whether the mean force felt  is below or under the predictions, hence we rebin the results
+in 50 logarithmically spaced bins and compute the mean and standard deviation. 
+We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
+there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the force force is otherwise
+in excellent agreement with Newton's law.
 
-\begin{figure}%[ht]
+The top right and middle right panels of Fig. \ref{fig:den_force_fracErr} show the fractional error on the force along the radial and tangential directions, also calculated from a single time step.
+The mean of the radial component is up to 80 per cent away from the exact law at sub-grid distances, but quickly drops under ten per cent for larger distances. 
+The transverse fractional error peaks exactly at the gris distance,  and is about two time smaller than the radial counterpart.
+
+\begin{figure*}%[ht]
   \begin{center}
     \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
-  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
-    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
-    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
-    \label{fig:den_force_ppext0}}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10.eps}
+  \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
+    in a single time step. 
+    ({\it top right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+  In a full simulation run, the scatter averages over many time steps, 
+    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
+    ({\it middle row}:) Top row organized in 50 logarithmic bins.  
+    ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
+    \label{fig:den_force_fracErr}
+    {\bf (Why is frac err radial slightly under 0? Problem in the units?)}}
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}%[ht]
-  \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
-  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
-  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
-    \label{fig:den_force_fracErr}}
-\end{center}
-\end{figure}
-
-\section{Systematics}
-\label{sec:systematics}
-
-\subsection{Mesh force at grid distances}
-
-The biggest problem with the straightforward pp force calculation is that the results 
+At grid scales, the fractional error is larger than {\small PMFAST}, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC.
+This prescription is responsible for the larger scatter about the theoretical value, but, as mentioned earlier,
+NGP interpolation is essential to our implementation of the pp part.
+At the same time, the biggest problem with the straightforward pp force calculation is that the results 
 are anisotropic and depend on the location of the fine mesh with respect 
 to the particles. As an example, consider two particles on either side of a grid 
-cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized 1-grid cell separation.
+cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized one-grid cell separation.
  If, instead, the mesh was shifted such that they were
 within the same cell, they would experience the much larger pp force. 
+This is clearly seen in the top left panel of Fig. \ref{fig:den_force_fracErr}, where particles physically close, but in different grid cells, 
+feel a force up to an order of magnitude too small.
 This effect is especially pronounced at the early stages of the simulation where
 the density is more homogeneous, and leads to mesh artefacts appearing
-in the density field. In order to minimize this systematic effect, 
+in the density field {\bf (Should I show a density plot?)}.
+
+In order to minimize these two systematic effect -- the large scatter and the anisotropy -- 
 we randomly shift the particle distribution relative to the mesh by a small
@@
 We ensure that, on average, this solution balances out the mesh feature,
-by tuning the force kernels such as to provide  force as evenly balanced as possible at grid cell distances
+by tuning the force kernels such as to provide a force as evenly balanced as possible, both at grid cell distances
 and at the cutoff length ($r_{c}=16$ fine cells).
-These adjustments are performed from the pairwise force test described in section \ref{sec:accuracy}.
-We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
-since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
-As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
-a choice that  must be carefully balanced.
+These adjustments are performed from the pairwise force test mentioned above (and described in MPT).
+The bottom panels of Fig. \ref{fig:den_force_fracErr} show the effect of averaging over ten time steps, 
+where the random offset is applied on the particles (and on the hole) at the end of each force calculation. 
+We observe that the agreement extends at the per cent level down to $r \sim 1.5$, beyond which point
+the discretize effect of the NGP causes an underestimate of the force of up to 60 per cent.
+{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}
 
-We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
-The power spectrum is completely wrong, due to the large scatter in the force from the fine mesh.
-In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
-did not have this problem since it used CIC interpolation on both meshes.  
-This test was performed with CITA256 simulations of very large box size,
-which should in principle agree with linear predictions up to the resolution.
-The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
-about 60 time steps.
-We see that deviations are enormous when the random offset is turned off.
+We note that this inevitable loss of force in the NGP scheme is one of the driving argument to extend the pp force outside the fine mesh cell,
+since the scattering about the actual $1/r^{2}$ law drops rapidly as the distance increases.
+As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a computational price,
+but at least we have the option to run the code with high precision.
+
+We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the otherwise default code configuration.
+This test was performed with a CITA256 simulation of very large box size,
+the output redshifts are very early (the upper most curve ($z=10$) was obtained after
+about 60 time steps), such that the power spectrum should agree with linear theory up to the resolution limit.
+Instead, we observe that the power spectrum grows completely wrong, due to the large scatter in the force from the fine mesh,
+and to the anisotropic nature of the P$^3$M calculations mentioned above.
+When these effects are not averaged over, the errors directly add up at each time step,
+which explains why later time are worst.
+We recall that {\small PMFAST} did not have this problem since it used CIC interpolation on both meshes.  
+{\bf (Anything else to add here?)}
 
@@
     \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
-  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA256 realizations that are 1000 Mpc/$h$ per side.
+  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is 1000 Mpc/$h$ per side.
 The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration.
-  The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
+  The dots were obtained by simply removing the random offset that is normally applied at each time step. \label{fig:disp_mesh}}
 \end{center}
@@
 
-{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}
+%\begin{figure*}%[ht]
+ % \begin{center}
+  %   \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
+   % \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
+  % \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
+  %  \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
+ % \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+ %   This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
+  %  ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+   % The left panels are calculated from a single time step, while the right panels show the average over time time steps.
+ %   \label{fig:den_force_N10}}
+%\end{center}
+%\end{figure*}
+
+%\begin{figure}%[ht]
+ % \begin{center}
+  %   \includegraphics[width=3.2in]{graphs/densityForce_ppext=10_rebin.eps}
+  %\caption{ Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+   % This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$, averaged over 10 time steps.
+   % Results are rebined to show how well the mean value follows the thoretical curve down to the grid size.
+   % \label{fig:F_rebin}}
+%\end{center}
+%\end{figure}
 

### 2012-06-21 paper.tex
Updated force plots with rebins and N=10 averages

@@
 In parallel, many utilities and extensions have been added to the standard code package, 
-including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
+including a runtime halo finder, a tuneable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
 We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
-of the code and its utilities, when applicable.
+of the code and its utilities, when applicable {\bf (We should include a density picture somewhere. Ideally a RANGER4000, for which we show the
+power spectrum and the halo mass function.)}.
 \end{abstract}

### 2012-06-29 Conclusion.tex
New Force plots, description of the pp algorithm

--- 2012-05-01 Conclusion.tex
@@
 \section{Conclusion}
+
+This paper describes {\small CUBEP3M}, a public and  massively parallel P$^3$M N-body code that inherits from{\small PMFAST} 
+ and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
+We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests
+that have been performed. We also describe various utilities and extensions that come with the public release, 
+including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
+{\small CUBEP3M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
+it has already been used  for a large number of scientific applications, and it is our hope that the current documentation will 
+help the community in interpreting its outcome.
 
@@
 \section*{Acknowledgements}
+
+The CITA simulations were run on the Sunnyvale cluster at CITA.
+ITI was supported by The Southeast Physics
+Network (SEPNet) and the Science and Technology Facilities Council
+grants ST/F002858/1 and ST/I000976/1. The authors acknowledge the
+TeraGrid and the Texas Advanced Computing Center (TACC) at The
+University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
+providing HPC and visualization resources that have contributed to the
+research results reported within this paper. ITI also acknowledges the Partnership for Advanced
+Computing in Europe (PRACE) grant 2010PA0442 which supported the code
+scaling studies. {\bf (Ue-Li, Hugh, Vincent, JD, any financial acknowledgements to add here?)}

### 2012-06-29 Extendedpp.tex
New Force plots, description of the pp algorithm

@@
  To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
- plot in a CITA128 realization, analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to  six layers of fine cells
- in the force test part of the code {\bf (do pp\_ext2 as well?)}. 
+ plot in a CITA128 realization, analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to  two layers of fine cells
+ in the force test part of the code. 
  We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
- The same applies to the fractional error on the radial and tangential components of the force, as seen in the right panel,
- which are now at least 5 time smaller than in the default P$^{3}$M algorithm {\bf (Check these numbers again)}.
- When averaging over 10 time steps, the improvement is only mild, as the calculation is already very accurate.
+ The fractional error on the radial and tangential components of the force, as seen in the right panel,
+ are now at least 5 time smaller than in the default P$^{3}$M algorithm.
+ When averaging over 10 time steps, we observe that the improvement is only mild, as the calculation is already very accurate {\bf (To verify/quantify)}.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin.eps}
-     \includegraphics[width=3.2in]{graphs/densityForce_ppext=6_rebin_N10.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6_rebin_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new.eps}
+     \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_new_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new_N10.eps}
     \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
   in the same CITA128 realization as that shown in Fig. \ref{fig:den_force_fracErr}, 
-  except that the exact pp force has been extended to six fine mesh layers around each particle
+  except that the exact pp force has been extended to two fine mesh layers around each particle
   in the force test code.
   Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
-  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
+  distance of the order of 2 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
   The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
@@
   ({\it bottom row}:) Same as top row, but averaged over ten time steps.
-{ \bf (I should add pp\_ext = 2. ) }
     \label{fig:den_force_fracErr_ppext6}}
@@
  To quantify the accuracy improvement versus computing time requirements, we performed the following test.
- We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
- and with $256^{3}$ particles. We evolve these CITA256 realizations with different ranges for the pp calculation, and compare 
+ We generate a set of initial conditions at a starting redshift of $z = 50$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$,
+ and with $512^{3}$ particles. We evolve these CITA512 realizations with different ranges for the pp calculation, and compare 
  the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
@@
  Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
- when extending  PM to P$^{3}$M first, and when adding successive layers of fine cell mesh where the pp force is extended.
-We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the fine grid calculations
-are more accurate as the distance increases. For this reason, it seems that a range of a few layers, between 3 and 6, suffices 
+ when extending  PM to P$^{3}$M first, and when adding successively one and two layers of fine cell mesh in which the pp force is extended.
+We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the mesh calculations
+become accurate enough as the distance increases. For this reason, it seems that a range of a two layers, suffices 
 to reduce most of the undesired NGP scatter.
@@
 This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
-to evolved the particles to the final redshift. Table \ref{table:cpu_pp_ext} presents this usage.
+to evolved the particles to the final redshift. For instance, in our CITA512 simulations, 
+the difference between $N_{layer} = 1$ and $2$ is about a factor of $2.3$ in real time.
+This number will change depending on the problem at hand and on the machine, and we recommand
+to perform performance gain vs resource usage tests on smaller runs before performing large scale simulations
+with the extended force.
 
@@
   \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
-  These CITA256 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$ 
-  until $z=0$. The only difference between the runs is the ranges of the pp calculation.}
+  These CITA512 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 50$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$ 
+  until $z=2.0$. The only difference between the runs is the ranges of the pp calculation. The inset shows details about the resolution turnaround. }
     \label{fig:power}
@@
 
-\begin{table}
-\begin{center}
-\caption{Scaling in {\small CPU} resources as a function of the range of the pp interaction.
-$N_{layers}$ refers to the number of fine mesh layers around a given cell, inside of which the force calculation
-is purely given by the pp contribution. }
-\begin{tabular}{|l|c|c|}
-\hline 
-             Type         & time (h)   \\
-                  \hline
-PM                         & 1.77 \\
-P$^{3}$M             & 2.09 \\
-\hline
-$N_{layers}$       &          \\
-\hline
- $1$ & 8.74 \\
- $2$ & 11.47\\
- $3$ & 14.30 \\
- $4$ & 18.87\\
- $5$ & 22.52\\
- $6$ & 29.62\\
- $7$ & 34.82\\
- $8$ & 47.00\\
+%\begin{table}
+%\begin{center}
+%\caption{Scaling in {\small CPU} resources as a function of the range of the pp interaction.
+%$N_{layers}$ refers to the number of fine mesh layers around a given cell, inside of which the force calculation
+%is purely given by the pp contribution. }
+%\begin{tabular}{|l|c|c|}
+%\hline 
+%             Type         & time (h)   \\
+%                  \hline
+%PM                         & 1.77 \\
+%P$^{3}$M             & 2.09 \\
+%\hline
+%$N_{layers}$       &          \\
+%\hline
+% $1$ & 8.74 \\
+% $2$ & 11.47\\
+% $3$ & 14.30 \\
+ %$4$ & 18.87\\
+ %$5$ & 22.52\\
+ %$6$ & 29.62\\
+ %$7$ & 34.82\\
+ %$8$ & 47.00\\
+%\hline \hline
+%\end{tabular}
+%\label{table:cpu_pp_ext}
+%\end{center}
+%\end{table}
 
-\hline \hline
-\end{tabular}
-\label{table:cpu_pp_ext}
-\end{center}
-\end{table}
-
-Perhaps the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different CITA256 runs.
+The power spectrum does not provide the complete story, and one of the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different CITA512 runs.
 Fig.
-\ref{fig:MassFunction_extpp} presents this comparison at redshift zero.
-About 11,000 haloes were found in the end, and we observe a good match with the Sheth-Tormen predictions.
+\ref{fig:MassFunction_extpp} presents this comparison at redshift two.
+About 56,000 haloes were found in the end, yielding a halo number density of about $0.0166$ per $\mbox{Mpc/$h$}^{3}$.
+We observe that the simulation undershoot the Sheth-Tormen predictions, which is caused by XXXXX. 
+We also observe that the pure PM code yields up 10 per cent less haloes than the P$^{3}$M version, 
+whereas the extended pp algorithm generally recovers  up to 10 per cent more haloes.
+The difference in performance between an extended pp  range of two vs four cells is rather mild, from where
+we conclude that $N_{layers} = 2$ is the most optimal choice.
 
@@
   \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
-  \caption{ Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. 
-  The smallest haloes have a mass equivalent to 20 particles.{\bf (y-axis?)}}
+  \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. 
+  The smallest haloes shown here have a mass equivalent to 100 particles, and fall in the lowest mass bin.
+  ({\it bottom}:) Ratio between the different curves and that of the default P$^3$M configuration.}
     \label{fig:MassFunction_extpp}

### 2012-06-29 Introduction.tex
New Force plots, description of the pp algorithm

@@
 
-Tests on constraints of the redshift step size were performed 
+Tests on constraints of the redshift step size and on improvements of the force calculations were performed 
 on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
 Each node contains 2 Quad Core Intel(R) Xeon(R) E5310 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
-We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
-with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
-$256^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
-Tests on improvements of the force calculations were performed on the same machine, but with  a box size of $500 h^{-1}\mbox{Mpc}$.
-Hereafter we refer to these simulation sets as the CITA256 configurations, and specify which volume was used in each context.
+Each of these tests were performed on the same cluster, but with  different box sizes, starting redshifts and particle numbers.
+Hereafter we refer to these simulation sets as the CITA configurations, and specify which parameters were used in each case.
 

### 2012-06-29 PoissonSolver.tex
added a rebined force plot to show how the mean behaves
New Force plots, description of the pp algorithm

@@
 the pp implementation enhances the code resolution by almost an order of magnitude.
-We impose a softening length of one tenth of a grid cell to prevent
-large scattering; particles seperated by less than this distance have their 
-particle-particle force set to zero.
 
@@
 
-Because these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
+Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
 both with the density and the kernel (see [Eq. 7-8] of MPT).
@@
 
-As mentioned in section \ref{sec:structure} an summarized in FIg. \ref{fig:particle_mesh},
+As mentioned in section \ref{sec:structure} and summarized in FIg. \ref{fig:particle_mesh},
 the force kernels are first  in the code initialization stage.
@@
 
-As an option, {\small PMFAST} could run with a different set of kernels, described in MPT as the `least square matching' , 
+The pp force is calculated during the fine mesh velocity update, which avoids loading the particle list twice and allows the operation to be threaded without significant additional work.
+During this process, the particles within a given fine mesh tile are first read in via the linked list,
+then their velocity is updated with the fine mesh force component, according to their location within the tile. 
+In order to organize the particle-particle interactions, we proceed in constructing a set of threaded fine-cell linked list chains for each coarse cell. 
+We then calculate the pairwise force between all particles that lie within the same fine mesh cell, exluding pairs whose separation is smaller than a softening length of one tenth of a grid cell; particles seperated by less than this distance have their
+particle-particle force set to zero.
+As this proceeds, we accumulate the  pp force applied on each particle and then determine the maximum force element of the pp contribution, which is also taken into account when constraining the length of the global timestep.
+
+This kind of force softening is generally required by any code to prevent
+large scattering as $r \rightarrow 0$, which can otherwise slow the calculation down, 
+and to reduce the two-body relaxation, which can affect the numerical convergence. 
+Many force softening scheme can be found in the litterature, including Plummer force, uniform or linear density profiles or the spline-softened model 
+(see \cite{Dyer1993} for a review). In this case, a sharp force cutoff corresponds to a particle density profile that resembles that of a hollow shell.
+In comparison with other techniques, this  force softening is the easiest to code, the fastest to execute, but has a discontinuous profile at the cutoff lenght,
+which certainly has an impact on the smallest resolved structure. 
+{\bf (Have we tried anything else? Why did we choose this one? How does this cutoff translates in terms of energy and momentum conservation? Unstable?)}
+
+In addition, {\small PMFAST} could run with a different set of force kernels, described in MPT as the `least square matching' , 
 which basically adjust the kernels on cell-by-cell basis based on minimization of the deviation with respect
@@
 requires solving the system of equations with the new configuration, a straightforward operation.
-{\bf (results are presented somewhere?)}
+ With the inclusion of the random shifting, it is not clear how much improvement 
+one would recover from this other kernel matching.  It is certainly something
+we will investigate and document in the near future.
 

### 2012-06-29 ScalingPerformance.tex
New Force plots, description of the pp algorithm

@@
 weak-scaling up to 4096 processes (over a billion particles), with the N-body calculation only incurring a 10 per cent overhead 
-at runtime (compared to 8 processes) for a balanced workload {\bf (How do we cite Hugh's work on blue gene?)}.  In order to 
+at runtime (compared to 8 processes) for a balanced workload \footnote{\tt http://web.archive.org/web/20060925132146/http://www-03.ibm.com/servers/deepcomputing/pdf/Blue\_Gene\_Applications\_Paper\_CubePM\_032306.pdf}.
+In order to
 accommodate the limited amount of memory available per processing core on the 

### 2012-06-29 Systematics.tex
New Force plots, description of the pp algorithm

@@
 This  creates a discrepancy up to an order of magnitude, in loss or gain, depending on the location of the pair with respect to the centre of the cell.
-As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
-from Newton's law in order to preserve periodic boundary conditions.
+As the separation approaches a tenth of the full box size or so, the force on the coarse mesh 
+scatters significantly about Newton's law due to periodic boundary conditions. 
+As mentioned at the end of section \ref{sec:Poisson}, the longest range of force kernel can either model accurately Newtonian gravity,
+or the structure growth of the largest modes, but not both. For the sake of demonstrating the accuracy of the force calculation,
+we chose the first option in this section.
 
-It is not clear from this scatter plot whether the mean force felt  is below or under the predictions, hence we rebin the results
+It is not clear from these scatter plots whether the mean force felt  is below or under the predictions, hence we rebin the results
 in 50 logarithmically spaced bins and compute the mean and standard deviation. 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new.eps}
   \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.

### 2012-06-29 Extendedpp.tex
new figures for pp_extended force

@@
     \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new.eps}
-     \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_new_N10.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new_N10.eps}
+     \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_N10_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new.eps}
     \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,

### 2012-07-09 Conclusion.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 
-This paper describes {\small CUBEP3M}, a public and  massively parallel P$^3$M N-body code that inherits from{\small PMFAST} 
+This paper describes {\small CUBEP3M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
  and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
@@
 help the community in interpreting its outcome.
-
 The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
-compiling and running strategy is can be found at {\tt www.wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.
+compiling and running strategy is can be found on the CITA wiki page\footnote{\tt www.wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.
 
@@
 Network (SEPNet) and the Science and Technology Facilities Council
-grants ST/F002858/1 and ST/I000976/1. The authors acknowledge the
+grants ST/F002858/1 and ST/I000976/1. 
+Computations for the SciNet1024 runs were performed on the GPC supercomputer at the SciNet HPC Consortium.
+SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto.
+The authors acknowledge the
 TeraGrid and the Texas Advanced Computing Center (TACC) at The
@@
 Computing in Europe (PRACE) grant 2010PA0442 which supported the code
-scaling studies. {\bf (Ue-Li, Hugh, Vincent, JD, any financial acknowledgements to add here?)}
+scaling studies. UEP and JDE are supported by the NSERC of Canada.
+%{\bf ( Hugh, Vincent, JD, any financial acknowledgements to add here?)}

### 2012-07-09 Extendedpp.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
  in the force test part of the code. 
- We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
+ We observe that the scatter about the theoretical curve has reduced significantly, down to the few percent level, 
+ and is still well balanced around the theoretical predictions.
  The fractional error on the radial and tangential components of the force, as seen in the right panel,
  are now at least 5 time smaller than in the default P$^{3}$M algorithm.
- When averaging over 10 time steps, we observe that the improvement is only mild, as the calculation is already very accurate {\bf (To verify/quantify)}.
+ When averaging over 10 time steps, we observe that improvement is relatively mild, as the calculation is already very accurate.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new.eps}
-     \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_N10_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_new-1.eps}
+     \includegraphics[width=3.2in]{graphs/densityForce_ppext=2_rebin_N10_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new-1.eps}
     \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
@@
 This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
-to evolved the particles to the final redshift. For instance, in our CITA512 simulations, 
-the difference between $N_{layer} = 1$ and $2$ is about a factor of $2.3$ in real time.
+to evolved the particles to the final redshift. For instance, in our SciNet512 simulations, 
+the difference between the default configuration and $N_{layer} = 1$ and $2$ is about a factor of $2.78$ in real time,
+and about $6.5$ for $N_{layer} = 2$.
 This number will change depending on the problem at hand and on the machine, and we recommand
@@
   \begin{center}
-  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
-  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
-  These CITA512 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 50$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$ 
-  until $z=2.0$. The only difference between the runs is the ranges of the pp calculation. The inset shows details about the resolution turnaround. }
+  \epsfig{file=graphs/pp_ext_new_z1.eps,width=0.49\textwidth}
+  \caption{ Dimensionless power spectrum for varying range of the exact pp force.
+  These SciNet512 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$ 
+  until $z=1.0$. The only difference between the runs is the ranges of the pp calculation. The inset shows details about the resolution turnaround, and the
+  thick vertical line corresponds to the coarse mesh scale. }
     \label{fig:power}
@@
 
-The power spectrum does not provide the complete story, and one of the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different CITA512 runs.
+The power spectrum does not provide the complete story, and one of the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different SciNet512 runs.
 Fig.
-\ref{fig:MassFunction_extpp} presents this comparison at redshift two.
-About 56,000 haloes were found in the end, yielding a halo number density of about $0.0166$ per $\mbox{Mpc/$h$}^{3}$.
-We observe that the simulation undershoot the Sheth-Tormen predictions, which is caused by XXXXX. 
-We also observe that the pure PM code yields up 10 per cent less haloes than the P$^{3}$M version, 
-whereas the extended pp algorithm generally recovers  up to 10 per cent more haloes.
+\ref{fig:MassFunction_extpp} presents this comparison at redshift $z = 1.0$.
+About 76,000 haloes were found in the end, yielding a halo number density of about $0.0225$ per $\mbox{Mpc/$h$}^{3}$.
+We observe that the simulation undershoot the Sheth-Tormen predictions, which is caused by the relatively low resolution of the configuration
+comparered to the RANGER4000 run. 
+We also observe that the pure PM code yields up 10 per cent less haloes than the P$^{3}$M version, over most of the mass range,
+whereas the extended pp algorithm generally recovers  up to 10 per cent more haloes in the range $10^{11} - 10^{13}$.
 The difference in performance between an extended pp  range of two vs four cells is rather mild, from where
@@
   \begin{center}
-  \epsfig{file=graphs/MassFunction_pp_ranges.eps,width=0.49\textwidth}
-  \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen. 
+  \epsfig{file=graphs/MassFunction_pp_ranges_new_z1.eps,width=0.49\textwidth}
+  \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen, at $z = 1.0$. 
   The smallest haloes shown here have a mass equivalent to 100 particles, and fall in the lowest mass bin.

### 2012-07-09 Halos.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 
-A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
-with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
-The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
-contains 200-2000 particles and so on. Comparisons with the linear predictions of \citet{2001MNRAS.323....1S} shows
-results consistent the known flaws of the model, namely that low-mass bias is over-predicted and high-mass bias is under-predicted \citep{2010ApJ...724..878T}.
-
-\begin{figure}%[ht]
+%A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
+%with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
+%The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
+%contains 200-2000 particles and so on. Comparisons with the linear predictions of \citet{2001MNRAS.323....1S} shows
+%results consistent the known flaws of the model, namely that low-mass bias is over-predicted and high-mass bias is under-predicted \citep{2010ApJ...724..878T}.
+%
+%\begin{figure}%[ht]
 %  \vskip -0.5cm 
-  \begin{center}
-    \includegraphics[width=3.2in]{graphs/bias.eps}
-  \end{center}
-  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
-  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
-  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2000 and so on.
-  The dashed lines are linear predictions from \protect \citet{2001MNRAS.323....1S}.
-    \label{fig:halo}}
-\end{figure}
+%  \begin{center}
+%    \includegraphics[width=3.2in]{graphs/bias.eps}
+%  \end{center}
+%  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
+%  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
+%  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2000 and so on.
+%  The dashed lines are linear predictions from \protect \citet{2001MNRAS.323....1S}.
+%    \label{fig:halo}}
+%\end{figure}
 

### 2012-07-09 Initial.tex
Updated force plots with rebins and N=10 averages
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
-\subsection{Initial Conditions}
+\subsection{Initial conditions}
 \label{subsec:init}
@@
 we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
+This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.
+For the Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshift.
+Consequently, the size of the first few redshift jumps in such simulations are rather large, and therefore less accurate. 
+One way to fix this is to constrain the size of the jumps, as discussed in section \ref{subsec:z_jumps}.
+However,  the simulations take much more time to run, and this solution is therfore not adequate for 
+very large configurations.
 

### 2012-07-09 Introduction.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
- \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, 2012arXiv1205.4989H}, 
+ \citep{2010arXiv1008.3506Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
 formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
@@
 Hereafter we refer to these simulation sets as the CITA configurations, and specify which parameters were used in each case.
+Some complimentary runs were also performed on the SciNet GPC cluster \citep{Scinet}.
 

### 2012-07-09 PoissonSolver.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 
-This kind of force softening is generally required by any code to prevent
-large scattering as $r \rightarrow 0$, which can otherwise slow the calculation down, 
+Force softening is generally required by any code to prevent
+large scattering as $r \rightarrow 0$ that can otherwise slow the calculation down, 
 and to reduce the two-body relaxation, which can affect the numerical convergence. 
 Many force softening scheme can be found in the litterature, including Plummer force, uniform or linear density profiles or the spline-softened model 
-(see \cite{Dyer1993} for a review). In this case, a sharp force cutoff corresponds to a particle density profile that resembles that of a hollow shell.
-In comparison with other techniques, this  force softening is the easiest to code, the fastest to execute, but has a discontinuous profile at the cutoff lenght,
-which certainly has an impact on the smallest resolved structure. 
-{\bf (Have we tried anything else? Why did we choose this one? How does this cutoff translates in terms of energy and momentum conservation? Unstable?)}
+(see \citet{1993ApJ...409...60D} for a review). In the current case, a sharp force cutoff corresponds to a particle interacting with a hollow shell.
+In comparison with other techniques, this  force softening is the easiest to code and the fastest to execute. 
+Generally, it is desirable to match the smoothness of the force to the order of the time integration. 
+A Plummer force is infinitely differentiable, which is a sufficient but not necessary condition for our 2nd order time integration.  
+Also, one of the drawback of Plummer's softening is that the resolution degrades smoothly: the effects of the smoothing are present at all radii. 
+In comparison, the uniform density and hollow shell alternatives both have the advantage that the deviations from $1/r^2$ are minimized. 
+Although all the results presented in this paper were obtained with the sharp cutoff softening, other schemes can easily be adopted as these 
+are typically a single line changes to the code. \citep{1993ApJ...409...60D} argues in favor on the uniform density profile scheme -- which is more physical --
+and future developments of {\small CUBEP3M} will incorporate this option.
 

### 2012-07-09 ScalingPerformance.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

--- 2012-06-21 ScalingPerformance.tex
@@
 weak-scaling up to 4096 processes (over a billion particles), with the N-body calculation only incurring a 10 per cent overhead 
-at runtime (compared to 8 processes) for a balanced workload {\bf (How do we cite Hugh's work on blue gene?)}.  In order to 
+at runtime (compared to 8 processes) for a balanced workload\footnote{\tt http://web.archive.org/web/20060925132146/ http://www-03.ibm.com/servers/deepcomputing/pdf/ Blue\_Gene\_Applications\_Paper\_CubePM\_032306.pdf}.
+In order to
 accommodate the limited amount of memory available per processing core on the 
@@
 unprecedented size.
-
-{\bf (We should also compare the memory usage to other codes like Gadget. Ilian mentioned Gadget uses 90 Bytes per particle. Is cubep3m really using 130 Bytes per particle? Check with mem\_usage.)}

### 2012-07-09 Structure.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 and allows to optimize the code for higher resolution configurations.
+In terms of memory, {\small CUBEP3M} uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like {\small GADGET}, which 
+lies around $90$ per particles.
 %\subsection{Code overview}
@@
 
-Other considerations that need to be considered is that any decomposition geometry somehow 
+Other constraints that need to be considered is that any decomposition geometry somehow 
 limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
@@
 it suffers from load imbalance when probing deep into the non-linear regime.
+This latter effect can be overcome by allocating extra memory per node, either for the full run,
+or for the first part of it, follow by a restart configuration with more memory.

### 2012-07-09 Systematics.tex
new figures for pp_extended force
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

@@
 
- \subsection{Power Spectrum}
+ \subsection{Density and Power Spectrum}
  \label{subsec:powerspectrum}
@@
 where the angle bracket corresponds to an ensemble (or volume) average.
-We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
-observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
-The drop of power at high-$k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
-The fluctuations at low-$k$ come from the white noise imposed in the initial conditions. The vertical line corresponds to the scale of the coarse mesh.
+Fig. \ref{fig:density} presents  a 2-dimensional density projection of a RANGER4000 simulation, which evolved 4000$^3$ particles until redshift zero.
+We then plot in Fig. \ref{fig:power_highres}  its dimensionless power spectrum  and
+observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is within five per cent up to $k = 1.0 h \mbox{Mpc}^{-1}$.
+The drop of power at higher $k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
+The fluctuations at low-$k$ come from the white noise imposed in the initial conditions, and it was show in \citet{2012MNRAS.419.2949N} and 
+\citet{2012MNRAS.423.2288H} that
+samples of a few hundreds of realizations average to the correct value.
+
+\begin{figure*}%[ht]
+  \begin{center}
+%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
+\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
+  \caption{Detail of a dark matter density projection, measured at $z=0$ in a RANGER4000 simulation.
+  Particles are shown in pink, halos in blue in the online version. The full projection is 64 times larger,
+  but does not provide much details on the structure on a paper format.
+    \label{fig:density}}
+\end{center}
+\end{figure*}
 
@@
   The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
-    \label{fig:power_highres} {\bf Should I show another redshift instead? Also, the legend box cuts the text: need to fix this. Finally, I could also show linear theory.}}
+    \label{fig:power_highres}}
 \end{center}
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10_new.eps}
-    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
+    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
   \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
@@
     ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
-    \label{fig:den_force_fracErr}
-    {\bf (Why is frac err radial slightly under 0? Problem in the units?)}}
+    \label{fig:den_force_fracErr}}
 \end{center}
@@
 the density is more homogeneous, and leads to mesh artefacts appearing
-in the density field {\bf (Should I show a density plot?)}.
+in the density field.
 
@@
 the discretize effect of the NGP causes an underestimate of the force of up to 60 per cent.
-{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}
+%{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}
 
@@
 We recall that {\small PMFAST} did not have this problem since it used CIC interpolation on both meshes.  
-{\bf (Anything else to add here?)}
+%{\bf (Anything else to add here?)}
 
@@
 \subsection{Constraining redshift jumps}
+\label{subsec:z_jumps}
 
@@
   \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
- The starting redshift ofwas raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
+ The starting redshift was raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
   The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
@@
 
-{\bf (Any other systematics effects we want to discuss here?)}
+%{\bf (Any other systematics effects we want to discuss here?)}

### 2012-07-09 paper.tex
Removed halo bias plot, updated extended pp_section with 1024 runs on scinet, with description of timing.

--- 2012-06-08 paper.tex
@@
 \title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M}
-\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
+\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, J. D.  Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
@@
 In parallel, many utilities and extensions have been added to the standard code package, 
-including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
+including a runtime halo finder, a tuneable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
 We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
@@
 \input{Extensions}
+\input{Initial}
+\input{PID}
+\input{Extendedpp}
 \input{Conclusion}

### 2012-07-11 paper_new.tex
Merged tex files, added rsoft power spectrum

\documentclass[useAMS,usenatbib]{mn2e}

% The usenatbib command allows the use of Patrick Daly's natbib.sty for
% cross-referencing.
%
% If you wish to typeset the paper in Times font (if you do not have the
% PostScript Type 1 Computer Modern fonts you will need to do this to get
% smoother fonts in a PDF file) then uncomment the next line
% \usepackage{Times}

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
\input{abbrev}
\input{psfig.sty}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{txfonts}

\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
Ilian T. Iliev$^{3}$, Hugh Merz$^{4}$, \newauthor
J.D. Emberson$^{1,5}$ and Vincent Desjacques$^{6,7}$\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Ontario, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$SHARCNET, Laurentian University, P3E 2C6, Ontario, Canada\\
$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
$^{6}$Institute for Theoretical Physics, University of Z\"{u}rich, Z\"{u}rich, CH 8057, Switzerland\\
$^{7}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}

\begin{document}

%\date{Accepted 1988 December 15. Received 1988 December 14; in original form 1988 October 11}
\date{\today}

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2011}

\maketitle

\label{firstpage}

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
is now one of the fastest highest performance public N-body code. 
Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
interactions inside and across neighboring cells.
Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
even at this scale.  In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
In parallel, many utilities and extensions have been added to the standard code package, 
including a runtime halo finder, a tuneable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

\begin{keywords}
N-body simulations --- Large scale structure of Universe --- Dark matter
\end{keywords}

%%%%%%%%%%%%%%%

\section{Introduction}

Many physical and astrophysical systems are subject to non-linear dynamics
and rely on N-body simulations to describe the evolution of bodies. 
One of the main field of application is the modelling of large scale structures, 
which are driven by the sole force of gravity. Recent observations of the 
cosmic microwave background \citep{2009ApJS..180..330K,2011ApJS..192...18K}, of galaxy clustering 
\citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 
2010MNRAS.401.1429D} of weak gravitational lensing \citep{2012AAS...21913001H, 2009ApJ...703.2232S}
and of supernovae redshift-distance relations all point towards a standard 
model of cosmology, in which dark energy and collision-less dark matter occupy 
more than 95 per cent of the total energy density of the universe. In such a 
paradigm, pure N-body code are perfectly suited to describe the dynamics, as 
long as baryonic physics is not very important, or at least we understand how 
the baryonic fluid feeds back on the dark matter structure. The next generation 
of measurements aim at constraining the cosmological parameters at the per cent 
level, and the theoretical understanding of the non-linear dynamics that govern 
structure formation heavily relies on numerical simulations.

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation 
scale can provide tight constraints on the dark energy equation of scale 
\citep{Eisenstein:2005su,2006PhRvD..74l3507T, 2007MNRAS.381.1053P,2009arXiv0902.4680S}. 
The most optimal estimates of the uncertainty requires the 
knowledge of the matter power spectrum covariance matrix, which is only accurate 
when measured from a large sample of N-body simulations \citep{2005MNRAS.360L..82R, 
2009ApJ...700..479T, 2011ApJ...726....7T}. For the same reasons, the most accurate 
estimates of weak gravitational lensing signal is obtained by propagating photons 
in past light cones that are extracted from simulated density fields 
\citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
Another area where large-scale N-body simulations have in recent years been 
instrumental are in simulations of early cosmic structures and reionization 
\citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
2011arXiv1107.4772I}. The reionization process is primarily driven by low-mass 
galaxies, which for both observational and theoretical reasons, need to be resolved 
in fairly large volumes, which demands simulations with a very large dynamic range.   
%{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
%accurate predictions/forecasts, etc.)}

The basic problem that is addressed with N-body codes is the time evolution of an ensemble of $N$ particles
that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operations, a cost that 
exceeds the memory and speed of current machines for large problems.
Solving the problem  on a mesh \citep{1981csup.book.....H} reduces to $O(N\mbox{log}N)$ the number of operations,
as it is possible to solve for the particle-mesh (PM) interaction with fast Fourier transforms techniques with high performance libraries such as {\small FFTW} \citep{FFTW3}.

With the advent of large computing facilities, parallel computations have now become 
common practice, and N-body codes have evolved both in performance and complexity. 
Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field. 
These often have the advantage to balance the work load across the computing units, which enable fast calculations even in high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to mesh-refined codes  \citep{1991ApJ...368L..23C}, 
which in the end are not designed to perform fast PM calculations on large scales.

%{\bf ( PM code by japanese? Others codes I should mention?)}

{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code designed specifically to optimize the PM algorithm,
both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T},
The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP} parallel computation.
This this setup enable the code to evolve large cosmological systems both rapidly and accurately, on relatively modest clusters.
One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
and  for computing the long range FFTs.
As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
in buffer zones is done at the same time as particles are passed to adjacent nodes,
and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
designed specifically to preserve a high processor load.

Since its first published version, {\small PMFAST} has evolved in many aspects. 
The first major improvement was to transform the volume decomposition in multi-node configurations 
from slabs to cubes. The problem with slabs is that they do not scale well to large runs: 
as the number of cells per dimension increases, the thickness of each slab shrinks rapidly,
until it reaches the hard limit of a single cell layer.  With this enhancement, the code name was changed to {\small CUBEPM}. Soon after, it incorporated particle-particle (pp) interactions at the sub-grid level, 
and was finally renamed {\small CUBEP3M}. The public package now includes a significant number of new features: the pp force
can be extended to an arbitrary range, the size of the redshift jumps can be constrained for improved accuracy during the first time steps,
a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification that can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features.
 {\small PMFAST} was equipped with a multi-time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available. 
 It also contains support for gas cosmological evolution through a TVD MHD module,
 and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.

{\small CUBEP3M} is one of the most competitive public N-body code and
has been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
 \citep{2010arXiv1008.3506Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}. 
It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

%\subsection{Simulation suites}
%\label{subsec:suites}

Since {\small CUBEP3M} is not new, the existing  accuracy and systematic tests were performed by different groups, on different machines,
and with different geometric and parallelization configurations. It is not an ideal situation in which to quantify the performance, and each test must be viewed as a
separate measurement that quantifies a specific aspect of the code. 
We have tried to keep to a minimum the number of such different runs, and although the detailed numbers vary 
with the problem size and the machines, the general trends are are rather universal.

Tests on constraints of the redshift step size and on improvements of the force calculations were performed 
on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
Each node contains 2 Quad Core Intel(R) Xeon(R) E5310 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
Each of these tests were performed on the same cluster, but with  different box sizes, starting redshifts and particle numbers.
Hereafter we refer to these simulation sets as the CITA configurations, and specify which parameters were used in each case.
Some complimentary runs were also performed on the SciNet GPC cluster \citep{Scinet}.

For tests of the code accuracy, of the non-Gaussian initial conditions generator and of the run time halo finder algorithm, 
we used a third simulation configuration series that was run at the Texas Advanced Computing Centre (TACC) on Ranger, a SunBlade 
x6420 system with AMD x86\_64 Opteron Quad Core 2.3 GHz `Barcelona' processors and Infiniband networking.
These RANGER4000 simulations evolved $4000^{3}$ particles 
from $z=100$ to $z=0$ with a box side of $3.2 h^{-1}\mbox{Gpc}$
on 4000 cores.

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
including much larger runs that were produced at various high performance computing centers;
section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
 to the default configuration, and conclude afterwards.

%\input{../cubep3m_paper/Structure}

\section{Review of the Code Structure}
\label{sec:structure}

An optimal large scale N-body code must address many challenges: minimize the memory footprint to allow larger dynamical range,
minimize the passing of information across computing nodes, reduce and accelerate the memory accesses to the large scale arrays, 
make efficient use of high performance libraries to speed up standard calculations like Fourier transforms, just to name a few.
In the realm of parallel programming, high efficiency  can be assessed when a high load is balanced across all processors
most of the time. In this section, we present the general strategies adopted to address these challenges\footnote{ 
Many originate directly from MPT and were preserved in {\small CUBEP3M};
those will be briefly mentioned, and we shall refer the reader to the original {\small PMFAST} paper for greater details.}.
We start with a walkthrough the code flow, and briefly discuss some specific sections that depart from standard N-body codes,
while referring the reader to future sections for detailed discussions on selected topics.

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on
a second level of cubical decomposition of the local volumes, the {\it tiles}. To make this possible, the fine grid arrays are constructed such as to support parallel memory access. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location. The force matching between the two meshes is performed by introducing a cutoff length, $r_{c}$, in the definition of the two force kernels. The value of $r_{c}=16$ fine cells was found to balance the communication 
overhead between processes and the accuracy of the match between the two meshes.

The computation of the short range force requires each tile to store the fine grid density of a region that includes a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than $r_{c}$, and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy.
This is basically to fully compensate for the coarse mesh calculations, whose CIC interpolation scheme 
reaches two coarse cells deep beyond the cutoff.

When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, center of mass, etc. 
It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.

%\subsection{Memory foot-print and communication strategy}
%\label{subsec:memory}

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.
In terms of memory, {\small CUBEP3M} uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like {\small GADGET}, which 
lies around $90$ per particles.
%\subsection{Code overview}
%\label{subsec:overview}

The code flow is presented in Fig. \ref{fig:structure} and \ref{fig:particle_mesh}.
Before entering the main loop, the code starts with an initialization stage, 
in which many declared variables are assigned default values,
the redshift checkpoints are read, the {\small FFTW} plans are created, and the {\small MPI} communicators are defined.
The phase-space array  is obtained from the output of the initial conditions generator,
and the force kernels on both grids are constructed from the specific geometry of the simulation.
For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
although they are actually distinct calls in the code.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized on many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the single core {\small MPI} process is active\footnote{Other libraries such as {\small P3DFFT} ({\tt http://code.google.com/p/p3dfft/}) currently permit
 an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.
 As in {\small PMFAST}, the particle position and velocity updates are performed in a leap frog scheme \citep{1981csup.book.....H}.

\begin{figure}
\begin{verbatim}
program cubep3m
   call initialize
   do
       call timestep
       call particle_mesh
       if(checkpoint_step) then
          call checkpoint
       elseif(last_step)
          exit
       endif
   enddo
   call finalize
end program cubep3m
\end{verbatim}
\caption{Overall structure of the code simplified for readability.}
\label{fig:structure}
\end{figure}

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position + apply random offset
   call link_list
   call particle_pass
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f
      if(pp = .true.) then       
         call link_list_pp
         call force_pp
         call update_velocity_pp
         if(extended_pp = .true.) then
            call link_list_pp_extended
            call force_pp_extended
            call update_velocity_pp_extended       
         endif
      endif
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm. We have included the section that concerns the standard pp and the extended pp force calculation, to illustrate that they follows similar linked-list logic. }
\label{fig:particle_mesh}
\end{figure}

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time. 
The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 
and clears the {\small MPI} communicators. We have collected those operations under the subroutine {\tt finalize} for concision.

Other constraints that need to be considered is that any decomposition geometry somehow 
limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
--  such as {\small CUBEP3M} -- requires the number of {\small MPI} processes to be a perfect cube.
Also,  since the decomposition is volumetric, as opposed to density dependent -- 
it suffers from load imbalance when probing deep into the non-linear regime.
This latter effect can be overcome by allocating extra memory per node, either for the full run,
or for the first part of it, follow by a restart configuration with more memory.

%%%%%%%%%%%%%%%%%%%
%\input{../cubep3m_paper/PoissonSolver}

\section{Poisson Solver}
\label{sec:Poisson}

This section reviews how Poisson's equation is solved on a double-mesh configuration. 
Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
a `cloud-in-cell' (CIC) interpolation for the coarse grid,  and a `nearest-grid-point' (NGP) interpolation 
for the fine grid \citep{1981csup.book.....H}. This choice is motivated by the fact that the most straightforward 
way to implement a P$^3$M algorithm on a mesh is to have exactly zero mesh force inside a grid, 
which is only true for the NGP interpolation. Although CIC generally has a smoother and more accurate force,
the pp implementation enhances the code resolution by almost an order of magnitude.

The code units inherit from \citep{2004NewA....9..443T} and are summarized here for completeness.
The comoving length of a fine grid cell is set to one,
such that the unit length in simulation unit is 
\begin{eqnarray}
1\mathcal{L} = a \frac{L}{N} 
\end{eqnarray}
where $a$ is the scale factor, $N$ is the total number of cells along one dimension,
and $L$ is the comoving volume in $h^{-1}\mbox{Mpc}$.
The mean comoving mass density is also set to unity in simulation units, 
which, in physical units, corresponds to 
\begin{eqnarray}
1\mathcal{D} = \rho_{m}(0) a^{-3} = \Omega_{m} \rho_{c} a^{-3} = \frac{3 \Omega_{m} H_{o}^{2}}{8 \pi G a^{3} }
\end{eqnarray}
$\Omega_{m}$ is the matter density, $H_{o}$ is the Hubble's constant, $\rho_{c}$ is the critical density today,
and $G$ is Newton's constant. The mass unit is found with $\mathcal{M} = \mathcal{DL}^{3}$.
Specifying the value of $G$ on the grid fixes the time unit, and with $G_{grid}$ = 1/$(6 \pi a)$,
we get:
\begin{eqnarray}
1 \mathcal{T} = \frac{2a^{2}}{3}\frac{1}{\sqrt{\Omega_{m}H_{o}^{2}}}
\end{eqnarray}
These choices completely determine the convertion between physical and simulation units.
For instance, the velocity units are given by $1\mathcal{V} = \mathcal{L}$/$\mathcal{T}$.

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
Gravity fields are curl-free, which allows us to relate the potential $\phi({\bf x})$ to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
We solve this equation in Fourier space, where we write:
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{-4 \pi G \tilde{\rho}({\bf k})}{k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}   
\label{eq:poisson_solution_pot}
\end{eqnarray}
and
\begin{eqnarray}
{\bf F}({\bf x}) = - m {\bf \nabla} \phi({\bf x}) 
\label{eq:Force_sol}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation at the final step, which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate, even though it requires four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions and define the force kernel {\boldmath $\omega$}$_{F}$ such as:
\begin{eqnarray}
 {\bf F}({\bf x}) =  \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
Because the gradient acting on \ref{eq:poisson_solution_pot} affects only un-prime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
\begin{eqnarray}
\mbox{\boldmath $\omega$}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

Following the spherically symmetric matching technique of MPT (section 2.1), 
we split  the force kernel into two components, for the short and long range respectively, and 
match the overlapping region with a polynomial. Namely, we have:
\begin{eqnarray}
\mbox{\boldmath $\omega$}_{s}(r) = \begin{cases} \mbox{\boldmath $\omega$}_{F}(r) -  \mbox{\boldmath $\beta$}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
0 & \mbox{otherwise} 
\end{cases}
\end{eqnarray}
and
\begin{eqnarray}
\mbox{\boldmath $\omega$}_{l}(r) = \begin{cases} \mbox{\boldmath $\beta$}(r) &\mbox{if  } \mbox{$r$ $\le$ $r_{c}$ } \\
 \mbox{\boldmath $\omega$}_{F}(r)  &\mbox{otherwise} 
\end{cases}
\end{eqnarray}
The vector $\mbox{\boldmath $\beta$}(r)$ is related to the fourth order polynomial that is used in the potential case described in MPT by
 $ \mbox{\boldmath $\beta$} = - {\bf \nabla} \alpha(r)$. The coefficients are found by matching the boundary conditions at $r_{c}$ up to the second derivative,
 and we get
  \begin{eqnarray}
   \mbox{\boldmath $\beta$}(r) = \bigg[ -\frac{7 r}{4 r_{c}^{3}} + \frac{3 r^{3}}{4 r^{5}}\bigg] \hat{\bf r}
  \end{eqnarray}

Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.  By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.

As mentioned in section \ref{sec:structure} and summarized in FIg. \ref{fig:particle_mesh},
the force kernels are first  in the code initialization stage.
Eq. \ref{eq:poisson_solution_force} is then solved with fast Fourier transform along each directions, 
 and is applied onto particles in the {\tt update\_velocity} subroutine.

The pp force is calculated during the fine mesh velocity update, which avoids loading the particle list twice and allows the operation to be threaded without significant additional work. During this process, the particles within a given fine mesh tile are first read in via the linked list, 
then their velocity is updated with the fine mesh force component, according to their location within the tile. 
In order to organize the particle-particle interactions, we proceed in constructing a set of threaded fine-cell linked list chains for each coarse cell. 
We then calculate the pairwise force between all particles that lie within the same fine mesh cell, exluding pairs whose separation is smaller than a softening length of 
one tenth of a grid cell; particles seperated by less than this distance have their 
particle-particle force set to zero.  As this proceeds, we accumulate the  pp force applied on each particle and then determine the maximum force element of the pp contribution, which is also taken into account when constraining the length of the global timestep.

Force softening is generally required by any code to prevent
large scattering as $r \rightarrow 0$ that can otherwise slow the calculation down, 
and to reduce the two-body relaxation, which can affect the numerical convergence. 
Many force softening scheme can be found in the litterature, including Plummer force, uniform or linear density profiles or the spline-softened model 
(see \citet{1993ApJ...409...60D} for a review). In the current case, a sharp force cutoff corresponds to a particle interacting with a hollow shell.
In comparison with other techniques, this  force softening is the easiest to code and the fastest to execute. 
Generally, it is desirable to match the smoothness of the force to the order of the time integration. 
A Plummer force is infinitely differentiable, which is a sufficient but not necessary condition for our 2nd order time integration.  
Also, one of the drawback of Plummer's softening is that the resolution degrades smoothly: the effects of the smoothing are present at all radii. 
In comparison, the uniform density and hollow shell alternatives both have the advantage that the deviations from $1/r^2$ are minimized. 
Although all the results presented in this paper were obtained with the sharp cutoff softening, other schemes can easily be adopted as these 
are typically a single line changes to the code. \citep{1993ApJ...409...60D} argues in favor on the uniform density profile scheme -- which is more physical --
and future developments of {\small CUBEP3M} will incorporate this option.

The choice of softening length is motivated by a trade off between accuracy and run time.
 Larger values reduce the structure formation but run quicker. 
 We show in Fig. \ref{fig:rsoft} the impact of changing this parameters on the power spectrum.
 For this test, we produced SciNet256 simulations, starting at $z=100$ and evolving to $z=0.5$, with a box size of $100 h^{-1}\mbox{Mpc}$.
 We also record in Table \ref{table:rsoft} the real run time for each trials, from where 
we see the strong effect of computation time the softening lenght has.
In this test, which is purposefully probing rather deep in the non-linear regime, reducing the softening length
to half it value ($1/20^{th}$ of a grid cell) double the run time. The gain in power spectrum is less than two per cent.
Similarly, increasing to $0.2$ the softening length reduces almost by a half the run time, but suffers from a five per cent loss in power at the resolution turn-around.
One tenth of grid cells seems to be the optimal choice in this trade off, however it should really be considered a free parameter.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_rsoft.eps}
  \caption{Dark matter power spectrum, measured at $z=0.5$ in a series of SciNet256 simulations in which the softening length is varied.
  The simulations started off at $z= 100$, and have a box size of $100 h^{-1}\mbox{Mpc}$.
    \label{fig:rsoft}}
\end{center}
\end{figure}

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the softening length.}
\begin{tabular}{|l|c|c|}
\hline 
$r_{soft}$         & time (h)   \\                 
\hline
 $0.5$ & 5.75 \\
 $0.3$ & 8.17\\
 $0.2$ & 10.09 \\
 $0.1$ & 18.67\\
 $0.05$ & 31.07 \\
% $0.03$ & XXX\\
% $0.01$ & XXX\\
\hline
\end{tabular}
\label{table:rsoft}
\end{center}
\end{table}

In addition, {\small PMFAST} could run with a different set of force kernels, described in MPT as the `least square matching' , 
which basically adjust the kernels on cell-by-cell basis based on minimization of the deviation with respect
to the Newtonian predictions. This was originally computed such as to optimize the force calculation for
the case where both grids are obtained from CIC interpolation. Moving to a mix CIC/NGP scheme
requires solving the system of equations with the new configuration, a straightforward operation.
 With the inclusion of the random shifting, it is not clear how much improvement 
one would recover from this other kernel matching.  It is certainly something
we will investigate and document in the near future.

Finally, a choice must be done concerning the longest range of the coarse mesh force. Gravity can be either a) an accurate $1/r^2$ force, as far as the volume allows, 
or b) modified to correctly match the periodicity of the boundary conditions. By default, the code is configured along to the second choice,
which accurately models the growth of structures at very large scales. However, detailed studies of gravitational collapse of a single large object would benefit 
from the first settings, even though the code is not optimal for such settings.

%%%%%%%%%%%%%%%%%%%%%%%%
%\input{../cubep3m_paper/ScalingPerformance}

\section{Scaling Performances}
\label{sec:scaling}

\begin{figure*}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_curie.eps}
    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_new.eps}
%  \vskip -1.2cm 
%  \vskip -0.5cm 
  \caption{Scaling of {\small CUBEP3M} on Curie fat nodes (left) and 
    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
    in each case. Dashed line indicates the ideal weak 
    scaling. The data are listed in Table \ref{summary_scaling_table}.
    \label{scaling}
% \vskip -0.9cm 
}
\end{center}
\end{figure*}

\begin{table*}%[ht]
  \vskip -0.5cm 
  \begin{center}
\caption{Scaling of {\small CUBEP3M} on Curie. Speedup is 
scaled to the smallest run.}
\label{summary_scaling_table}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size ($h^{-1}$Mpc)
\\[2mm]\hline
%\hline
32  &  1.00 & - &3.2 & $256^3$ & 256\\
256  & 7.21 & 8 &3.55 & $512^3$  & 512\\
864  & 25.63 & 27 &4.8 & $864^3$  & 864\\
2048  & 61.87 & 64 &26.48 & $2048^3$ & 2048 \\
\hline
\end{tabular}
\caption{Scaling of  {\small CUBEP3M} on Ranger. Speedup is scaled to the smallest run.}
\label{summary_scaling_table2}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size ($h^{-1}$Mpc)
\\[2mm]\hline
%\hline
864    & 1.00  & -    &258   & $1728^3$  & 6.3\\
4096   & 4.53  & 4.74 &320   & $3072^3$  & 11.4\\
10976  & 9.78  & 12.7 &845   & $5488^3$  & 20\\
21952  & 14.73 & 25.4 &561   & $5488^3$  & 20 \\
\hline
\end{tabular}
\end{center}
  \vskip -0.7cm 
\end{table*}

%the Ranger system at the Texas Supercomputing
%Center (in top 20 in the world) which is a SunBlade x6420 
%with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
% ``Barcelona'' processors and Infiniband networking. It has 
%a total of 62976 computing cores and 125952 GB of total 
%memory. Its nodes consist of 4 Quad Core processors and 32 GB 
%of shared RAM. For efficiency reasons (local memory access) we 
%typically use smaller MPI 'nodes' consisting of one Quad Core 
%processor and 8 GB of RAM.

The parallel algorithm of {\small CUBEP3M} is designed for `weak' 
scaling, i.e. if the number of cores and the problem size 
increase in proportion to each other, then for ideal scaling the 
wall-clock time should remain the same. This is to be in contrasted with `strong' 
scaling codes, whereby the same problem solved on more cores should take 
proportionately less wall-clock time. This weak scaling requirement 
is dictated by the problems we are typically investigating (very 
large and computationally-intensive) and our goals, which are to 
address such large problems in the most efficient way, rather than 
for the least wall-clock time. Furthermore, we recall that there is no explicit 
load balancing feature, thus the code is maximally efficient when the sub-domains
contain roughly an  number of particles. This is true for most
cosmological-size volumes that do not resolve too deep in the non-linear regime, 
but not for e.g. simulations of a single highly-resolved galaxy.

Because of the volumetric decomposition, the total number of {\small MPI} processes needs
to be a perfect cube. Also, for maximal resource usage, the number of tiles per node 
should be a multiple of the number of available {\small CPU}s per {\small MPI} process,
such that no core sits idle in the threaded block.
Given the available freedom in the parallel configuration, as long as the load is balanced, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes:
the information exchange between cores that are part of the same motherboard is generally much faster.
In addition, having fewer {\small MPI} processes reduces the total amount of buffer zones, 
freeing memory that can be used to increase the mesh resolution. 
As one probes deeper into the non-linear regime however, 
the formation of dense objects can cause memory problems in such configurations, and increasing 
the number of {\small MPI} processes helps to ensure memory locality,
especially in non-uniform memory access (NUMA) environments.

%Since the current Curie is using Intel Nehalem architecture, while 
%the thin Curie nodes will use the newer Westermere Intel architecture, 
%we also show the scaling of our code on the Westermere-based Lonestar 
%computer at the Texas Advanced Computing Centre

 The intermediary version of the code -- {\small CUBEPM} --
was first  ported to the IBM Blue Gene/L platform, and achieved 
weak-scaling up to 4096 processes (over a billion particles), with the N-body calculation only incurring a 10 per cent overhead 
at runtime (compared to 8 processes) for a balanced workload\footnote{\tt http://web.archive.org/web/20060925132146/ http://www-03.ibm.com/servers/deepcomputing/pdf/ Blue\_Gene\_Applications\_Paper\_CubePM\_032306.pdf}.  In order to 
accommodate the limited amount of memory available per processing core on the 
Blue Gene/L platform machines, it was necessary to perform the long range {\small MPI FFT}
with a volumetric decomposition \citep{3DFFT}.
Slab decomposition would have required a volume too large to fit in system 
memory given the constraints in the simulation geometry.

The scaling of {\small CUBEP3M}  was first  tested with a dedicated series 
of simulations -- the CURIE simulation suite-- by increasing the size and number of cores on the `fat' 
(i.e. large-memory) nodes of the  Curie supercomputer at the Tr\`{e}s Grand Centre de Calcul (TGCC) in France. 
 For appropriate direct comparison,
all these simulations were performed using the same particle mass 
($M_{\rm particle}=1.07\times10^{11}M_\odot$) and force resolution 
(softening length 50 $h^{-1}$kpc). The box sizes used range from 256 $h^{-1}$Mpc
to 2048 $h^{-1}$Mpc, and the number of particles from $256^3$ to $2048^3$.
Simulations were run on 32 up to 2048 computing cores, also starting from 
redshift $z=100$, and evolving until $z=0$. Our results are shown in Fig. \ref{scaling} and in Table \ref{summary_scaling_table}, and present excellent scaling, within 
$\sim3$ per cent of ideal, at least for up to 2048 cores.

We have also ran {\small CUBEP3M} on a much larger number of cores, 
from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
particles on Ranger and on JUROPA at the J\"ulich Supercomputing Centre in Germany, 
which is an Intel Xeon X5570 
(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.
Since it is not practical to perform dedicated scaling tests on such a large number of
computing cores, we instead list in Table \ref{summary_scaling_table2} 
the data directly extracted from production runs. We have found the 
code to scale within 1.5 per cent of ideal up to 4096 cores. 
For larger sizes ($\ge$10,976 cores), the scaling is less ideal,
due to increased communication 
costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
and load balancing issues, but still within $\sim20$ per cent of ideal. 
These first three Ranger runs were performed 
with 4 {\small MPI} processes and 4 threads per Ranger node (`4way')\footnote{For these very large runs, 
we used a NUMA script {\it tacc\_affinity}, specially-provided by the technical staff, 
that bind the memory usage to local sockets, thus ensuring memory affinity. 
This becomes important because the memory sockets per node 
(32 GB RAM/node on Ranger) are actually not equal-access. Generally, the local 
memory of each processor has much shorter access time.}.

Furthermore, due to the increasing clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain a number of particles
well above the 
average, thereby requiring more memory per {\small MPI} process
in order to run until the end. 
As a consequence,  throughout most of their late evolution, the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
was developed for us by the TACC technical staff and allowed to run more efficiently 
in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
performance somewhat, as does the imperfect load balancing that arises in such situations.
This can be seen in the rightmost point of FIg. \ref{scaling} (right panel), where the scaling 42 per cent below the ideal.
We note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus satisfactory even at extremely large number of cores.
 We expect the code to handle even larger problems efficiently, and is thus well suited to run on
  next generation Petascale systems.

Finally, we note that several special fixes had to be developed by  the TACC 
and JUROPA technical staff in order for our largest runs to work properly.
In particular, we encountered unexpected problems from software libraries such as 
{\small MPICH} and {\small FFTW} when applied to calculations of such 
unprecedented size.

%%%%%%%%%%%%%%%%%%%%%%%
%\input{../cubep3m_paper/Systematics}

\section{Accuracy and Systematics}
\label{sec:accuracy}

 This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
 We start with a demonstration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
 The halo mass function also assess the code capacity to model gravitational collapse, 
 but depends on the particular halo finder used.
 We thus post-pone the discussion on this aspect until in section \ref{sec:halo},
and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
 with comparisons to Newton's law of gravitation.

 \subsection{Density and Power Spectrum}
 \label{subsec:powerspectrum}

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
Fig. \ref{fig:density} presents  a 2-dimensional density projection of a RANGER4000 simulation, which evolved 4000$^3$ particles until redshift zero.
We then plot in Fig. \ref{fig:power_highres}  its dimensionless power spectrum  and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is within five per cent up to $k = 1.0 h \mbox{Mpc}^{-1}$.
The drop of power at higher $k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
The fluctuations at low-$k$ come from the white noise imposed in the initial conditions, and it was show in \citet{2012MNRAS.419.2949N} and 
\citet{2012MNRAS.423.2288H} that
samples of a few hundreds of realizations average to the correct value.

\begin{figure*}%[ht]
  \begin{center}
%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
  \caption{Detail of a dark matter density projection, measured at $z=0$ in a RANGER4000 simulation.
  Particles are shown in pink, halos in blue in the online version. The full projection is 64 times larger,
  but does not provide much details on the structure on a paper format.
    \label{fig:density}}
\end{center}
\end{figure*}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}. 
  The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

\subsection{Mesh force at grid distances}
\label{subsec:force}

The force test presented in MPT was carried by placing two particles at random locations on the grid, calculate the force between them, then iterate
over different locations. This pairwise force test is useful to quantify the accuracy on a cell-by-cell basis, but lacks the statistics that occur in an actual time step calculation.  
The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle during a single step, is presented in the top left panel of Fig. \ref{fig:den_force_fracErr}.
This force versus distance plot was obtained from a CITA128 realization, and the calculation proceeds in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, thus creating a `hole', compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the hole.

Particles in the same fine cell as the hole follow the exact $1/r^{2}$ curve. The scatter at 
  distances of the order of the fine grid is caused by the NGP interpolation scheme:
  particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
  but still feel the same mesh force at grid cell distances.
This  creates a discrepancy up to an order of magnitude, in loss or gain, depending on the location of the pair with respect to the centre of the cell.
As the separation approaches a tenth of the full box size or so, the force on the coarse mesh 
scatters significantly about Newton's law due to periodic boundary conditions. 
As mentioned at the end of section \ref{sec:Poisson}, the longest range of force kernel can either model accurately Newtonian gravity,
or the structure growth of the largest modes, but not both. For the sake of demonstrating the accuracy of the force calculation,
we chose the first option in this section.

It is not clear from these scatter plots whether the mean force felt  is below or under the predictions, hence we rebin the results
in 50 logarithmically spaced bins and compute the mean and standard deviation. 
We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the force force is otherwise
in excellent agreement with Newton's law.

The top right and middle right panels of Fig. \ref{fig:den_force_fracErr} show the fractional error on the force along the radial and tangential directions, also calculated from a single time step.
The mean of the radial component is up to 80 per cent away from the exact law at sub-grid distances, but quickly drops under ten per cent for larger distances. 
The transverse fractional error peaks exactly at the gris distance,  and is about two time smaller than the radial counterpart.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
  \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
    in a single time step. 
    ({\it top right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  In a full simulation run, the scatter averages over many time steps, 
    thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
    ({\it middle row}:) Top row organized in 50 logarithmic bins.  
    ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure*}

At grid scales, the fractional error is larger than {\small PMFAST}, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC. This prescription is responsible for the larger scatter about the theoretical value, but, as mentioned earlier, 
NGP interpolation is essential to our implementation of the pp part.
At the same time, the biggest problem with the straightforward pp force calculation is that the results 
are anisotropic and depend on the location of the fine mesh with respect 
to the particles. As an example, consider two particles on either side of a grid 
cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized one-grid cell separation.
 If, instead, the mesh was shifted such that they were
within the same cell, they would experience the much larger pp force. 
This is clearly seen in the top left panel of Fig. \ref{fig:den_force_fracErr}, where particles physically close, but in different grid cells, 
feel a force up to an order of magnitude too small.
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field.

In order to minimize these two systematic effect -- the large scatter and the anisotropy -- 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it is applied during the particle position update,
and suppresses the mesh behaviour that otherwise grows over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population, a necessary step 
if one needs to correlate the initial and final positions of the particles for instance,
or for hybrid dark matter -- MHD simulations.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the force kernels such as to provide a force as evenly balanced as possible, both at grid cell distances
and at the cutoff length ($r_{c}=16$ fine cells).
These adjustments are performed from the pairwise force test mentioned above (and described in MPT).
The bottom panels of Fig. \ref{fig:den_force_fracErr} show the effect of averaging over ten time steps, 
where the random offset is applied on the particles (and on the hole) at the end of each force calculation. 
We observe that the agreement extends at the per cent level down to $r \sim 1.5$, beyond which point
the discretize effect of the NGP causes an underestimate of the force of up to 60 per cent.
%{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}

We note that this inevitable loss of force in the NGP scheme is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a computational price,
but at least we have the option to run the code with high precision.

We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the otherwise default code configuration.
This test was performed with a CITA256 simulation of very large box size,
the output redshifts are very early (the upper most curve ($z=10$) was obtained after
about 60 time steps), such that the power spectrum should agree with linear theory up to the resolution limit.
Instead, we observe that the power spectrum grows completely wrong, due to the large scatter in the force from the fine mesh,
and to the anisotropic nature of the P$^3$M calculations mentioned above.
When these effects are not averaged over, the errors directly add up at each time step,
which explains why later time are worst.
We recall that {\small PMFAST} did not have this problem since it used CIC interpolation on both meshes.  
%{\bf (Anything else to add here?)}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is 1000 Mpc/$h$ per side. The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random offset that is normally applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

%\begin{figure*}%[ht]
 % \begin{center}
  %   \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N1.eps}
   % \includegraphics[width=3.2in]{graphs/densityForce_Zoom_N10.eps}
  % \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N1.eps}
  %  \includegraphics[width=3.2in]{graphs/densityForce_fracErr_N10.eps}
 % \caption{({\it top}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
 %   This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$.
  %  ({\it bottom}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
   % The left panels are calculated from a single time step, while the right panels show the average over time time steps.
 %   \label{fig:den_force_N10}}
%\end{center}
%\end{figure*}

%\begin{figure}%[ht]
 % \begin{center}
  %   \includegraphics[width=3.2in]{graphs/densityForce_ppext=10_rebin.eps}
  %\caption{ Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
   % This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$, averaged over 10 time steps.
   % Results are rebined to show how well the mean value follows the thoretical curve down to the grid size.
   % \label{fig:F_rebin}}
%\end{center}
%\end{figure}

\subsection{Constraining redshift jumps}
\label{subsec:z_jumps}

At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
If the expansion jump is too large, the size of the residual errors can become significant, and one can observe, for instance,
a growth of structure that does not match the predictions of  linear theory even at the largest scales.
One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
$\mbox{d}a/(a + \mbox{d}a)$ and is set to $0.05$ by default.  Generally, a simulation should start at a redshift high enough that
the initial dimensionless power spectrum is well under unity at all scales. This ensures that the Zel'dovich approximation
 holds at the per cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
 truncation error will be significant at the early time steps.

It is possible to reduce this effect, and thereby improve significantly 
the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of a series of CITA256 realizations that originate from the same initial conditions, 
and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
test runs in order to optimize a given configuration.  
As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
We should mention that with a proper use of the non-Gaussian initial conditions, it is possible to start the simulations at much later redshifts, where the time step sizes are dominated by a stronger mesh force.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
 The starting redshift was raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
\end{center}
\end{figure}

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the value of $r_{max}$. The tests were performed 
on the CITA Sunnyvale cluster, and general trends could vary slightly on other machines.}
\begin{tabular}{|l|c|c|}
\hline 
$r_{max}$         & time (h)   \\                 
\hline
 $0.1$ & 1.46 \\
 $0.06$ & 1.48\\
 $0.01$ & 1.67 \\
 $0.006$ & 1.91\\
 $0.003$ & 2.83 \\
 $0.002$ & 4.13\\
 $0.001$ & 8.15\\
\hline
\end{tabular}
\label{table:ra_max}
\end{center}
\end{table}

%{\bf (Any other systematics effects we want to discuss here?)}

%\input{../cubep3m_paper/Halos}

\section{Runtime Halo Finder}
\label{sec:halo}

%Spherical overdensities, search algorithm, 
%provided halo information, halo bias, comparison to PS and ST.
%{\bf (Ilian, You can lead the way here...)}

We have implemented a halo finding procedure, which we have developed 
based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
In the interest of speed and efficiency the halo catalogues are constructed 
on-the-fly at a pre-determined list of redshifts. The halo finding is 
massively-parallel and threaded based on the main {\small CUBEP3M} data structures 
discussed in section \ref{sec:structure}. The code first builds the 
fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
proceeds to search for all local density maxima above a certain
threshold (typically set to a factor of 100 above mean density) within the local tile. 
It then uses parabolic interpolation on the density field to determine more precisely
the location of the maximum within the densest cell, and records the peak 
position and value. %The halo centre determined this way agrees closely with 
%the centre-of-mass of the halo particles discussed below.

Once the list of peak positions is generated, they are sorted from the highest 
to the lowest density value. Then each of the halo candidates is inspected 
independently, starting with the highest peak. The grid mass is accumulated 
in spherical shells of fine grid cells surrounding the maximum, until the 
mean density within the halo drops below a pre-defined overdensity cutoff 
(usually set to 178 in units of the mean, in accordance with the top-hat 
collapse model). As we accumulate the mass, we remove it from the mesh, so that no 
mass element is double-counted. This method is thus inappropriate for finding 
sub-haloes since within this framework, those are naturally incorporated in their 
host haloes. Because the haloes are found on a discrete grid, it is 
possible, especially for those with lower mass, to overshoot the target overdensity.
To minimize this efect, we correct the halo mass and radius with an analytical density profile. 
We use the Truncated Isothermal Sphere (TIS) 
profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
$\sim130$, and a simple $1/r^2$ law for lower overdensities {\bf (lower or higher?)}. 
These yield a similar outer slope to the Navarro, Frenk and White 
\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
and matches well the virialization shock position given by the Bertschinger 
self-similar collapse solution \citep{1985ApJS...58...39B}.

Once the correct halo mass, radius and position are determined, we find all 
particles that are within the halo radius. Their positions and velocities are
used to calculate the halo centre-of-mass, bulk velocity, internal velocity 
dispersion and the three angular momentum components, all of which are then 
included in the final halo catalogues. We also calculate the total mass of
all particles within the halo radius, also listed in the halo data. This mass
is very close, but  typically slightly lower, than the halo mass calculated 
based on the gridded density field. The  centre-of-mass found this way closely follows 
that found from the peak location, which is based on the gridded mass distribution.

%Compared to Tinker
%44% low for 50 particles
%22% low for 400 particles
%12% low for 1000 particles

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$, based on a
    RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
    particles (solid, red in the online version). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced  from a RANGER4000 simulation at $z=0$ is shown in Fig. \ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, this one is based on the
SO search algorithm, whose masses are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10$ per cent for all haloes with masses corresponding to
1000 particles or more. Lower-mass haloes are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass haloes. It was shown, however, that using more sophisticated
halo finders (available only through post-processing calculations due to their heavier memory
footprint) it is possible to recover the expected mass function.

%A second test of the accuracy of the halo finder algorithm is to extract the halo power spectrum $P_{h}(k)$ and compare the halo bias
%with theoretical predictions. The halo bias is defined as $b(k) = \sqrt{P(k)/P_{h}(k)}$ and is shown in Fig. \ref{fig:halo}.
%The results are organized in four mass bins, where the lowest mass consists of 20-200 particles, the second bin 
%contains 200-2000 particles and so on. Comparisons with the linear predictions of \citet{2001MNRAS.323....1S} shows
%results consistent the known flaws of the model, namely that low-mass bias is over-predicted and high-mass bias is under-predicted \citep{2010ApJ...724..878T}.
%
%\begin{figure}%[ht]
%  \vskip -0.5cm 
%  \begin{center}
%    \includegraphics[width=3.2in]{graphs/bias.eps}
%  \end{center}
%  \caption{Halo bias, measured at $z=0$ from a RANGER4000 simulation with a side length of $3.2 h^{-1}$Gpc.
%  Bottom to top curves correspond to haloes of different masses, from light to massive. The four bins are decades
%  of halo masses; the first contains haloes made of 20-200 particles, the second of 200-2000 and so on.
%  The dashed lines are linear predictions from \protect \citet{2001MNRAS.323....1S}.
%    \label{fig:halo}}
%\end{figure}

%{\bf Show some examples and comparisions to analytical fits and other halo finders.}

%\input{../cubep3m_paper/Extensions}

\section{Beyond the standard configuration}
\label{sec:extensions}

The preceding descriptions and discussions apply to the standard configuration of the code, 
as described at the beginning of section \ref{sec:structure}. A few extensions have been recently developed
in order to enlarge the range of applications of {\small CUBEP3M}, and this section briefly 
describe the most important improvements.

\subsection{Initial conditions}
\label{subsec:init}

As mention in section \ref{sec:structure}, the code starts off by reading a set of initial conditions.
These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
local volume. Although many applications are typically based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.
For the Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshift.
Consequently, the size of the first few redshift jumps in such simulations are rather large, and therefore less accurate. 
One way to fix this is to constrain the size of the jumps, as discussed in section \ref{subsec:z_jumps}.
However,  the simulations take much more time to run, and this solution is therfore not adequate for 
very large configurations.

The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
is extended to include non-Gaussian features of the `local' form,
$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
to the Bardeen potential $\Phi({\bf x})$ (see \cite{2004PhR...402..103B} for a review). 
We adopted the CMB convention,
in which $\Phi$ is calculated immediately after the matter-
radiation equality (and not at redshift $z=0$ as in the large scale
structure convention). For consistency, $\phi({\bf x})$ is normalized
to the amplitude of scalar perturbations inferred by CMB measurements
($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
before the inclusion of the matter transfer function, and the initial 
particle positions and velocities are finally computed from $\Phi({\bf x})$ 
according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two RANGER4000 realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{2004PhRvD..69j3513S,2008PhRvD..78l3534T}.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The two curves represent two RANGER4000 realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator. The two curves differ at the sub-per cent level, as seen in the top panel,
  and the non-Gaussian realization is in much better agreement with the one-loop perturbation theory calculations.   
    \label{fig:init}}
\end{center}
\end{figure}

\subsection{Particle identification tags}
\label{PID}

A system of particle identification can be turned on, which basically allows to track each particle's trajectory
between checkpoints. Such a tool is useful for a number of applications, from reconstruction of halo merging history to tracking individual particle
trajectories.
The particle tag system has been implemented as an array of double integers, {\tt PID}, 
and assigns a unique integer to each particle during the initialization stage. The location of the tag on the {\tt PID} array 
matches the location of the corresponding particle on the {\tt xv} array, hence it acts as if the latter array had an extra dimension.
The location change only when particles exist the local volume, in which case the tag is sent along adjacent nodes 
with the particle in the {\tt pass\_particle} subroutine, and deleted particles result in deleted flags.
Similarly to the phase space array, the {\tt PID} array gets written to file at each particle checkpoint.
Since the array takes a significant amount of memory, we opted to store the tags in a seperate object
-- as opposed to adding  an extra dimension to the existing {\tt xv} array, such that it can be turned off 
when memory becomes an issue.

\subsection{Extended range of the pp force}
\label{subsec:extendedpp}

One of the main source of error in the calculation of the force occurs when on the smallest scales of the fine grid.
The approximation by which particles in a neighbouring mesh grid can be placed at the centre of the cell
is less accurate, which cause a maximal scatter around the exact $1/r^2$ law.
A solution to minimize this error consists in extending the pp force calculation outside a single cell,
which inevitably reintroduces a $N^2$ number of operations. Our goal is to add the flexibility to have a code
that runs slower, but produces results with a higher precision.

To allow this feature, we  have to choose how far outside a cell we want the exact pp force.  
Since the force kernels on both meshes are organized in terms of grids, the simplest way to implement this 
feature is to shut down the mesh kernels in a region of specified size, and allow the pp force to extend therein.
Concretely, these regions are constructed as cubic layers of fine mesh grids around a central cell; 
the freedom we have is to choose the number of such layers.

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_fracErr_ppext6} a force versus distance
 plot in a CITA128 realization, analogous to Fig. \ref{fig:den_force_fracErr}, but the pp force has been extended to  two layers of fine cells
 in the force test part of the code. 
 We observe that the scatter about the theoretical curve has reduced significantly, down to the few percent level, 
 and is still well balanced around the theoretical predictions.
 The fractional error on the radial and tangential components of the force, as seen in the right panel,
 are now at least 5 time smaller than in the default P$^{3}$M algorithm.
 When averaging over 10 time steps, we observe that improvement is relatively mild, as the calculation is already very accurate.

 \begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_new-1.eps}
     \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_N10_new-1.eps}
    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new-1.eps}
    \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
  in the same CITA128 realization as that shown in Fig. \ref{fig:den_force_fracErr}, 
  except that the exact pp force has been extended to two fine mesh layers around each particle
  in the force test code.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 2 fine grid, compared to Fig. \ref{fig:den_force_fracErr}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances, and further suppressed by averaging over many time steps.
  %\caption{
  ({\it top right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  This was also obtained over a single time step.
  ({\it bottom row}:) Same as top row, but averaged over ten time steps.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure*}

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = 50$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$,
 and with $512^{3}$ particles. We evolve these CITA512 realizations with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
 Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
 when extending  PM to P$^{3}$M first, and when adding successively one and two layers of fine cell mesh in which the pp force is extended.
We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the mesh calculations
become accurate enough as the distance increases. For this reason, it seems that a range of a two layers, suffices 
to reduce most of the undesired NGP scatter.

Extending the pp calculation comes at a price, since the number of operation scales as  $N^{2}$ in the sub-domain. 
This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
to evolved the particles to the final redshift. For instance, in our SciNet512 simulations, 
the difference between the default configuration and $N_{layer} = 1$ and $2$ is about a factor of $2.78$ in real time,
and about $6.5$ for $N_{layer} = 2$.
This number will change depending on the problem at hand and on the machine, and we recommand
to perform performance gain vs resource usage tests on smaller runs before performing large scale simulations
with the extended force.

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext_new_z1.eps,width=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force.
  These SciNet512 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$ 
  until $z=1.0$. The only difference between the runs is the ranges of the pp calculation. The inset shows details about the resolution turnaround, and the
  thick vertical line corresponds to the coarse mesh scale. }
    \label{fig:power}
  \end{center}
\end{figure}

%\begin{table}
%\begin{center}
%\caption{Scaling in {\small CPU} resources as a function of the range of the pp interaction.
%$N_{layers}$ refers to the number of fine mesh layers around a given cell, inside of which the force calculation
%is purely given by the pp contribution. }
%\begin{tabular}{|l|c|c|}
%\hline 
%             Type         & time (h)   \\
%                  \hline
%PM                         & 1.77 \\
%P$^{3}$M             & 2.09 \\
%\hline
%$N_{layers}$       &          \\
%\hline
% $1$ & 8.74 \\
% $2$ & 11.47\\
% $3$ & 14.30 \\
 %$4$ & 18.87\\
 %$5$ & 22.52\\
 %$6$ & 29.62\\
 %$7$ & 34.82\\
 %$8$ & 47.00\\
%\hline \hline
%\end{tabular}
%\label{table:cpu_pp_ext}
%\end{center}
%\end{table}

The power spectrum does not provide the complete story, and one of the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different SciNet512 runs. Fig. \ref{fig:MassFunction_extpp} presents this comparison at redshift $z = 1.0$. About 76,000 haloes were found in the end, yielding a halo number density of about $0.0225$ per $\mbox{Mpc/$h$}^{3}$. 
We observe that the simulation undershoot the Sheth-Tormen predictions, which is caused by the relatively low resolution of the configuration
comparered to the RANGER4000 run. 
We also observe that the pure PM code yields up 10 per cent less haloes than the P$^{3}$M version, over most of the mass range,
whereas the extended pp algorithm generally recovers  up to 10 per cent more haloes in the range $10^{11} - 10^{13}$.
The difference in performance between an extended pp  range of two vs four cells is rather mild, from where
we conclude that $N_{layers} = 2$ is the most optimal choice.

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/MassFunction_pp_ranges_new_z1.eps,width=0.49\textwidth}
  \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen, at $z = 1.0$. 
  The smallest haloes shown here have a mass equivalent to 100 particles, and fall in the lowest mass bin.
  ({\it bottom}:) Ratio between the different curves and that of the default P$^3$M configuration.}
    \label{fig:MassFunction_extpp}
  \end{center}
\end{figure}

%\input{../cubep3m_paper/Conclusion}

\section{Conclusion}

This paper describes {\small CUBEP3M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
 and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests
that have been performed. We also describe various utilities and extensions that come with the public release, 
including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
{\small CUBEP3M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
it has already been used  for a large number of scientific applications, and it is our hope that the current documentation will 
help the community in interpreting its outcome.
The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
compiling and running strategy is can be found on the CITA wiki page\footnote{\tt www.wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.

\section*{Acknowledgements}

The CITA simulations were run on the Sunnyvale cluster at CITA.
ITI was supported by The Southeast Physics
Network (SEPNet) and the Science and Technology Facilities Council
grants ST/F002858/1 and ST/I000976/1. 
Computations for the SciNet1024 runs were performed on the GPC supercomputer at the SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto. 
The authors acknowledge the
TeraGrid and the Texas Advanced Computing Center (TACC) at The
University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
providing HPC and visualization resources that have contributed to the
research results reported within this paper. ITI also acknowledges the Partnership for Advanced
Computing in Europe (PRACE) grant 2010PA0442 which supported the code
scaling studies. UEP and JDE are supported by the NSERC of Canada.
%{\bf ( Hugh, Vincent, any financial acknowledgements to add here?)}

\bibliographystyle{mn2e}
\bibliography{mybib3_new}{}
%\bibliographystyle{amsplain}

\bsp

\label{lastpage}

\end{document}

### 2012-07-31 paper_new.tex
Latest changes, with comments from ewveryone, and a new test with modifications to the fine kernel.

@@
 Ilian T. Iliev$^{3}$, Hugh Merz$^{4}$, \newauthor
-J.D. Emberson$^{1,5}$ and Vincent Desjacques$^{6,7}$\\
+J.D. Emberson$^{1,5}$ and Vincent Desjacques$^{6}$\\
 %\footnotemark[1]\thanks{This file has been amended to
@@
 $^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
-$^{6}$Institute for Theoretical Physics, University of Z\"{u}rich, Z\"{u}rich, CH 8057, Switzerland\\
-$^{7}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}
+$^{6}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}
 
@@
 \begin{abstract}
-This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
-is now one of the fastest highest performance public N-body code. 
-Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
-interactions inside and across neighboring cells.
-Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
-allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
+This paper presents {\small CUBEP3M}, a high performance cosmological N-body code
+and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
+a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
+{\small CUBEP3M} is fast, has a low memory imprint, and 
+has been run on up to $20,000$ cores, achieving close to ideal weak scaling
 even at this scale.
+It is well suited for a vast number of science applications that 
+require either large samples of non-linear realizations or 
+very large dark matter N-body simulations,
+including  cosmological reionization, baryonic acoustic oscillations, weak lensing or
+non-Gaussian statistics.
+%Since there is no explicit load balancing algorithm, the code runs optimally 
+%for homogeneous systems such as those which prevail in large scale structures studies. 
+%It has an interface with a MHD-TVD module and a radiative transfer code, 
+%which opens the door to a vast number of science applications.
+%or a large number of realizations for statistical studies.
+%one of the fastest highest performance public N-body code. 
+%Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
+%interactions inside and across neighbouring cells. 
+%Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
+%allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
+%even at this scale.
 In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
-In parallel, many utilities and extensions have been added to the standard code package, 
-including a runtime halo finder, a tuneable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
 We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
@@
 The basic problem that is addressed with N-body codes is the time evolution of an ensemble of $N$ particles
-that is subject to gravitational attraction. The brute force calculation requires $O(N^{2})$ operations, a cost that 
+that is subject to gravitational attraction. 
+The brute force calculation requires $O(N^{2})$ operations, a cost that 
 exceeds the memory and speed of current machines for large problems.
@@
 
-%{\bf ( PM code by japanese? Others codes I should mention?)}
-
-{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) is one of the first code designed specifically to optimize the PM algorithm,
-both in terms of speed and memory usage. It uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T},
-The long range gravitational force is computed on a  grid four times coarser, such as to minimize the {\small MPI} communication time
+Although such codes are needed to study systems that are spatially heterogeneous like individual clusters or haloes, AGNs or other compact objects,
+many applications are interested in studying large cosmological volumes in which the matter distribution is rather homogeneous.
+In such environments, the load balancing algorithm can be removed, and the effort
+can thereby be put towards speed, memory compactness and scalability.  
+{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) was one of the first code designed specifically to optimize the PM algorithm,
+both in terms of speed and memory usage, and uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T}.
+The long range gravitational force is computed on a  grid $4^{3}$ times coarser, such as to minimize the {\small MPI} communication time
 and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
 to be stored at a given time, allowing for {\small OPENMP} parallel computation.
-This this setup enable the code to evolve large cosmological systems both rapidly and accurately, on relatively modest clusters.
-One of the main advantage of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
-and the global {\small MPI} communications are cut down to the minimum: for passing particle at the beginning of each time step,
+This setup enables the code to evolve large cosmological systems both rapidly and accurately, even on relatively modest clusters.
+One of the main advantages of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
+and the global {\small MPI} communications are cut down to the minimum: for passing particles at the beginning of each time step,
 and  for computing the long range FFTs.
-As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
-in buffer zones is done at the same time as particles are passed to adjacent nodes,
-and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
-designed specifically to preserve a high processor load.
 
@@
  {\small PMFAST} was equipped with a multi-time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available.
- It also contains support for gas cosmological evolution through a TVD MHD module,
- and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
-
-{\small CUBEP3M} is one of the most competitive public N-body code and
-has been involved in a number of scientific applications over the last few years,
+
+ The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
+ {\small CUBEP3M} is therefore one of the most competitive and versatile public N-body code, 
+ and has been involved in a number of scientific applications over the last few years,
 spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
@@
 reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}. 
+Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
+and we expect that this will pave the way to an increasing number of science projects. 
+Notably, the fact that the force calculation is multi-layered makes the code extendible, 
+and opens the possibility to run {\small CUBEP3M} on hybrid CPU-GPU clusters.
 It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
@@
 We have tried to keep to a minimum the number of such different runs, and although the detailed numbers vary 
-with the problem size and the machines, the general trends are are rather universal.
+with the problem size and the machines, the general trends are rather universal.
 
@@
 and allows to optimize the code for higher resolution configurations.
-In terms of memory, {\small CUBEP3M} uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like {\small GADGET}, which 
-lies around $90$ per particles.
+In terms of memory, {\small CUBEP3M} can be configured such as the main footprint is dominated exclusively by the 
+particle phase space array, in which case it uses about $60$ bytes per particles. 
+For a fixed problem size, this can be achieved by increasing the number of tiles, thereby reducing the size of the local fine mesh arrays.
+There is a limit at which we can do such a volume breakdown, though, since the physical 
+volume on each tile must be much larger than the connection length between the two meshes (see section \ref{sec:Poisson}).
+%
+%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
+For comparison, (Lean) {\small GADGET-2} code uses around $90$ per particles.
 %\subsection{Code overview}
@@
 The phase-space array  is obtained from the output of the initial conditions generator,
-and the force kernels on both grids are constructed from the specific geometry of the simulation.
+and the force kernels on both grids are constructed by reading precomputed kernels which are then adjusted to the specific simulation size.
 For clarity, all these operations are collected under the subroutine call {\tt initialize} in Fig. \ref{fig:structure}, 
 although they are actually distinct calls in the code.
+
+As described in MPT, access to particles is accelerated with the use of linked lists, deletion of `ghost' particles
+in buffer zones is done at the same time as particles are passed to adjacent nodes,
+and the global FFTs are performed with a slab decomposition of the volumes via a special file transfer interface, 
+designed specifically to preserve a high processor load.
 
@@
 \caption{Overall structure of the two-level mesh algorithm.
-We have included the section that concerns the standard pp and the extended pp force calculation, to illustrate that they follows similar linked-list logic.
+We have included the section that concerns the standard pp and the extended pp force calculation, to illustrate that they follow similar linked-list logic.
 }
@@
 and writes them to file. 
+These restart files start with a small header that 
+contains the local  number of particles, the current redshift and time step, and the constraints on the time step jump
+from the previous iteration; the header is followed by the phase space of each particles. The particle identification tags are written 
+with a similar fashion in distinct files to simplify the post-processing coding. 
+This general I/O strategy allows for highly efficient {\small MPI} parallel read and write,
+by default in binary format for compactness, and has been shown to scale well to large data sets.
+
 Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time.
@@
 Also,  since the decomposition is volumetric, as opposed to density dependent -- 
-it suffers from load imbalance when probing deep into the non-linear regime.
-This latter effect can be overcome by allocating extra memory per node, either for the full run,
-or for the first part of it, follow by a restart configuration with more memory.
+it suffers from load imbalance whenever the particles are not evenly distributed.
+However, large scale cosmology problems are very weakly affected by this effect, which
+can generally  be overcome by allocating a mild amount of extra memory per node.
+As mentioned in section \ref{sec:scaling}, large runs can even be optimized by  allocating only a small amount for the first part of it,
+follow by a restart with a configuration that allows more memory per node.
 
@@
 \end{eqnarray}
-Because the gradient acting on \ref{eq:poisson_solution_pot} affects only un-prime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
+Because the gradient acting on \ref{eq:poisson_solution_pot} affects only unprime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
 \begin{eqnarray}
@@
 
-As mentioned in section \ref{sec:structure} and summarized in FIg. \ref{fig:particle_mesh},
+As mentioned in section \ref{sec:structure} and summarized in Fig. \ref{fig:particle_mesh},
 the force kernels are first  in the code initialization stage.
-Eq. \ref{eq:poisson_solution_force} is then solved with fast Fourier transform along each directions, 
+Eq. \ref{eq:poisson_solution_force} is then solved with fast Fourier transforms along each direction, 
  and is applied onto particles in the {\tt update\_velocity} subroutine.
@@
 In order to organize the particle-particle interactions, we proceed in constructing a set of threaded fine-cell linked list chains for each coarse cell. 
-We then calculate the pairwise force between all particles that lie within the same fine mesh cell, exluding pairs whose separation is smaller than a softening length of 
-one tenth of a grid cell; particles seperated by less than this distance have their 
+We then calculate the pairwise force between all particles that lie within the same fine mesh cell, excluding pairs whose separation is smaller than a softening length $r_{soft}$; particles separated by less than this distance have their
 particle-particle force set to zero.
-As this proceeds, we accumulate the  pp force applied on each particle and then determine the maximum force element of the pp contribution, which is also taken into account when constraining the length of the global timestep.
+As this proceeds, we accumulate the  pp force applied on each particle and then determine the maximum force element of the pp contribution, which is also taken into account when constraining the length of the global time step.
 
@@
 and to reduce the two-body relaxation, which can affect the numerical convergence. 
-Many force softening scheme can be found in the litterature, including Plummer force, uniform or linear density profiles or the spline-softened model 
+Many force softening schemes can be found in the literature, including Plummer force, uniform or linear density profiles or the spline-softened model 
 (see \citet{1993ApJ...409...60D} for a review). In the current case, a sharp force cutoff corresponds to a particle interacting with a hollow shell.
@@
 Generally, it is desirable to match the smoothness of the force to the order of the time integration. 
-A Plummer force is infinitely differentiable, which is a sufficient but not necessary condition for our 2nd order time integration.  
-Also, one of the drawback of Plummer's softening is that the resolution degrades smoothly: the effects of the smoothing are present at all radii. 
+A Plummer force is infinitely differentiable, which is a sufficient but not necessary condition for our $2^{nd}$ order time integration.  
+Also, one of the drawbacks of Plummer's softening is that the resolution degrades smoothly: the effects of the smoothing are present at all radii. 
 In comparison, the uniform density and hollow shell alternatives both have the advantage that the deviations from $1/r^2$ are minimized. 
 Although all the results presented in this paper were obtained with the sharp cutoff softening, other schemes can easily be adopted as these 
-are typically a single line changes to the code. \citep{1993ApJ...409...60D} argues in favor on the uniform density profile scheme -- which is more physical --
+are typically single line changes to the code. \citet{1993ApJ...409...60D} argues in favour on the uniform density profile scheme -- which is more physical --
 and future developments of {\small CUBEP3M} will incorporate this option.
@@
  We also record in Table \ref{table:rsoft} the real run time for each trials, from where 
-we see the strong effect of computation time the softening lenght has.
+we see the strong effect of computation time the softening length has.
 In this test, which is purposefully probing rather deep in the non-linear regime, reducing the softening length
@@
 Similarly, increasing to $0.2$ the softening length reduces almost by a half the run time, but suffers from a five per cent loss in power at the resolution turn-around.
-One tenth of grid cells seems to be the optimal choice in this trade off, however it should really be considered a free parameter.
+One tenth of grid cell seems to be the optimal choice in this trade off, and is therefor the default value, however it should really be considered a free parameter.
 
@@
 
-In addition, {\small PMFAST} could run with a different set of force kernels, described in MPT as the `least square matching' , 
+In addition, {\small PMFAST} could run with a different set of force kernels, described in MPT as  `least square matching' , 
 which basically adjust the kernels on cell-by-cell basis based on minimization of the deviation with respect
-to the Newtonian predictions. This was originally computed such as to optimize the force calculation for
+to Newtonian predictions. This was originally computed such as to optimize the force calculation for
 the case where both grids are obtained from CIC interpolation. Moving to a mix CIC/NGP scheme
@@
 which accurately models the growth of structures at very large scales. However, detailed studies of gravitational collapse of a single large object would benefit 
-from the first settings, even though the code is not optimal for such settings.
+from the first setting, even though the code is not meant to evolve such systems that generally require load balance control.
 
@@
 load balancing feature, thus the code is maximally efficient when the sub-domains
-contain roughly an  number of particles. This is true for most
+contain roughly an equal number of particles. This is true for most
 cosmological-size volumes that do not resolve too deep in the non-linear regime, 
@@
 such that no core sits idle in the threaded block.
-Given the available freedom in the parallel configuration, as long as the load is balanced, it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes:
+Given the available freedom in the parallel configuration, as long as the load is balanced, 
+it is generally good practice to maximize the number of {\small OPENMP} threads and minimize the number of {\small MPI} processes:
 the information exchange between cores that are part of the same motherboard is generally much faster.
@@
 freeing memory that can be used to increase the mesh resolution. 
+However, it has been observed that for the case of non-uniform memory access (NUMA) systems, 
+the code is optimize when only the cores that share the same socket should be  {\small OPENMP} threaded.
 As one probes deeper into the non-linear regime however, 
@@
 the number of {\small MPI} processes helps to ensure memory locality,
-especially in non-uniform memory access (NUMA) environments.
+especially in NUMA environments.
 
@@
 performance somewhat, as does the imperfect load balancing that arises in such situations.
-This can be seen in the rightmost point of FIg. \ref{scaling} (right panel), where the scaling 42 per cent below the ideal.
+This can be seen in the rightmost point of Fig. \ref{scaling} (right panel), where the scaling is $42$ per cent below the ideal.
 We note that we still get $\sim1.5$ speedup
@@
  but depends on the particular halo finder used.
- We thus post-pone the discussion on this aspect until in section \ref{sec:halo},
+ We thus post-pone the discussion on this aspect until section \ref{sec:halo},
 and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
@@
 
-One of the most reliable way to assess the simulation's accuracy at evolving particles
+One of the most reliable ways to assess the simulation's accuracy at evolving particles
 is to measure the density power spectrum at late time, and compare to non-linear prediction. 
@@
 We then plot in Fig. \ref{fig:power_highres}  its dimensionless power spectrum  and
-observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is within five per cent up to $k = 1.0 h \mbox{Mpc}^{-1}$.
+observe that the agreement with the non-linear prediction of \citet{2003MNRAS.341.1311S} is within five per cent up to $k = 1.0 h \mbox{Mpc}^{-1}$.
+The code exhibits  a $\sim 5$ per cent over estimate compared 
+  to the theory for $ 0.2 < k < 0.6 h\mbox{Mpc}^{-1}$, an effect generally caused by inaccuracies at the early time steps. It can be removed by starting the code at a later redshift,
+  but then the code has less time to relax, and other quantities like the halo mass functions or the four point functions become less accurate.
 The drop of power at higher $k$ is partly caused by the finite mesh resolution, partly from expected deviations about the predictions.
@@
   \caption{Detail of a dark matter density projection, measured at $z=0$ in a RANGER4000 simulation.
-  Particles are shown in pink, halos in blue in the online version. The full projection is 64 times larger,
+  Particles are shown in pink, haloes in blue in the online version. The full projection is 64 times larger,
   but does not provide much details on the structure on a paper format.
@@
   \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
-  compared to the non-linear predictions of {\small CAMB}. 
+  compared to the non-linear predictions of {\small HALOFIT}. 
   The vertical line corresponds to the scale of the coarse mesh, while the dotted line represents the Poisson noise.
+  The particle masses are of $5.68 \times 10^{10} M_{\odot}$. 
     \label{fig:power_highres}}
@@
 
-The force test presented in MPT was carried by placing two particles at random locations on the grid, calculate the force between them, then iterate
+The force test presented in MPT was carried by placing two particles at random locations on the grid, calculating the force between them, then iterating
 over different locations. This pairwise force test is useful to quantify the accuracy on a cell-by-cell basis, but lacks the statistics that occur in an actual time step calculation.  
@@
 or the structure growth of the largest modes, but not both. For the sake of demonstrating the accuracy of the force calculation,
-we chose the first option in this section.
-
-It is not clear from these scatter plots whether the mean force felt  is below or under the predictions, hence we rebin the results
+we chose the first option in this section, but this is not how the code would normally be used.
+
+The top right panel of Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial direction (top) and the fractional 
+tangential contribution (bottom), also calculated from a single time step.
+The fractional error of the radial component is up to 60 per cent at sub-grid distances, but remains under ten per cent for  distances larger than a few cells. 
+The transverse fractional error peaks exactly at the grid distance,  and is about two time smaller than the radial counterpart.
+
+Although these scatter plots show the contributions from individual particles and cells, it is not clear
+whether the mean radial force felt is higher or lower the predictions, hence we rebin the results
 in 50 logarithmically spaced bins and compute the mean and standard deviation. 
 We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
-there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the force force is otherwise
+there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the numerical calculations are otherwise
 in excellent agreement with Newton's law.
-
-The top right and middle right panels of Fig. \ref{fig:den_force_fracErr} show the fractional error on the force along the radial and tangential directions, also calculated from a single time step.
-The mean of the radial component is up to 80 per cent away from the exact law at sub-grid distances, but quickly drops under ten per cent for larger distances. 
-The transverse fractional error peaks exactly at the gris distance,  and is about two time smaller than the radial counterpart.
+The transverse force is somehow ill-defined, since it is by definition a positive number.
+Since we know that it averages out to zero over many time steps, we plot only the scatter about the mean, represented in the figure by the solid line.
 
@@
     \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new-1.eps}
+    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
     \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new-1.eps}
+%    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
+    \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
     \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
-  \caption{({\it top left}:) Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
+%    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
+  \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
     This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
     in a single time step. 
-    ({\it top right}:) Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+    ({\it top right}:) Fractional error on the force in the P$^3$M algorithm in the radial direction (top) and fractional tangential contribution (bottom).
   In a full simulation run, the scatter averages over many time steps, 
     thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
-    ({\it middle row}:) Top row organized in 50 logarithmic bins.  
+    ({\it middle row}:) Top row organized in 50 logarithmic bins; we plot only the scatter about the mean for the tangential contribution.  
     ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
@@
 
-In order to minimize these two systematic effect -- the large scatter and the anisotropy -- 
+In order to minimize these two systematic effects -- the large scatter and the anisotropy -- 
 we randomly shift the particle distribution relative to the mesh by a small
@@
 and suppresses the mesh behaviour that otherwise grows over multiple time steps.
-It is possible to shift back the particles at the end of each time steps,
+It is possible to shift back the particles at the end of each time step,
 which prevents a random drift of the whole population, a necessary step 
@@
 We observe that the agreement extends at the per cent level down to $r \sim 1.5$, beyond which point
-the discretize effect of the NGP causes an underestimate of the force of up to 60 per cent.
+the discretization effect of the NGP causes a cumulative underestimate of the force, which reaches up 60 per cent.
 %{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}
 
-We note that this inevitable loss of force in the NGP scheme is one of the driving argument to extend the pp force outside the fine mesh cell,
+We note that this inevitable loss of force in the NGP scheme is one of the driving arguments to extend the pp force outside the fine mesh cell,
 since the scattering about the actual $1/r^{2}$ law drops rapidly as the distance increases.
@@
 When these effects are not averaged over, the errors directly add up at each time step,
-which explains why later time are worst.
+which explains why later times are worst.
 We recall that {\small PMFAST} did not have this problem since it used CIC interpolation on both meshes.  
@@
 %\end{figure}
+Although most of the code tests presented in this paper were obtained with the configuration described above, 
+we present here a recent improvement in the fine force kernel that compensates for the systematic underestimate
+at grid distances. Basically, the idea is to boost the small distance elements of the kernel by a small amount,
+which somehow corrects for the loss observed in the bottom panels of Fig. \ref{fig:den_force_fracErr}.
+We show here the impact of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
+the force from the first two layers of neighbours is boosted by $60$ per cent, and c) by $80$ per cent.
+We observe that we are able to improve the power at turn around by almost a factor of two, which doubles the resolution.  
+This test was performed in a series of SciNet256 simulations with a volume of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
+We see from the figure that in the original configuration, the departure from the non-linear predictions occur at $k\sim 3.0 h\mbox{Mpc}^{-1}$,
+whereas the third trial allows for a resolution down to $k\sim 6.0 h\mbox{Mpc}^{-1}$. 
+%We note that these trials have not yet been 
+%tried on very large runs, and we recommend that some more testing should be done to assess the robustness of this correction.
+
+\begin{figure}%[ht]
+  \begin{center}
+    \includegraphics[width=3.2in]{graphs/power_fudge.eps}
+  \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is 150 Mpc/$h$ per side. 
+  The thick solid line represents predictions from {\small CAMB}, the thin solid line is the default configuration of the code, and the three other lines represent
+  three different corrections that are applied at grid distances to compensate for the systematic force underestimation seen in the bottom panels of Fig. \ref{fig:den_force_fracErr}. }
+\end{center}
+\end{figure}
 
@@
 
-At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
+At early stages of the simulation, the density field is homogenous, causing the force of gravity to be
 rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
@@
 and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
-We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
-allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
+We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to $0.006$ 
+allows the simulation to recover about $30$ per cent of dimensionless power at the turn over scale, in this simulation configuration.
 This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
@@
 As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
-In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
-We should mention that with a proper use of the non-Gaussian initial conditions, it is possible to start the simulations at much later redshifts, where the time step sizes are dominated by a stronger mesh force.
+In this test case, reducing further at $0.001$ shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
+The default configuration of the code is set to $r_{max}=0.05$, but at the light of this recent test, we recommend reducing to the value $0.01$ or even $0.005$.
+We also mention here that with a proper use of second order initial conditions (also see section \ref{subsec:init}), it is possible to start the simulations at much later redshifts
+without loosing much accuracy.
 
@@
 \caption{Scaling in {\small CPU} resources as a function of the value of $r_{max}$. The tests were performed 
-on the CITA Sunnyvale cluster, and general trends could vary slightly on other machines.}
+on the CITA Sunnyvale cluster, and general trends could vary slightly on other machines. The time tabulated on the right column
+corresponds to the full run time of the CITA256 simulations, which evolved $128^{3}$ particles from $z = 200$ to $z = 0$.}
 \begin{tabular}{|l|c|c|}
@@
 Once the list of peak positions is generated, they are sorted from the highest 
-to the lowest density value. Then each of the halo candidates is inspected 
+to the lowest density. Then each of the halo candidates is inspected 
 independently, starting with the highest peak. The grid mass is accumulated 
@@
 possible, especially for those with lower mass, to overshoot the target overdensity.
-To minimize this efect, we correct the halo mass and radius with an analytical density profile. 
+To minimize this effect, we correct the halo mass and radius with an analytical density profile. 
 We use the Truncated Isothermal Sphere (TIS) 
 profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
-$\sim130$, and a simple $1/r^2$ law for lower overdensities {\bf (lower or higher?)}. 
+$\sim130$, and a simple $1/r^2$ law for lower overdensities lower. 
 These yield a similar outer slope to the Navarro, Frenk and White 
 \citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
-and matches well the virialization shock position given by the Bertschinger 
-self-similar collapse solution \citep{1985ApJS...58...39B}.
-
-Once the correct halo mass, radius and position are determined, we find all 
+and matches well the virialization shock position given by the
+self-similar collapse solution of \citet{1985ApJS...58...39B}.
+
+After the correct halo mass, radius and position are determined, we find all 
 particles that are within the halo radius. Their positions and velocities are
@@
     precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
+    The particle masses are of $5.68 \times 10^{10} M_{\odot}$.
     \label{mf}}
@@
 to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
-by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
-SO halo finder, which misses some of the low-mass haloes. It was shown, however, that using more sophisticated
+by $\sim40$ per cent for 50 particles. This is largely due to the grid-based nature of our
+SO halo finder, which misses some of the low-mass haloes. It was shown that using more sophisticated
 halo finders (available only through post-processing calculations due to their heavier memory
-footprint) it is possible to recover the expected mass function.
+footprint) it is possible to recover most of the expected number count.
 
@@
 in order to enlarge the range of applications of {\small CUBEP3M}, and this section briefly 
-describe the most important improvements.
+describes the most important improvements.
 
@@
 These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
-local volume. Although many applications are typically based on initial conditions that follow Gaussian statistics,
+local volume. \footnote{We remind that for Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshifts.
+Consequently, the size of the first few redshift jumps in such simulations can become rather large, and therefore less accurate, 
+which is why we must carefully constrain the size of the jumps, as discussed in section \ref{subsec:z_jumps}.
+Another solution to this issue is to use second order perturbation theory to generate the initial conditions,
+which is not implemented yet in the public package.}
+Although many applications are typically based on initial conditions that follow Gaussian statistics,
 we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
 This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.
-For the Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshift.
-Consequently, the size of the first few redshift jumps in such simulations are rather large, and therefore less accurate. 
-One way to fix this is to constrain the size of the jumps, as discussed in section \ref{subsec:z_jumps}.
-However,  the simulations take much more time to run, and this solution is therfore not adequate for 
-very large configurations.
 
@@
 but one of which had non-Gaussian initial conditions.
-We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
-agrees much better with one loop perturbation theory \citep{2004PhRvD..69j3513S,2008PhRvD..78l3534T}.
+We see that the difference between the two power spectra is at the sub-per cent level, and that the ratio of the two power spectra
+are well described with one loop perturbation theory \citep{2004PhRvD..69j3513S,2008PhRvD..78l3534T}.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
+    \includegraphics[width=3.2in]{graphs/dm_power_ngp_z0_fNL0_50_w_1loop_PT.eps}
   \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
@@
 The two curves differ at the sub-per cent level, as seen in the top panel,
-  and the non-Gaussian realization is in much better agreement with the one-loop perturbation theory calculations.   
+  and the one-loop perturbation theory calculations accurately describes the ratio between the two curves up to $k\sim 0.4h\mbox{Mpc}^{-1}$ in this case.   
     \label{fig:init}}
@@
 
-One of the main source of error in the calculation of the force occurs when on the smallest scales of the fine grid.
+One of the main sources of error in the calculation of the force occurs from the PM interaction at the smallest scales of the fine grid.
 The approximation by which particles in a neighbouring mesh grid can be placed at the centre of the cell
-is less accurate, which cause a maximal scatter around the exact $1/r^2$ law.
+is less accurate, which causes a maximal scatter around the exact $1/r^2$ law.
 A solution to minimize this error consists in extending the pp force calculation outside a single cell,
-which inevitably reintroduces a $N^2$ number of operations. Our goal is to add the flexibility to have a code
-that runs slower, but produces results with a higher precision.
+which inevitably reintroduces a number of operations that scales as $N^2$. 
+Our goal is to add the flexibility to have a code that runs slower, but produces results with a higher precision.
 
@@
  to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. 
-We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
- we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
+ We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cell for which we killed the mesh kernels,
+ we compute the separation and the force between each pair and update their velocities simultaneously with Newton's third law. 
  To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
@@
  The fractional error on the radial and tangential components of the force, as seen in the right panel,
- are now at least 5 time smaller than in the default P$^{3}$M algorithm.
+ are now at least five times smaller than in the default P$^{3}$M algorithm.
  When averaging over 10 time steps, we observe that improvement is relatively mild, as the calculation is already very accurate.
@@
     \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_new-1.eps}
+    \includegraphics[width=3.0in]{graphs/DeltaF_pp2.eps}
+%    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_new-1.eps}
      \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_N10_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new-1.eps}
+    \includegraphics[width=3.0in]{graphs/DeltaF_pp2_N10.eps}
+%    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new-1.eps}
     \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, compared with the exact $1/r^{2}$ law,
@@
   %\caption{
-  ({\it top right}:) Fractional error on the force in the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
+  ({\it top right}:) Fractional error on the force in the P$^3$M algorithm, in the radial direction (top) and fractional tangential contribution (bottom).
   This was also obtained over a single time step.
@@
  We generate a set of initial conditions at a starting redshift of $z = 50$, with a box size equal to $ 150 h^{-1}\mbox{Mpc}$,
- and with $512^{3}$ particles. We evolve these CITA512 realizations with different ranges for the pp calculation, and compare 
+ and with $512^{3}$ particles. We evolve these SciNet512 realizations with different ranges for the pp calculation, and compare 
  the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
@@
 We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the mesh calculations
-become accurate enough as the distance increases. For this reason, it seems that a range of a two layers, suffices 
+become accurate enough as the distance increases. For this reason, it seems that a range of two layers suffices 
 to reduce most of the undesired NGP scatter.
 
-Extending the pp calculation comes at a price, since the number of operation scales as  $N^{2}$ in the sub-domain. 
-This cost is best capture by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
+Extending the pp calculation comes at a price, since the number of operations scales as  $N^{2}$ in the sub-domain. 
+This cost is best captured by the increase of real time required by a fixed number of dedicated  {\small CPU}s 
 to evolved the particles to the final redshift. For instance, in our SciNet512 simulations, 
-the difference between the default configuration and $N_{layer} = 1$ and $2$ is about a factor of $2.78$ in real time,
+the difference between the default configuration and $N_{layer} = 1$  is about a factor of $2.78$ in real time,
 and about $6.5$ for $N_{layer} = 2$.
-This number will change depending on the problem at hand and on the machine, and we recommand
+This number will change depending on the problem at hand and on the machine, and we recommend
 to perform performance gain vs resource usage tests on smaller runs before performing large scale simulations
@@
 
-The power spectrum does not provide the complete story, and one of the most relevant way to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different SciNet512 runs.
+The power spectrum does not provide the complete story, and one of the most relevant ways to quantify the improvement of the calculation is to measure the impact on the halo mass function from these different SciNet512 runs.
 Fig.
@@
 We observe that the simulation undershoot the Sheth-Tormen predictions, which is caused by the relatively low resolution of the configuration
-comparered to the RANGER4000 run. 
-We also observe that the pure PM code yields up 10 per cent less haloes than the P$^{3}$M version, over most of the mass range,
+compared to the RANGER4000 run. 
+We also observe that the pure PM code yields up to 10 per cent less haloes than the P$^{3}$M version, over most of the mass range,
 whereas the extended pp algorithm generally recovers  up to 10 per cent more haloes in the range $10^{11} - 10^{13}$.
@@
 The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
-compiling and running strategy is can be found on the CITA wiki page\footnote{\tt www.wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.
+compiling and running strategy is can be found on the CITA wiki page\footnote{\tt wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.
 
@@
 scaling studies. UEP and JDE are supported by the NSERC of Canada.
-%{\bf ( Hugh, Vincent, any financial acknowledgements to add here?)}
+VD acknowledges support by the Swiss National Science Foundation.
+%{\bf ( Hugh, any financial acknowledgements to add here?)}
 

### 2012-07-31 paper_new.tex
Improved the text about the kernel correction

@@
 the force from the first two layers of neighbours is boosted by $60$ per cent, and c) by $80$ per cent.
-We observe that we are able to improve the power at turn around by almost a factor of two, which doubles the resolution.  
+We observe that with the trial a), we are able to improve the power at turn around by a factor of $1.5$, which almost doubles the resolution.  
 This test was performed in a series of SciNet256 simulations with a volume of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
 We see from the figure that in the original configuration, the departure from the non-linear predictions occur at $k\sim 3.0 h\mbox{Mpc}^{-1}$,
-whereas the third trial allows for a resolution down to $k\sim 6.0 h\mbox{Mpc}^{-1}$. 
+whereas the first trial allows for a similar resolution down to $k\sim 4.5 h\mbox{Mpc}^{-1}$. 
+Other correction schemes are more aggressive, and although they recover power at even smaller scales, 
+they exhibit a small overestimate compare to non-linear predictions in the range $2.0 < k < 4.0 h\mbox{Mpc}^{-1}$, 
+and more testing should be done prior to using these corrections.
+
+We therefore recommend that the fine force kernel should be considered as an adjustable parameter;
+`no correction' is considered as the  conservative mode, which is accurate only up to the resolution turn around 
+observed in the dimensionless power spectrum. Higher resolution can be achieved with the modified fine force kernel following `trial a',
+and should be considered as an actual resolution improvement, with no cost in accuracy nor in run time;
+`trial b' and `trial c' should be considered as too aggressive. We note that we have not exhausted the list of possible correction,
+and that future work in that direction might result in even better resolution.
+
 %We note that these trials have not yet been 

### 2012-08-01 paper_new.tex
Corrected Fig. 15

@@
 
-\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M}
-\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, JD Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
+\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP$^3$M]{{\small} High Performance P$^{3}$M N-body code: CUBEP$^3$M}
+\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, J.D. Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
 \thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
@@
 \begin{abstract}
-This paper presents {\small CUBEP3M}, a high performance cosmological N-body code
+This paper presents {\small CUBEP$^3$M}, a high performance cosmological N-body code
 and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
 a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
-{\small CUBEP3M} is fast, has a low memory imprint, and 
+{\small CUBEP$^3$M} is fast, has a low memory imprint, and 
 has been run on up to $20,000$ cores, achieving close to ideal weak scaling
@@
 non-Gaussian statistics.
-%Since there is no explicit load balancing algorithm, the code runs optimally 
-%for homogeneous systems such as those which prevail in large scale structures studies. 
-%It has an interface with a MHD-TVD module and a radiative transfer code, 
-%which opens the door to a vast number of science applications.
-%or a large number of realizations for statistical studies.
-%one of the fastest highest performance public N-body code. 
-%Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
-%interactions inside and across neighbouring cells. 
-%Among important improvements,  {\small CUBEP3M} has a volumetric decomposition,
-%allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
-%even at this scale.
-In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
 We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
@@
 Soon after, it incorporated particle-particle (pp) interactions at the sub-grid level,
-and was finally renamed {\small CUBEP3M}. The public package now includes a significant number of new features: the pp force
+and was finally renamed {\small CUBEP$^3$M}. The public package now includes a significant number of new features: the pp force
 can be extended to an arbitrary range, the size of the redshift jumps can be constrained for improved accuracy during the first time steps,
 a runtime halo finder has been implemented, the expansion has also been generalized to include a redshift dependent equation of state of dark energy, there is a system of unique particle identification that can be switched on or off, and the initial condition generator has been generalized as to include non-Gaussian features.
- {\small PMFAST} was equipped with a multi-time stepping option that has not been tested on {\small CUBEP3M} yet, but which is, in principle at least, still available.
+ {\small PMFAST} was equipped with a multi-time stepping option that has not been tested on {\small CUBEP$^3$M} yet, but which is, in principle at least, still available.
 
  The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
- {\small CUBEP3M} is therefore one of the most competitive and versatile public N-body code, 
+ {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
  and has been involved in a number of scientific applications over the last few years,
@@
 Notably, the fact that the force calculation is multi-layered makes the code extendible, 
-and opens the possibility to run {\small CUBEP3M} on hybrid CPU-GPU clusters.
+and opens the possibility to run {\small CUBEP$^3$M} on hybrid CPU-GPU clusters.
 It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
@@
 
-Since {\small CUBEP3M} is not new, the existing  accuracy and systematic tests were performed by different groups, on different machines,
+Since {\small CUBEP$^3$M} is not new, the existing  accuracy and systematic tests were performed by different groups, on different machines,
 and with different geometric and parallelization configurations. It is not an ideal situation in which to quantify the performance, and each test must be viewed as a
@@
 most of the time. In this section, we present the general strategies adopted to address these challenges\footnote{ 
-Many originate directly from MPT and were preserved in {\small CUBEP3M};
+Many originate directly from MPT and were preserved in {\small CUBEP$^3$M};
 those will be briefly mentioned, and we shall refer the reader to the original {\small PMFAST} paper for greater details.}.
@@
 
-As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
+As mentioned in the Introduction section, {\small CUBEP$^3$M} is a {\small FORTRAN90} 
 N-body code that solves Poisson's equation on a two-level mesh, 
@@
 and allows to optimize the code for higher resolution configurations.
-In terms of memory, {\small CUBEP3M} can be configured such as the main footprint is dominated exclusively by the 
+In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
 particle phase space array, in which case it uses about $60$ bytes per particles. 
@@
 creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
-with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
+with the exception  that {\small CUBEP$^3$M} decomposes the volume into cubes (as opposed to slabs). 
 The loop over tiles and the particle exchange are thus performed in three dimensions.
@@
 limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
---  such as {\small CUBEP3M} -- requires the number of {\small MPI} processes to be a perfect cube.
+--  such as {\small CUBEP$^3$M} -- requires the number of {\small MPI} processes to be a perfect cube.
 Also,  since the decomposition is volumetric, as opposed to density dependent -- 
@@
 Many parts of the algorithm are identical to {\small PMFAST}, hence we refer the reader 
-to section 2 of MPT for more details. In {\small CUBEP3M}, the mass default assignment scheme are
+to section 2 of MPT for more details. In {\small CUBEP$^3$M}, the mass default assignment scheme are
 a `cloud-in-cell' (CIC) interpolation for the coarse grid,  and a `nearest-grid-point' (NGP) interpolation 
@@
 are typically single line changes to the code. \citet{1993ApJ...409...60D} argues in favour on the uniform density profile scheme -- which is more physical --
-and future developments of {\small CUBEP3M} will incorporate this option.
+and future developments of {\small CUBEP$^3$M} will incorporate this option.
 
@@
 %  \vskip -0.5cm 
-  \caption{Scaling of {\small CUBEP3M} on Curie fat nodes (left) and 
+  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
     on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
@@
   \begin{center}
-\caption{Scaling of {\small CUBEP3M} on Curie. Speedup is 
+\caption{Scaling of {\small CUBEP$^3$M} on Curie. Speedup is 
 scaled to the smallest run.}
@@
 \end{tabular}
-\caption{Scaling of  {\small CUBEP3M} on Ranger. Speedup is scaled to the smallest run.}
+\caption{Scaling of  {\small CUBEP$^3$M} on Ranger. Speedup is scaled to the smallest run.}
 \label{summary_scaling_table2}
@@
 
-The parallel algorithm of {\small CUBEP3M} is designed for `weak' 
+The parallel algorithm of {\small CUBEP$^3$M} is designed for `weak' 
 scaling, i.e. if the number of cores and the problem size 
@@
 
-The scaling of {\small CUBEP3M}  was first  tested with a dedicated series 
+The scaling of {\small CUBEP$^3$M}  was first  tested with a dedicated series 
 of simulations -- the CURIE simulation suite-- by increasing the size and number of cores on the `fat' 
@@
 
-We have also ran {\small CUBEP3M} on a much larger number of cores, 
+We have also ran {\small CUBEP$^3$M} on a much larger number of cores, 
 from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
@@
 a growth of structure that does not match the predictions of  linear theory even at the largest scales.
-One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
+One therefore needs to choose a maximum step size. In {\small CUBEP$^3$M}, this is controlled by $r_{max}$, which is the fractional step size,
 $\mbox{d}a/(a + \mbox{d}a)$ and is set to $0.05$ by default.  Generally, a simulation should start at a redshift high enough that
@@
 on-the-fly at a pre-determined list of redshifts. The halo finding is 
-massively-parallel and threaded based on the main {\small CUBEP3M} data structures 
+massively-parallel and threaded based on the main {\small CUBEP$^3$M} data structures 
 discussed in section \ref{sec:structure}. The code first builds the 
@@
 as described at the beginning of section \ref{sec:structure}. A few extensions have been recently developed
-in order to enlarge the range of applications of {\small CUBEP3M}, and this section briefly 
+in order to enlarge the range of applications of {\small CUBEP$^3$M}, and this section briefly 
 describes the most important improvements.
@@
 
-The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
+The original code that provides Gaussian initial conditions for {\small CUBEP$^3$M} 
 is extended to include non-Gaussian features of the `local' form,
@@
   \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen, at $z = 1.0$. 
-  The smallest haloes shown here have a mass equivalent to 100 particles, and fall in the lowest mass bin.
+  The smallest haloes shown here have a mass equivalent to 20 particles, each of $2.79 \times 10^{9} M{\odot}$, and fall in the lowest mass bin.
   ({\it bottom}:) Ratio between the different curves and that of the default P$^3$M configuration.}
@@
 
-This paper describes {\small CUBEP3M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
+This paper describes {\small CUBEP$^3$M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
  and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
@@
 including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
-{\small CUBEP3M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
+{\small CUBEP$^3$M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
 it has already been used  for a large number of scientific applications, and it is our hope that the current documentation will 

### 2012-08-13 paper_new.tex
latest changes about the new kernels, more details about initial conditions and memory footprint, and included recommendations by Ilian.

@@
 
-\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP$^3$M]{{\small} High Performance P$^{3}$M N-body code: CUBEP$^3$M}
+\title[CUBEP$^3$M]{High Performance P$^{3}$M N-body code: CUBEP$^3$M}
 \author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, J.D. Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
@@
 \begin{abstract}
-This paper presents {\small CUBEP$^3$M}, a high performance cosmological N-body code
+This paper presents {\small CUBEP$^3$M}, a high performance, publicly-available, cosmological N-body code
 and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
@@
 even at this scale.
-It is well suited for a vast number of science applications that 
+It is well suited and has already been used for a broad number of science applications that 
 require either large samples of non-linear realizations or 
@@
 can thereby be put towards speed, memory compactness and scalability.  
-{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) was one of the first code designed specifically to optimize the PM algorithm,
-both in terms of speed and memory usage, and uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T}.
+{\small PMFAST} (\citet{2005NewA...10..393M}, MPT hereafter) was one of the first code designed specifically to optimize the PM algorithm,
+both in terms of speed and memory usage, and uses a two-level mesh algorithm based on the gravity solver of \citet{2003AAS...203.9703T}.
 The long range gravitational force is computed on a  grid $4^{3}$ times coarser, such as to minimize the {\small MPI} communication time
@@
 we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
-including much larger runs that were produced at various high performance computing centers;
+including much larger runs that were produced at various high performance computing centres;
 section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
@@
 
-When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, center of mass, etc.
+When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, centre of mass, etc.
 It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.
@@
 \item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
+\item{{\tt send\_buf, recv\_buf} store the particles to be {\small MPI}-passed to the neighbouring sub-volumes}
 \item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
@@
 and allows to optimize the code for higher resolution configurations.
-In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
-particle phase space array, in which case it uses about $60$ bytes per particles. 
-For a fixed problem size, this can be achieved by increasing the number of tiles, thereby reducing the size of the local fine mesh arrays.
+
+{\bf In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
+particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
+For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
 There is a limit at which we can do such a volume breakdown, though, since the physical 
-volume on each tile must be much larger than the connection length between the two meshes (see section \ref{sec:Poisson}).
+volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
+In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
+if little particle passing is expected. 
+However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
+A higher number of cells per tile with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy.
 %
 %uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
-For comparison, (Lean) {\small GADGET-2} code uses around $90$ per particles.
+For comparison, Lean-{\small GADGET-2} code uses  $84$ per particles \citep{Gadget2Lean}.}
+
 %\subsection{Code overview}
@@
 contains the local  number of particles, the current redshift and time step, and the constraints on the time step jump
-from the previous iteration; the header is followed by the phase space of each particles. The particle identification tags are written 
+from the previous iteration; the header is followed by the position and velocity of each particle. The particle identification tags are written 
 with a similar fashion in distinct files to simplify the post-processing coding. 
@@
 by default in binary format for compactness, and has been shown to scale well to large data sets. 
+At this stage, the performance depends only on the setup and efficiency of the parallel file system.
 
@@
 
-Other constraints that need to be considered is that any decomposition geometry somehow 
-limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
+Other constraints that need to be considered is that any decomposition geometry 
+limits in some ways permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
 --  such as {\small CUBEP$^3$M} -- requires the number of {\small MPI} processes to be a perfect cube.
@@
 However, large scale cosmology problems are very weakly affected by this effect, which
-can generally  be overcome by allocating a mild amount of extra memory per node.
-As mentioned in section \ref{sec:scaling}, large runs can even be optimized by  allocating only a small amount for the first part of it,
-follow by a restart with a configuration that allows more memory per node.
+can generally  be overcome by allocating a small amount of extra memory per node.
+As mentioned in section \ref{sec:scaling}, large runs can even be optimized by  allocating 
+less memory to start with, follow by a restart with a configuration that allows more memory per node, if required.
 
@@
 \end{eqnarray}
-Because the gradient acting on \ref{eq:poisson_solution_pot} affects only unprime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
+Because the gradient acting on Eq. \ref{eq:poisson_solution_pot} affects only unprime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
 \begin{eqnarray}
@@
 and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
+{\bf A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
+This is caused by the uneven balance of the force about the $1/r^2$ law in the NGP interpolation scheme.
+Although most figures in this paper were made prior this correction, the new code configuration applies a 90 per cent boost to 
+the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).}
 
@@
 In this test, which is purposefully probing rather deep in the non-linear regime, reducing the softening length
-to half it value ($1/20^{th}$ of a grid cell) double the run time. The gain in power spectrum is less than two per cent.
+to half its value ($1/20^{th}$ of a grid cell) double the run time. The gain in power spectrum is less than two per cent.
 Similarly, increasing to $0.2$ the softening length reduces almost by a half the run time, but suffers from a five per cent loss in power at the resolution turn-around.
-One tenth of grid cell seems to be the optimal choice in this trade off, and is therefor the default value, however it should really be considered a free parameter.
+One tenth of grid cell seems to be the optimal choice in this trade off, and is therefore the default value, however it should really be considered a free parameter.
 
@@
 \begin{center}
-\caption{Scaling in {\small CPU} resources as a function of the softening length.}
+\caption{Scaling in {\small CPU} resources as a function of the softening length, in units of fine grid cells.}
 \begin{tabular}{|l|c|c|}
@@
 due to increased communication 
-costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
+costs, I/O overheads (a single time slice of $5488^3$ particles is 3.6 TB)
 and load balancing issues, but still within $\sim20$ per cent of ideal. 
@@
 with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
-Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
+Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to ensure 
 local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
@@
  but depends on the particular halo finder used.
- We thus post-pone the discussion on this aspect until section \ref{sec:halo},
+ We thus postpone the discussion on this aspect until section \ref{sec:halo},
 and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
@@
 \includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
-  \caption{Detail of a dark matter density projection, measured at $z=0$ in a RANGER4000 simulation.
-  Particles are shown in pink, haloes in blue in the online version. The full projection is 64 times larger,
-  but does not provide much details on the structure on a paper format.
+  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Êper side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
+  Particles are shown in pink, haloes in blue in the online version, and the colour scale is linear. The full projection is 64 times larger,
+  but we show here only a sub-volume, which provides more details on the structure.
     \label{fig:density}}
@@
 tangential contribution (bottom), also calculated from a single time step.
-The fractional error of the radial component is up to 60 per cent at sub-grid distances, but remains under ten per cent for  distances larger than a few cells. 
-The transverse fractional error peaks exactly at the grid distance,  and is about two time smaller than the radial counterpart.
+Again, particles that are in the same cell have zero fractional error in the radial direction, and zero tangential force.
+Particles feel a large scatter from the neighbouring fine mesh cells, but as the distance increase beyond five fine cells, the fractional error drops down to about 10 per cent.
+%The fractional error of the radial component is up to 60 per cent at sub-grid distances, but remains under ten per cent for  distances larger than a few cells. 
+The transverse fractional error peaks exactly at the grid distance,  and is about 50 per cent smaller than the radial counterpart.
 
@@
 We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
-there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the numerical calculations are otherwise
+there is a 50 per cent fractional error in the range $ 0.4 < r < 2$ (in units of fine grid), but the numerical calculations are otherwise
 in excellent agreement with Newton's law.
-The transverse force is somehow ill-defined, since it is by definition a positive number.
-Since we know that it averages out to zero over many time steps, we plot only the scatter about the mean, represented in the figure by the solid line.
+Since  the transverse force is, by definition, a positive number,
+and since we know that it averages out to zero over many time steps, we plot only the scatter about the mean, represented in the figure by the solid line.
 
@@
   \begin{center}
-    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
-%    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
+ %   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new-1.eps}
+   % \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
+    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new_kern.eps}
+    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new_kern.eps}
+    %\includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
+   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern.eps}
+   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern.eps}
+ %   \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
+   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern_N10.eps}
+   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern_N10.eps}
+%    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
+   % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
 %    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_N10_new-1.eps}
-  \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
+  \caption{({\it top left}:) Gravity force in the P$^3$M algorithm, versus distance, compared with the exact $1/r^{2}$ law.
+   Distance is in fine mesh cell units, and force in simulation units.
     This particular calculation was obtained in a CITA128 realization with  a box size of $100 h^{-1}\mbox{Mpc}$,
@@
     thanks to the inclusion of a random offset that is imposed on each particle, as discussed in the main text.
-    ({\it middle row}:) Top row organized in 50 logarithmic bins; we plot only the scatter about the mean for the tangential contribution.  
+    ({\it middle row}:) Top row organized in 50 logarithmic bins; for the tangential contribution,  we plot only the scatter about the mean.  
     ({\it bottom row}:) Same as middle row, but averaged over ten time steps.
@@
 We ensure that, on average, this solution balances out the mesh feature,
-by tuning the force kernels such as to provide a force as evenly balanced as possible, both at grid cell distances
-and at the cutoff length ($r_{c}=16$ fine cells).
-These adjustments are performed from the pairwise force test mentioned above (and described in MPT).
+by tuning the force kernels such as to provide a force as evenly balanced as possible, both at the cutoff length ($r_{c}=16$ fine cells) 
+and at grid cell distances, discussed in section \ref{sec:Poisson}.
+Both of these adjustments are performed from the above force test and the pairwise force test mentioned above (and described in MPT).
 The bottom panels of Fig. \ref{fig:den_force_fracErr} show the effect of averaging over ten time steps, 
 where the random offset is applied on the particles (and on the hole) at the end of each force calculation. 
-We observe that the agreement extends at the per cent level down to $r \sim 1.5$, beyond which point
-the discretization effect of the NGP causes a cumulative underestimate of the force, which reaches up 60 per cent.
+We observe that the agreement with Newton's law is at the per cent level for  $r \ge 1.5$. 
+At smaller distances, we still observe a scatter, but it is about two times smaller than during a single time step for $r > 1 $.
+The force at sub-grid distances is, on average, biased on the low side by 20-40 per cent,
+as caused by the discretization effect of the NGP.
 %{\bf (Can't we beat this by tempering with the fine force kernel? Will the least square fit enhance this?)}
@@
     \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
-  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is 1000 Mpc/$h$ per side.
+  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is $1000 h^{-1}$ Mpc per side. 
   The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
@@
 %\end{figure}
-Although most of the code tests presented in this paper were obtained with the configuration described above, 
-we present here a recent improvement in the fine force kernel that compensates for the systematic underestimate
-at grid distances. Basically, the idea is to boost the small distance elements of the kernel by a small amount,
-which somehow corrects for the loss observed in the bottom panels of Fig. \ref{fig:den_force_fracErr}.
-We show here the impact of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
+
+{\bf 
+As mentioned in section \ref{sec:Poisson}, most of the tests presented in this paper were obtained without the fine force kernel correction at  $r=1$.
+For completeness, we present here the motivation behind the recent improvement that compensates for a systematic underestimate
+at grid distances. 
+Without the correction, the force of gravity looks like Fig. \ref{fig:den_force_fracErr_kern}, 
+where we observe that the fractional error in the radial direction is systematically on the low side, by up to 50 per cent. 
+This is caused by the fact that in the strong regime of a $1/r^2$ law, the mean force from a cell of uniform density, as felt by a particle, 
+is not equivalent to that of the same cell collapsed to its centre -- the actual position of equivalence would be at a closer distance.
+Basically, the idea is to boost the small distance elements of the kernel by a small amount,
+which thereby corrects for the loss of force above-described, and observed in Fig. \ref{fig:den_force_fracErr_kern}.
+
+We  show in Fig. \ref{fig:fudge} the impact on the power spectrum of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
 the force from the first two layers of neighbours is boosted by $60$ per cent, and c) by $80$ per cent.
@@
 whereas the first trial allows for a similar resolution down to $k\sim 4.5 h\mbox{Mpc}^{-1}$. 
-Other correction schemes are more aggressive, and although they recover power at even smaller scales, 
+The two other correction schemes are more aggressive, and although they recover power at even smaller scales, 
 they exhibit a small overestimate compare to non-linear predictions in the range $2.0 < k < 4.0 h\mbox{Mpc}^{-1}$, 
 and more testing should be done prior to using these corrections. 
-
-We therefore recommend that the fine force kernel should be considered as an adjustable parameter;
-`no correction' is considered as the  conservative mode, which is accurate only up to the resolution turn around 
-observed in the dimensionless power spectrum. Higher resolution can be achieved with the modified fine force kernel following `trial a',
-and should be considered as an actual resolution improvement, with no cost in accuracy nor in run time;
-`trial b' and `trial c' should be considered as too aggressive. We note that we have not exhausted the list of possible correction,
-and that future work in that direction might result in even better resolution.
+We have also tried to increase even more the boost factor in the case a), but once again, the effect started to propagate to larger scales, 
+an effect we need to avoid. %We show in Fig. \ref{fig:den_force_fracErr_kern} the impact of the trial a) on the force of gravity. 
+%We observe that when averaged over ten time steps, the scatter  is five to ten times larger than with the original kernel, 
+%both in the radial and tangential directions, but the radial part is better balanced about the theoretical prediction.
+
+We therefore conclude that the fine force kernel should be subject to a boost factor following `trial a', i.e. 
+$\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$.
+It is thinkable that other corrections could outperform this one, hence the kernel correction factor
+should be considered as an adjustable parameter;
+`no correction' is considered as the conservative mode, which is accurate only up to the turn around 
+observed in the dimensionless power spectrum, and which has been used in most of the tests presented in this paper. 
+Since higher resolution can be achieved at no cost with the corrected fine force kernel, it has now been adopted in the default configuration;
+`trial b' and `trial c' should be considered as too aggressive. We note, again, that we have not exhausted the list of possible correction,
+and that future work in that direction might result in even better resolution gain.}
 
@@
   \begin{center}
+%    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
+  %  \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern_N10.eps}
+  % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
+    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
+  \caption{Fractional error in the gravity force, without the fine force kernel correction at $r=1$.
+      This particular calculation was obtained in a CITA128 realization with  a box size of $150 h^{-1}\mbox{Mpc}$,
+     averaged over ten time steps.
+    \label{fig:den_force_fracErr_kern}}
+\end{center}
+\end{figure}
+
+\begin{figure}%[ht]
+  \begin{center}
     \includegraphics[width=3.2in]{graphs/power_fudge.eps}
-  \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is 150 Mpc/$h$ per side. 
-  The thick solid line represents predictions from {\small CAMB}, the thin solid line is the default configuration of the code, and the three other lines represent
-  three different corrections that are applied at grid distances to compensate for the systematic force underestimation seen in the bottom panels of Fig. \ref{fig:den_force_fracErr}. }
+  \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is $150 h^{-1}$ Mpc per side. 
+  The thick solid line represents predictions from {\small CAMB}, the thin solid line is the default (i.e. without correction) configuration of the code, and the three other lines represent
+  three different corrections that are applied at grid distances to compensate for the systematic force underestimation seen in the bottom panels of Fig.
+\ref{fig:den_force_fracErr}.
+\label{fig:fudge}}
 \end{center}
@@
 The default configuration of the code is set to $r_{max}=0.05$, but at the light of this recent test, we recommend reducing to the value $0.01$ or even $0.005$.
-We also mention here that with a proper use of second order initial conditions (also see section \ref{subsec:init}), it is possible to start the simulations at much later redshifts
+We also mention here that with a proper use of second order initial conditions, it is possible to start the simulations at much later redshifts
 without loosing much accuracy.
@@
 profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
-$\sim130$, and a simple $1/r^2$ law for lower overdensities lower. 
-These yield a similar outer slope to the Navarro, Frenk and White 
+$\sim130$, and a simple $1/r^2$ law for lower overdensities. 
+TIS yields a similar outer slope to the Navarro, Frenk and White 
 \citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
@@
 These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
-local volume. \footnote{We remind that for Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshifts.
+local volume. {\bf Each {\small MPI} sub-volume generates its own random seed, 
+which are used to create a local noise map. The initial power spectrum is found by reading off a transfer function that is by default generated from {\small CAMB}, 
+and evolved to the desired initial redshift with the linear growth function. 
+The noise map is then Fourier transformed, each element is multiplied by the (square root of the) power spectrum, and the result is brought back to real space. 
+To ensure that the global volume is correct, a buffer layer in the gravitational potential is exchanged across each adjacent node.
+The particles are then moved from a uniform cell-centred position to a displaced position with Zel'dovich approximation\footnote{We remind that 
+for Zel'dovich approximation to hold in such cases, the simulations need to be started at very early redshifts.
 Consequently, the size of the first few redshift jumps in such simulations can become rather large, and therefore less accurate, 
@@
 Another solution to this issue is to use second order perturbation theory to generate the initial conditions,
-which is not implemented yet in the public package.}
-Although many applications are typically based on initial conditions that follow Gaussian statistics,
-we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
-This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.
+which is not implemented yet in the public package.}.
+The density field produced by this algorithm follows Gaussian statistics, and is well suited to describe many systems.
+To increase the range of applicability of this code, we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
+This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.}
 
@@
 Similarly to the phase space array, the {\tt PID} array gets written to file at each particle checkpoint.
-Since the array takes a significant amount of memory, we opted to store the tags in a seperate object
+Since the array takes a significant amount of memory, we opted to store the tags in a separate object
 -- as opposed to adding  an extra dimension to the existing {\tt xv} array, such that it can be turned off 
@@
  and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
+ This code is fast and has a memory footprint more than two times smaller than Lean-{\small GADGET-2}. 
 We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests

### 2012-08-20 MNRAS_V1/paper_new.tex
Ready to Commit?

--- 2012-08-13 paper_new.tex
@@
 \citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
-2011arXiv1107.4772I}. The reionization process is primarily driven by low-mass 
+2012MNRAS.423.2222I}. The reionization process is primarily driven by low-mass 
 galaxies, which for both observational and theoretical reasons, need to be resolved 
@@
 common practice, and N-body codes have evolved both in performance and complexity. 
-Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, and Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field.
+Many have opted for  `tree' algorithms, including {\small GADGET} \citep{2001NewA....6...79S, 2005MNRAS.364.1105S}, {\small PMT} \citep{1995ApJS...98..355X}, {\small GOTPM} \citep{2004NewA....9..111D}, or adaptive P$^3$M codes like Hydra \citep{1995ApJ...452..797C}, in which the local resolution increases with the density of the matter field.
 These often have the advantage to balance the work load across the computing units, which enable fast calculations even in high density regions. 
@@
 
- The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C2-Ray \citep{2006NewA...11..374M}.
+ The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C$^2$-ray \citep{2006NewA...11..374M}.
  {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
  and has been involved in a number of scientific applications over the last few years,
-spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, 2012arXiv1202.2332H},  BAO
- \citep{2010arXiv1008.3506Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
+spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2012MNRAS.421..832Y, 2012arXiv1202.2332H},  BAO
+ \citep{2011ApJ...728...35Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
 formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
-observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
-reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}. 
+observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.422.2164I} and
+reionization \citep{2012MNRAS.423.2222I,2012ApJ...750...20F,2011MNRAS.413.1353F,2012MNRAS.422..926M,2012MNRAS.424.1877D,2012MNRAS.424..762D}. 
 Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
@@
 Notably, the fact that the force calculation is multi-layered makes the code extendible, 
-and opens the possibility to run {\small CUBEP$^3$M} on hybrid CPU-GPU clusters.
+and opens the possibility to run {\small CUBEP$^3$M} on hybrid CP2010arXiv1012.0444YU-GPU clusters.
 It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
@@
 \item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
-\item{{\tt PID} stores the unique particle identification tags.}
+\item{{\tt PID} stores the unique particle identification tags as double integers.}
 \end{enumerate}
@@
 
-{\bf In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
+ In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
 particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
@@
 However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
-A higher number of cells per tile with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy.
-%
+A higher number of cells per tile -- typically $176^3$ -- with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy,
+and the memory per particle is closer to 100 - 140 bytes. Also, enabling the particle ID adds about 24 bytes per particle: 8 for the main array, and 8 for the two {\small MPI}-passing buffers as well.
 %uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
-For comparison, Lean-{\small GADGET-2} code uses  $84$ per particles \citep{Gadget2Lean}.}
+For comparison, Lean-{\small GADGET-2} code uses  $84$ bytes per particles \citep{Gadget2Lean}.
 
@@
 and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
-{\bf A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
+A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
 This is caused by the uneven balance of the force about the $1/r^2$ law in the NGP interpolation scheme.
 Although most figures in this paper were made prior this correction, the new code configuration applies a 90 per cent boost to 
-the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).}
+the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/power_rsoft.eps}
+    \includegraphics[width=3.2in]{graphs/fig3.eps}
   \caption{Dark matter power spectrum, measured at $z=0.5$ in a series of SciNet256 simulations in which the softening length is varied.
@@
   \begin{center}
-    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_curie.eps}
-    \includegraphics[width=3.0in]{graphs/scaling_cubep3m_new.eps}
+    \includegraphics[width=3.0in]{graphs/fig4a.eps}
+    \includegraphics[width=3.0in]{graphs/fig4b.eps}
 %  \vskip -1.2cm 
@@
 %    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
-\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
-  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Êper side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
+\includegraphics[width=5.2in]{graphs/fig5.eps}
+  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Ê per side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
   Particles are shown in pink, haloes in blue in the online version, and the colour scale is linear. The full projection is 64 times larger,
@@
   \begin{center}
-    \includegraphics[width=5.2in]{graphs/power_highres.eps}
+    \includegraphics[width=5.2in]{graphs/fig6.eps}
   \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
@@
    % \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_new_kern.eps}
-    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_new_kern.eps}
+    \includegraphics[width=3.0in]{graphs/fig7a.eps}
+    \includegraphics[width=3.0in]{graphs/fig7b.eps}
     %\includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new-1.eps}
-   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern.eps}
-   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern.eps}
+   \includegraphics[width=3.0in]{graphs/fig7c.eps}
+   \includegraphics[width=3.0in]{graphs/fig7d.eps}
  %   \includegraphics[width=3.0in]{graphs/DeltaF_pp0.eps}
-   \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_new_kern_N10.eps}
-   \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=0_rebin_new_kern_N10.eps}
+   \includegraphics[width=3.0in]{graphs/fig7e.eps}
+   \includegraphics[width=3.0in]{graphs/fig7f.eps}
 %    \includegraphics[width=3.0in]{graphs/densityForce_ppext=0_rebin_N10_new-1.eps}
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
+    \includegraphics[width=3.2in]{graphs/fig8.eps}
   \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a CITA256 realization that is $1000 h^{-1}$ Mpc per side. 
@@
 
-{\bf 
 As mentioned in section \ref{sec:Poisson}, most of the tests presented in this paper were obtained without the fine force kernel correction at  $r=1$.
@@
 We  show in Fig. \ref{fig:fudge} the impact on the power spectrum of three different trials: a) the force from the first layer of neighbours is boosted by $90$ per cent, b)
-the force from the first two layers of neighbours is boosted by $60$ per cent, and c) by $80$ per cent.
+the force from the first two layers of neighbours is boosted by $60$ per cent, and c) the force from two layers is boosted by $80$ per cent.
 We observe that with the trial a), we are able to improve the power at turn around by a factor of $1.5$, which almost doubles the resolution.  
-This test was performed in a series of SciNet256 simulations with a volume of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
+This test was performed in a series of SciNet256 simulations with a box length of $150 h^{-1} \mbox{Mpc}$ and at $z=0$.
 We see from the figure that in the original configuration, the departure from the non-linear predictions occur at $k\sim 3.0 h\mbox{Mpc}^{-1}$,
@@
 `trial b' and `trial c' should be considered as too aggressive. We note, again, that we have not exhausted the list of possible correction,
-and that future work in that direction might result in even better resolution gain.}
+and that future work in that direction might result in even better resolution gain.
 
@@
   % \includegraphics[width=3.0in]{graphs/DeltaF_pp0_new_kern.eps}
-    \includegraphics[width=3.0in]{graphs/DeltaF_pp0_N10.eps}
+    \includegraphics[width=3.0in]{graphs/fig9.eps}
   \caption{Fractional error in the gravity force, without the fine force kernel correction at $r=1$.
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/power_fudge.eps}
+    \includegraphics[width=3.2in]{graphs/fig10.eps}
   \caption{Dark matter power spectrum, measured at $z=0$, in a SciNet256 realization that is $150 h^{-1}$ Mpc per side. 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
+    \includegraphics[width=3.2in]{graphs/fig11.eps}
   \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
-  \end{center}
+    \includegraphics[width=3.2in]{graphs/fig12.eps}
   \caption{Simulated halo multiplicity function, 
@@
     \label{mf}}
+     \end{center}
 \end{figure}
@@
 These are organized as a set of $6 \times N$ phase-space arrays -- one per {\small MPI} process -- where $N$ is the number of particles in the
-local volume. {\bf Each {\small MPI} sub-volume generates its own random seed, 
-which are used to create a local noise map. The initial power spectrum is found by reading off a transfer function that is by default generated from {\small CAMB}, 
+local volume. Each {\small MPI} sub-volume generates its own random seed, 
+which are used to create a local noise map. The initial power spectrum is found by reading off a transfer function at redshift $z=0$ that is generated from {\small CAMB}  by default, 
 and evolved to the desired initial redshift with the linear growth function. 
@@
 To increase the range of applicability of this code, we have developed a non-Gaussian initial conditions generator that we briefly describe in this section. 
-This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.}
+%This becomes important especially for simulations in which the sensitivity to very small scales becomes more important.
 
@@
   \begin{center}
-    \includegraphics[width=3.2in]{graphs/dm_power_ngp_z0_fNL0_50_w_1loop_PT.eps}
+    \includegraphics[width=3.2in]{graphs/fig13.eps}
   \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
@@
   \begin{center}
-    \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/DeltaF_pp2.eps}
+    \includegraphics[width=3.0in]{graphs/fig14a.eps}
+    \includegraphics[width=3.0in]{graphs/fig14b.eps}
 %    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_new-1.eps}
-     \includegraphics[width=3.0in]{graphs/densityForce_ppext=2_rebin_N10_new-1.eps}
-    \includegraphics[width=3.0in]{graphs/DeltaF_pp2_N10.eps}
+     \includegraphics[width=3.0in]{graphs/fig14c.eps}
+    \includegraphics[width=3.0in]{graphs/fig14d.eps}
 %    \includegraphics[width=3.0in]{graphs/densityForce_fracErr_ppext=2_rebin_N10_new-1.eps}
@@
   \begin{center}
-  \epsfig{file=graphs/pp_ext_new_z1.eps,width=0.49\textwidth}
+  \epsfig{file=graphs/fig15,width=0.49\textwidth}
   \caption{ Dimensionless power spectrum for varying range of the exact pp force.
@@
   \begin{center}
-  \epsfig{file=graphs/MassFunction_pp_ranges_new_z1.eps,width=0.49\textwidth}
+  \epsfig{file=graphs/fig16.eps,width=0.49\textwidth}
   \caption{ ({\it top}:) Halo mass function for different ranges of the pp force calculation, compared to the predictions of Sheth-Tormen, at $z = 1.0$. 

### 2012-08-21 MNRAS_V1/paper_new.tex
Entered comments by Ue-Li

@@
 a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
-{\small CUBEP$^3$M} is fast, has a low memory imprint, and 
+{\small CUBEP$^3$M} is fast, has a memory imprint up to three times lower than other N-body codes, and 
 has been run on up to $20,000$ cores, achieving close to ideal weak scaling
@@
 Notably, the fact that the force calculation is multi-layered makes the code extendible, 
-and opens the possibility to run {\small CUBEP$^3$M} on hybrid CP2010arXiv1012.0444YU-GPU clusters.
+and opens the possibility to run {\small CUBEP$^3$M} on hybrid CPU-GPU clusters.
 It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.
@@
  In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
-particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
+particle phase space array, and the linked list, in which case it asymptotes to about $28$ bytes per particles. 
 For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
@@
 volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
-In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
-if little particle passing is expected. 
-However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
+This allows a low rate of particle exchange across nodes;  in the default configuration, the {\tt send\_buf} and  {recv\_buf} arrays 
+are much larger, in which case the memory per particle goes up to $44$ bytes. 
+Enabling the particle ID adds at least 8 bytes per particle from the main array, and as the {\small MPI}-communication buffers are allocated larger sizes,
+they contribute at most an additional 16 bytes as well. Hence a `lean' version with the ID enable uses down to $36$ bytes per particles.
+In practice, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- the {\small FFTW} for instance.
 A higher number of cells per tile -- typically $176^3$ -- with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy,
-and the memory per particle is closer to 100 - 140 bytes. Also, enabling the particle ID adds about 24 bytes per particle: 8 for the main array, and 8 for the two {\small MPI}-passing buffers as well.
+and the memory per particle is closer to 100 - 140 bytes. 
 %uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
@@
  and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
- This code is fast and has a memory footprint more than two times smaller than Lean-{\small GADGET-2}. 
+ This code is fast and has a memory footprint up to three times smaller than Lean-{\small GADGET-2}. 
 We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests
@@
 Computing in Europe (PRACE) grant 2010PA0442 which supported the code
-scaling studies. UEP and JDE are supported by the NSERC of Canada.
+scaling studies. ULP and JDE are supported by the NSERC of Canada.
 VD acknowledges support by the Swiss National Science Foundation.
@@
 
-\bibliographystyle{mn2e}
+%\bibliographystyle{mn2e}
+\bibliographystyle{hapj}
 \bibliography{mybib3_new}{}
-%\bibliographystyle{amsplain}
 
